### 2021-04-04 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.Rmd
Initial commit with data package

---
title: "An examination of accessibility to pilot COVID-19 vaccination sites in Hamilton, Ontario"
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliation: University of Toronto Scarborough
address:
  - code: McMaster University
    address: School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada
  - code: University of Toronto Scarborough
    address: Department of Geography & Planning, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON M1C1A4
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Abstract

journal: "Transport Findings"
date: "`r Sys.Date()`"
bibliography: References.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
always_allow_html: true

header-includes:
 \usepackage{float}
 \usepackage{xcolor}
 \floatplacement{figure}{H}
---

```{r install-data-package, include = FALSE}
# Run only once if needed to install data package `sobiEquity`
if (!require("vaccHamilton", character.only = TRUE)) {
      devtools::install_github("https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines", subdir = "vaccHamilton")
  }
```

```{r load-packages, include=FALSE}
#Load the packages needed to read and work with the data
#library(circlize)
library(grid)
library(gridExtra)
library(kableExtra)
library(png)
#library(readr)
library(sf)
library(tidyverse)
library(vaccHamilton)
```

Research Questions and Hypotheses
==========================

Words.

The objective of this paper is to investigate...Using data...we pose the following questions:

- Question 1
- Question 2

This paper is a reproducible research document [see @Brunsdon2020opening]; the code and data necessary to reproduce the tables and figures are available in a public repository\footnote{\url{https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines}}

Methods and Data
============

Words.

Findings
===================

Words.

References {#references .unnumbered}
==========

### 2021-04-04 README.Rmd
Initial commit with data package

---
output:
  github_document:
    pandoc_args: --webtex
always_allow_html: true
bibliography: Accessibility-Vaccination-Sites-Hamilton/references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# An examination of accessibility to pilot COVID-19 vaccination sites in Hamilton, Ontario

<!-- badges: start -->
<!-- badges: end -->

Antonio Páez (McMaster University)  
Christopher D. Higgins (University of Toronto Scarborough)

Paper submitted to *Transport Findings*

## Abstract

Words.

## Keywords

- COVID-19  
- Vaccination sites
- Accessibility  
- Hamilton  
- Ontario
- Reproducible research

### 2021-04-04 vaccHamilton/R/vaccHamilton.R

@@ -1,0 +1,210 @@
+#' vaccHamilton
+#'
+#' vaccHamilton is a companion package for the paper "An examination of accessibility to pilot COVID-19 vaccination sites in Hamilton, Ontario".
+#'
+#' The package includes the data used in the analysis.
+#'
+#' Contains information licensed under the Open Government Licence – Canada (https://open.canada.ca/en/open-government-licence-canada).
+#'
+#' Contains public sector Data made available under the City of Hamilton’s Open Data Licence (https://www.hamilton.ca/city-initiatives/strategies-actions/open-data-licence-terms-and-conditions).
+#'
+#' @author Antonio Paez, \email{paezha@@mcmaster.ca}
+#' @docType package
+#' @name vaccHamilton
+NULL
+
+#' Hamilton CMA Dissemination Areas with Selected Census Information.
+#'
+#' Boundaries of the Dissemination Areas in Hamilton CMA.
+#'
+#' @format A simple features data frame with 1,199 rows and 25 variables:
+#' \describe{
+#'   \item{Population}{Population in the DA}
+#'   \item{Households}{Number of households in DA}
+#'   \item{GeoUID}{Unique identifier of the DA}
+#'   \item{Type}{Geography type}
+#'   \item{Shape Area}{Area of the shape}
+#'   \item{Dwellings}{Number of dwellings in DA}
+#'   \item{CSD_UID}{Census Subdivision (CSD) unique identifier}
+#'   \item{CT_UID}{Census Tract (CT) unique identifier}
+#'   \item{CMA_UID}{Census Metropolitan Area (CMA) unique identifier}
+#'   \item{Region Name}{Name of the region within the CMA}
+#'   \item{Area (sq km)}{Area of the DA in square kilometers}
+#'   \item{v_CA16_190: 50 to 54 years}{Population age 50-54 in the DA}
+#'   \item{v_CA16_208: 55 to 59 years}{Population age 55-59 in the DA}
+#'   \item{v_CA16_226: 60 to 64 years}{Population age 60-64 in the DA}
+#'   \item{v_CA16_247: 65 to 69 years}{Population age 65-69 in the DA}
+#'   \item{v_CA16_265: 70 to 74 years}{Population age 70-74 in the DA}
+#'   \item{v_CA16_283: 75 to 79 years}{Population age 75-79 in the DA}
+#'   \item{v_CA16_301: 80 to 84 years}{Population age 80-84 in the DA}
+#'   \item{v_CA16_319: 85 years and over}{Population 85 years and older in the DA}
[…]

### 2021-04-05 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.Rmd
Migrated analysis to paper, drafted abstract and introductory section

@@
 ---
-title: "An examination of accessibility to pilot COVID-19 vaccination sites in Hamilton, Ontario"
+title: "An examination of the accessibility implications of a pilot COVID-19 vaccination program in Hamilton, Ontario"
 author:
@@
 abstract: |
-  Abstract
+  The province of Ontario in Canada announced the pilot for a new vaccination program, with designated pharmacies across the province now able to offer COVID-19 vaccines.
+The accessibility of this program raises questions about the cost of travel and the distribution of the cost among the population.
+In our examination of the City of Hamilton we find that selected sites do not serve well the rural and urban population of Hamilton, and that the associated cost of travel is expected to be disproportionally borne by lower income populations.
+Modest additions to the list of pilot sites in the city can substantially alleviate this inequity.
 
@@
 ```{r install-data-package, include = FALSE}
-# Run only once if needed to install data package `sobiEquity`
+# Run only once if needed to install data package `vaccHamilton`
 if (!require("vaccHamilton", character.only = TRUE)) {
-      devtools::install_github("https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines", subdir = "vaccHamilton")
+      devtools::install_github("https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines", 
+                               subdir = "vaccHamilton")
   }
@@
 #Load the packages needed to read and work with the data
-#library(circlize)
 library(grid)
@@
 library(png)
-#library(readr)
 library(sf)
@@
 
+```{r invoke-data, include=FALSE}
+data("data_da_2016")
+data("dwelling_network_points_2016")
+data("hamilton_cma")
+data("modes_55to69")
+data("pharmacy_locations")
+data("ttm_car")
+data("ttm_transit")
+data("ttm_walk")
+data("urban_types")
+```
+
+```{r population-age-55-plus, include=FALSE}
+# Data preparation
+## Calculate number of people age 55+ per DA:
+data_da_2016 <- data_da_2016 %>%
+  filter(`Region Name` == "Hamilton") %>%
+  mutate(Population_55to69 = (`v_CA16_208: 55 to 59 years` + 
+                  `v_CA16_226: 60 to 64 years` + 
+                  `v_CA16_247: 65 to 69 years`),
+         Population_70plus = (`v_CA16_265: 70 to 74 years` +
+                  `v_CA16_283: 75 to 79 years` +
+                  `v_CA16_301: 80 to 84 years` + 
+                  `v_CA16_319: 85 years and over`))
+```
+
+```{r select-pharmacies, include=FALSE}
+# Data preparation
+## Select pharmacy locations:
+pharmacy_locations <- pharmacy_locations %>%
+  slice(1:23)
+
+## Also filter selected pharmacies from travel time tables:
+ttm_car <- ttm_car %>%
+  dplyr::filter(OBJECTID <= 23)
+
+ttm_transit <- ttm_transit %>%
+  dplyr::filter(OBJECTID <= 23)
+
+ttm_walk <- ttm_walk %>%
+  dplyr::filter(OBJECTID <= 23)
+```
+
+```{r clip-da-geometry, include=FALSE}
+# Data preparation
+## This is purely cosmetic: clip `data_da_2016`:
+data_da_2016 <- data_da_2016 %>%
+  st_intersection(hamilton_cma)
+```
+
+```{r number-of-residential-units-per-DA, include=FALSE}
+# Data preparation
+## Find number of residential units in each DA. This is needed to assign the population to parcels
+residential_units <- dwelling_network_points_2016 %>%
+  st_drop_geometry() %>%
+  group_by(DAUID) %>%
+  summarize(residential_units = sum(Dwellings))
+```
+
+```{r join-population-to-parcels, include=FALSE}
+# Data preparation
+## Join number of people age 55+ per DA to dwellings and total number of residential units:
+dwelling_network_points_2016 <- dwelling_network_points_2016 %>%
+  left_join(data_da_2016 %>%
+              st_drop_geometry() %>%
+              dplyr::transmute(DAUID = GeoUID, 
+                               Population_55to69,
+                               Population_70plus),
+            by = "DAUID") %>%
+  left_join(residential_units,
+            by = "DAUID")
+```
+
+```{r calculate-population-per-parcel, include=FALSE}
+# Data preparation
+##Calculate population per point on the network.
+This is the population in the DA divided by the total number of residential units.
+This gives the average number of people per residential unit.
+Then, multiply by the number of dwellings (residential units) at the point:
+dwelling_network_points_2016 <- dwelling_network_points_2016 %>%
+  mutate(Population_55to69 = (Population_55to69/residential_units) * Dwellings,
+         Population_70plus = (Population_70plus/residential_units) * Dwellings)
+```
+
+```{r join-region-to-parcels, include=FALSE}
+dwelling_network_points_2016 <- dwelling_network_points_2016 %>%
+  st_join(urban_types)
+```
+
+```{r calculate-population-statistics, include=FALSE}
+population_statistics <- dwelling_network_points_2016 %>% 
+  st_drop_geometry() %>% 
+  dplyr::filter(!is.na(Type)) %>% 
+  group_by(Type) %>% 
+  summarize(Population_55to69 = sum(Population_55to69, 
+                                    na.rm = TRUE), 
+            Population_70plus = sum(Population_70plus, 
+                                    na.rm = TRUE))
+```
+
+```{r join-population-data-to-travel-time-tables, include=FALSE}
+# Data preparation
+## Join the population data to the travel time matrices:
+ttm_car <- ttm_car %>%
+  left_join(dwelling_network_points_2016 %>%
+              st_drop_geometry() %>%
+              dplyr::transmute(ID, 
+                            Population_55to69,
+                            Population_70plus,
+                            Urban_type = Type,
+                            DAUID, 
+                            TAZUID),
+            by = c("UID" = "ID"))
+
+ttm_transit <- ttm_transit %>%
+  left_join(dwelling_network_points_2016 %>%
+              st_drop_geometry() %>%
+              dplyr::transmute(ID, 
+                            Population_55to69,
+                            Population_70plus,
+                            Urban_type = Type,
+                            DAUID, 
+                            TAZUID),
+            by = c("UID" = "ID"))
+
+ttm_walk <- ttm_walk %>%
+  left_join(dwelling_network_points_2016 %>%
+              st_drop_geometry() %>%
+              dplyr::transmute(ID, 
+                            Population_55to69,
+                            Population_70plus,
+                            Urban_type = Type,
+                            DAUID, 
+                            TAZUID),
+            by = c("UID" = "ID"))
+```
+
+```{r join-pharmacy-data-to-travel-time-tables, include=FALSE}
+# Data preparation
+## Join the pharmacy data to the travel time matrices:
+ttm_car <- ttm_car %>%
+  left_join(pharmacy_locations %>%
+              st_drop_geometry() %>%
+              dplyr::select(ID, 
+                            Type),
+            by = c("OBJECTID" = "ID"))
+
+ttm_transit <- ttm_transit %>%
+  left_join(pharmacy_locations %>%
+              st_drop_geometry() %>%
+              dplyr::select(ID, 
+                            Type),
+            by = c("OBJECTID" = "ID"))
+
+ttm_walk <- ttm_walk %>%
+  left_join(pharmacy_locations %>%
+              st_drop_geometry() %>%
+              dplyr::select(ID, 
+                            Type),
+            by = c("OBJECTID" = "ID"))
+```
+
+```{r drop-missing-values, include=FALSE}
+# Data preparation
+## Drop parcels with missing values:
+ttm_car <- ttm_car %>%
+  drop_na(TAZUID,
+          Population_55to69)
+
+ttm_transit <- ttm_transit %>%
+  drop_na(TAZUID,
+          Population_55to69)
+
+ttm_walk <- ttm_walk %>%
+  drop_na(TAZUID,
+          Population_55to69)
+```
+
+```{r proportion-of-trips-by-mode, include=FALSE}
+# Data preparation
+## Calculate proportion of trips by mode per TAZ:
+modes <- modes_55to69 %>%
+  mutate(p_car = Driver/(Driver + Transit + Walk),
+         p_transit = Transit/(Driver + Transit + Walk),
+         p_walk = Walk/(Driver + Transit + Walk))
+```
+
+```{r complete-proportion-of-trips-by-mode, include=FALSE}
+# Data preparation
+## Assume that TAZ with NAs have a proportion of car trips of one:
+modes <- modes %>%
+  dplyr::mutate(p_car = replace_na(p_car, 1),
+                p_transit = replace_na(p_transit, 0),
+                p_walk = replace_na(p_walk, 0))
+```
+
+```{r join-proportion-of-trips-to-travel-time-tables, include=FALSE}
+# Data preparation
+## Join proportion of trips by mode to travel time tables:
+ttm_car <- ttm_car %>%
+  left_join(modes %>%
+              st_drop_geometry() %>%
+              transmute(TAZUID,
+                        p_car,
+                        p_transit,
+                        p_walk),
+            by = "TAZUID")
+
+ttm_transit <- ttm_transit %>%
+  left_join(modes %>%
+              st_drop_geometry() %>%
+              transmute(TAZUID,
+                        p_car,
+                        p_transit,
+                        p_walk),
+            by = "TAZUID")
+
+ttm_walk <- ttm_walk %>%
+  left_join(modes %>%
+              st_drop_geometry() %>%
+              transmute(TAZUID,
+                        p_car,
+                        p_transit,
+                        p_walk),
+            by = "TAZUID")
+```
+
+```{r income-quintiles, include=FALSE}
+# Data preparation
+## Calculate quintiles of median household income in Hamilton:
+mhi_q <- data_da_2016 %>%
+  dplyr::filter(`Region Name` == "Hamilton") %>%
+  pull(`v_CA16_2397: Median total income of households in 2015 ($)`) %>%
+  quantile(c(0, 0.2, 0.4, 0.6, 0.8, 1), 
+           na.rm = TRUE)
+```
+
+```{r label-income-quintiles, include=FALSE}
+# Data preparation
+## Label the DAs according to income quintiles
+data_da_2016 <- data_da_2016 %>%
+  rename(Median_household_income = `v_CA16_2397: Median total income of households in 2015 ($)`) %>%
+  mutate(income_quintile = case_when(Median_household_income < mhi_q[2] ~ "Bottom 20%",
+                                     Median_household_income >= mhi_q[2] & Median_household_income < mhi_q[3] ~ "Fourth 20%",
+                                     Median_household_income >= mhi_q[3] & Median_household_income < mhi_q[4] ~ "Third 20%",
+                                     Median_household_income >= mhi_q[4] & Median_household_income < mhi_q[5] ~ "Second 20%",
+                                     Median_household_income >= mhi_q[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile,
+                                  levels = c("Top 20%",
+                                             "Second 20%",
+                                             "Third 20%",
+                                             "Fourth 20%",
+                                             "Bottom 20%"),
+                                  ordered = TRUE))
+```
+
 Research Questions and Hypotheses
@@
 
-Words.
+Along with the provision of health care facilities to treat severe cases of COVID-19 [@Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs.
+The Province of Ontario, in Canada, announced on April 1st 2021 the expansion of a pilot program to offer vaccines in pharmacies in the City of Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
+This program is in addition to dedicated vaccination centers for people aged 70+.
+Twenty pharmacies in Hamilton were added to an earlier list of 325 locations in other cities across the province, and the program was extended to people aged 55 years old and over.
 
-The objective of this paper is to investigate...Using data...we pose the following questions:
+Critics were quick to point out that the list of pharmacies approved for Hamilton were mostly located in lower density parts of the city that are not well serviced by transit and are difficult to reach by foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}.
+Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast majority of the pharmacies are in suburban Hamilton.
+The issue is somewhat less clear cut when we consider that Hamilton's suburbs tend to be older (see Figure \ref{fig:population-map}).
+The population aged 55 to 69 in Hamilton is approximately `r prettyNum(round(population_statistics$Population_55to69[2]), big.mark = ",")` suburban, `r prettyNum(round(population_statistics$Population_55to69[3]), big.mark = ",")` urban, and only `r prettyNum(round(population_statistics$Population_55to69[3]), big.mark = ",")` rural.
+Given the target demographic for the program, it is possible that suburban sites could be convenient for mature and older adults.
+Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the normally inert Major of the city was caught flat footed by the announcement; see:\url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
+As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may introduce inequities in access.
 
-- Question 1
-- Question 2
+In this research, we investigate the accessibility implications of the sites selected for the pilot vaccination program. Concretely, we ask:
 
-This paper is a reproducible research document [see @Brunsdon2020opening]; the code and data necessary to reproduce the tables and figures are available in a public repository\footnote{\url{https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines}}
+- What is the estimated cost of travel to reach the vaccination sites, assuming that every person requires a vaccine?
+- What is the distribution of this cost across the population of the city?
+- How does the cost and its distribution change with the addition of candidate sites in urban Hamilton?
+
+We concentrate on the 55 to 69 years old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pilot.
+
+```{r pharmacies-map, echo=FALSE, fig.height=5, fig.align="center", fig.cap="\\label{fig:pharmacies-and-regions}Location of pharmacies in pilot and regions with the City of Hamilton"}
+#map_pharmacies <-
+ggplot() +
+  geom_sf(data = urban_types,
+          aes(fill = Type),
+          color = "lightgray") +
+  geom_sf(data = pharmacy_locations %>%
+            filter(Type == "Pilot"),
+          shape = 17,
+          size = 2.5) +
+  scale_color_manual(values = c("Candidate" = "black",
+                                "Pilot" = "black")) +
+  scale_fill_manual(values = c("Urban" = "firebrick2",
+                               "Suburban" = "cornflowerblue",
+                               "Rural" = "forestgreen")) +
+  guides(fill = guide_legend(title = "")) + 
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+```
+
+```{r population-map, echo=FALSE, warning=FALSE, fig.height=7, fig.align="center", fig.cap="\\label{fig:population-map}Distribution of population age 55+ in the City of Hamilton"}
+ggplot() +
+  geom_sf(data = data_da_2016 %>%
+            filter(`Region Name` == "Hamilton") %>%
+            dplyr::select(Population_55to69, Population_70plus) %>%
+            pivot_longer(cols = -geometry, names_to = "Age", values_to = "Population") %>%
+            mutate(Age = factor(Age,
+                                levels = c("Population_55to69",
+                                           "Population_70plus"),
+                                labels = c("Age: 55 to 69",
+                                           "Age: 70+"),
+                                ordered = TRUE)) %>%
+            st_as_sf(),
+          aes(fill = Population),
+          color = NA) +
+  geom_sf(data = urban_types,
+          aes(color = Type),
+          fill = NA) +
+  scale_fill_distiller(name = "Population",
+                       breaks = c(20, 50, 150, 400),
+                       palette = "OrRd", 
+                       trans = "log",
+                       direction = 1) +
+  scale_color_manual(name = "",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen"))  +
+  facet_wrap(~ Age, nrow = 2) +
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+```
 
@@
 
-Words.
+This paper is a reproducible research document [see @Brunsdon2020opening] conducted using open source tools for transportation analysis [@Lovelace2021open].
+All code and data necessary to reproduce the tables and figures are available in a public repository\footnote{\url{https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines}}.
 
@@
 
+```{r minimum-travel-time-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Retrieve the minimum travel time from a point to a pharmacy by each mode for people aged 55 to 69:
+min_ttm_car <- ttm_car %>%
+  dplyr::filter(Type == "Pilot") %>%
+  group_by(UID) %>%
+  summarize(travel_time = min(travel_time),
+            Population = first(Population_55to69),
+            Urban_type = first(Urban_type),
+            DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            p_car = first(p_car),
+            p_transit = first(p_transit),
+            p_walk = first(p_walk),
+            .groups = "drop")
+
+min_ttm_transit <- ttm_transit %>%
+  dplyr::filter(Type == "Pilot") %>%
+  group_by(UID) %>%
+  summarize(travel_time = min(travel_time),
+            Population = first(Population_55to69),
+            Urban_type = first(Urban_type),
+            DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            p_car = first(p_car),
+            p_transit = first(p_transit),
+            p_walk = first(p_walk),
+            .groups = "drop")
+
+min_ttm_walk <- ttm_walk %>%
+  dplyr::filter(Type == "Pilot") %>%
+  group_by(UID) %>%
+  summarize(travel_time = min(travel_time),
+            Population = first(Population_55to69),
+            Urban_type = first(Urban_type),
+            DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            p_car = first(p_car),
+            p_transit = first(p_transit),
+            p_walk = first(p_walk),
+            .groups = "drop")
+```
+
+```{r weighted-travel-time-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Calculate weighted travel time:
+min_ttm_car <- min_ttm_car %>%
+  mutate(weighted_travel_time = travel_time * p_car)
+
+min_ttm_transit <- min_ttm_transit %>%
+  mutate(weighted_travel_time = travel_time * p_transit)
+
+min_ttm_walk <- min_ttm_walk %>%
+  mutate(weighted_travel_time = travel_time * p_walk)
+```
+
+```{r bind-travel-time-tables-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Bind the three tables:
+min_ttm <- rbind(min_ttm_car,
+                 min_ttm_transit,
+                 min_ttm_walk)
+```
+
+```{r aggregate-weighted-travel-times-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Aggregate weighted travel times by mode:
+min_ttm <- min_ttm %>%
+  group_by(UID) %>%
+  summarize(DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            Population = first(Population),
+            Urban_type = first(Urban_type),
+            weighted_travel_time = sum(weighted_travel_time))
+```
+
+```{r calculate-person-hours-travel-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Calculate person-hours of travel:
+min_ttm <- min_ttm %>%
+  mutate(weighted_person_hours = Population * weighted_travel_time/60)
+```
+
+```{r summarize-results-by-taz-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Summarize by TAZ and join geometry:
+min_ttm_taz <- min_ttm %>%
+  group_by(TAZUID) %>%
+  summarize(weighted_travel_time = mean(weighted_travel_time),
+            weighted_person_hours = sum(weighted_person_hours)) %>%
+  left_join(modes %>%
+              dplyr::select(TAZUID),
+            by = "TAZUID") %>%
+  st_as_sf()
+```
+
+```{r include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Map weighted travel time:
+ggplot() +
+  geom_sf(data = min_ttm_taz,
+          aes(fill = weighted_travel_time),
+          color = NA) +
+  geom_sf(data = hamilton_cma,
+          fill = NA) +
+  scale_fill_distiller(palette = "OrRd", 
+                       direction = 1)
+```
+
+```{r include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Map weighted person-hours of travel:
+ggplot() +
+  geom_sf(data = min_ttm_taz,
+          aes(fill = weighted_person_hours),
+          color = NA) +
+  geom_sf(data = hamilton_cma,
+          fill = NA) +
+  scale_fill_distiller(palette = "OrRd", 
+                       direction = 1)
+```
+
+```{r join-income-information-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Join median household information to table:
+min_ttm <- min_ttm %>%
+  left_join(data_da_2016 %>%
+              st_drop_geometry() %>%
+              dplyr::transmute(DAUID = GeoUID,
+                               income_quintile),
+            by = "DAUID")
+```
+
+```{r plot-distribution-travel-time-by-income-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Distribution of weighted travel time by income:
+ttm_income_baseline <- ggplot(data = min_ttm, 
+       aes(x = weighted_travel_time, 
+           color= income_quintile)) +
+  geom_density(size = 0.5, adjust = 2.5) +
+  labs(x = "Travel time (in minutes; weighted by mode)") +
+  scale_color_manual(values = c("Top 20%" = "blue",
+                                "Second 20%" = "steelblue1",
+                                "Third 20%" = "gray40", 
+                                "Fourth 20%" = "tomato",
+                                "Bottom 20%" = "red")) +
+  theme_minimal() +
+  theme(text = element_text(size = 7),
+        legend.text = element_text(size = 7))
+```
+
+```{r plot-distribution-travel-time-by-region-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Distribution of weighted travel time by region:
+ttm_region_baseline <- ggplot(data = min_ttm %>%
+                                drop_na(Urban_type), 
+       aes(x = weighted_travel_time, 
+           color= Urban_type)) +
+  geom_density(size = 0.5, adjust = 3) +
+  labs(x = "Travel time (in minutes; weighted by mode)") +
+  scale_color_manual(name = "",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen"))  +
+  theme_minimal() +
+  theme(text = element_text(size = 7),
+        legend.text = element_text(size = 7))
+```
+
+```{r total-weighted-person-hours-by-group-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Total weighted person-hours by income:
+person_hours_income_baseline <- min_ttm %>%
+  group_by(income_quintile) %>%
+  summarize(total_weighted_person_hours = sum(weighted_person_hours),
+            total_population = sum(Population),
+            .groups = "drop") %>%
+  mutate(ratio = total_weighted_person_hours/total_population) %>%
+  rename(group = income_quintile)
+
+## Total weighted person-hours by region:
+person_hours_region_baseline <- min_ttm %>%
+  group_by(Urban_type) %>%
+  summarize(total_weighted_person_hours = sum(weighted_person_hours),
+            total_population = sum(Population),
+            .groups = "drop") %>%
+  drop_na() %>%
+  mutate(ratio = total_weighted_person_hours/total_population) %>%
+  rename(group = Urban_type)
+
+person_hours_baseline <- rbind(person_hours_income_baseline,
+                               person_hours_region_baseline)
+```
+
+```{r minimum-travel-time, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Retrieve the minimum travel time from a point to a pharmacy by each mode for people aged 55 to 69:
+min_ttm_car <- ttm_car %>%
+  #dplyr::filter(Type == "Pilot") %>%
+  group_by(UID) %>%
+  summarize(travel_time = min(travel_time),
+            Population = first(Population_55to69),
+            Urban_type = first(Urban_type),
+            DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            p_car = first(p_car),
+            p_transit = first(p_transit),
+            p_walk = first(p_walk),
+            .groups = "drop")
+
+min_ttm_transit <- ttm_transit %>%
+  #dplyr::filter(Type == "Pilot") %>%
+  group_by(UID) %>%
+  summarize(travel_time = min(travel_time),
+            Population = first(Population_55to69),
+            Urban_type = first(Urban_type),
+            DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            p_car = first(p_car),
+            p_transit = first(p_transit),
+            p_walk = first(p_walk),
+            .groups = "drop")
+
+min_ttm_walk <- ttm_walk %>%
+  #dplyr::filter(Type == "Pilot") %>%
+  group_by(UID) %>%
+  summarize(travel_time = min(travel_time),
+            Population = first(Population_55to69),
+            Urban_type = first(Urban_type),
+            DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            p_car = first(p_car),
+            p_transit = first(p_transit),
+            p_walk = first(p_walk),
+            .groups = "drop")
+```
+
+```{r weighted-travel-time, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Calculate weighted travel time:
+min_ttm_car <- min_ttm_car %>%
+  mutate(weighted_travel_time = travel_time * p_car)
+
+min_ttm_transit <- min_ttm_transit %>%
+  mutate(weighted_travel_time = travel_time * p_transit)
+
+min_ttm_walk <- min_ttm_walk %>%
+  mutate(weighted_travel_time = travel_time * p_walk)
+```
+
+```{r bind-travel-time-tables, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Bind the three tables:
+min_ttm <- rbind(min_ttm_car,
+                 min_ttm_transit,
+                 min_ttm_walk)
+```
+
+```{r aggregate-weighted-travel-times, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Aggregate weighted travel times by mode:
+min_ttm <- min_ttm %>%
+  group_by(UID) %>%
+  summarize(DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            Population = first(Population),
+            Urban_type = first(Urban_type),
+            weighted_travel_time = sum(weighted_travel_time))
+```
+
+```{r calculate-person-hours-travel, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Calculate person-hours of travel:
+min_ttm <- min_ttm %>%
+  mutate(weighted_person_hours = Population * weighted_travel_time/60)
+```
+
+```{r summarize-results-by-taz, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Summarize by TAZ and join geometry:
+min_ttm_taz <- min_ttm %>%
+  group_by(TAZUID) %>%
+  summarize(weighted_travel_time = mean(weighted_travel_time),
+            weighted_person_hours = sum(weighted_person_hours)) %>%
+  left_join(modes %>%
+              dplyr::select(TAZUID),
+            by = "TAZUID") %>%
+  st_as_sf()
+```
+
+```{r include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Map weighted travel time:
+ggplot() +
+  geom_sf(data = min_ttm_taz,
+          aes(fill = weighted_travel_time),
+          color = NA) +
+  geom_sf(data = hamilton_cma,
+          fill = NA) +
+  scale_fill_distiller(palette = "OrRd", 
+                       direction = 1)
+```
+
+```{r include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Map weighted person-hours of travel:
+ggplot() +
+  geom_sf(data = min_ttm_taz,
+          aes(fill = weighted_person_hours),
+          color = NA) +
+  geom_sf(data = hamilton_cma,
+          fill = NA) +
+  scale_fill_distiller(palette = "OrRd", 
+                       direction = 1)
+```
+
+```{r join-income-information, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Join median household information to table:
+min_ttm <- min_ttm %>%
+  left_join(data_da_2016 %>%
+              st_drop_geometry() %>%
+              dplyr::transmute(DAUID = GeoUID,
+                               income_quintile),
+            by = "DAUID")
+```
+
+```{r plot-distribution-travel-time-by-income, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Distribution of weighted travel time by income:
+ttm_income <- ggplot(data = min_ttm, 
+       aes(x = weighted_travel_time, 
+           color= income_quintile)) +
+  geom_density(size = 0.5, adjust = 2.5) +
+  labs(x = "Travel time (in minutes; weighted by mode)") +
+  scale_color_manual(name = "Income quintile",
+                     values = c("Top 20%" = "blue",
+                                "Second 20%" = "steelblue1",
+                                "Third 20%" = "gray40", 
+                                "Fourth 20%" = "tomato",
+                                "Bottom 20%" = "red")) +
+  theme_minimal() +
+  theme(text = element_text(size = 7),
+        legend.text = element_text(size = 7))
+```
+
+```{r plot-distribution-travel-time-by-region, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Distribution of weighted travel time by region:
+ttm_region <- ggplot(data = min_ttm %>%
+                                drop_na(Urban_type), 
+       aes(x = weighted_travel_time, 
+           color= Urban_type)) +
+  geom_density(size = 0.5, adjust = 3) +
+  labs(x = "Travel time (in minutes; weighted by mode)") +
+  scale_color_manual(name = "Region",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen"))  +
+  theme_minimal() +
+  theme(text = element_text(size = 7),
+        legend.text = element_text(size = 7))
+```
+
+```{r total-weighted-person-hours-by-group, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Total weighted person-hours by income:
+person_hours_income <- min_ttm %>%
+  group_by(income_quintile) %>%
+  summarize(total_weighted_person_hours = sum(weighted_person_hours),
+            total_population = sum(Population),
+            .groups = "drop") %>%
+  mutate(ratio = total_weighted_person_hours/total_population) %>%
+  rename(group = income_quintile)
+
+## Total weighted person-hours by region:
+person_hours_region <- min_ttm %>%
+  group_by(Urban_type) %>%
+  summarize(total_weighted_person_hours = sum(weighted_person_hours),
+            total_population = sum(Population),
+            .groups = "drop") %>%
+  drop_na() %>%
+  mutate(ratio = total_weighted_person_hours/total_population) %>%
+  rename(group = Urban_type)
+
+person_hours <- rbind(person_hours_income,
+                      person_hours_region)
+```
+
 Words.
+
+```{r figure-results, echo=FALSE, warning=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:results}Distribution of travel time (weighted by mode) for different population groups"}
+grid.arrange(ttm_income_baseline + 
+               ggtitle("By income: Pilot Program") + 
+               labs(x = "") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8), 
+                     axis.text = element_text(size = 7), 
+                     legend.position = "none"), 
+             ttm_income + 
+               ggtitle("By income: Scenario") + 
+               labs(x = "", y = "") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8), 
+                     axis.text = element_text(size = 7)), 
+             ttm_region_baseline + 
+               ggtitle("By region: Pilot Program") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8),
+                     axis.title = element_text(size = 8), 
+                     axis.text = element_text(size = 7), 
+                     legend.position = "none"), 
+             ttm_region + 
+               ggtitle("By region: Scenario") + 
+               labs(y = "") +
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8),
+                     axis.title = element_text(size = 8), 
+                     axis.text = element_text(size = 7)), 
+             nrow = 2)
+```
+
+```{r table-results, echo=FALSE}
+cbind(person_hours_baseline %>%
+        dplyr::select(group, total_population, total_weighted_person_hours, ratio), 
+      person_hours %>%
+        dplyr::select(total_weighted_person_hours, ratio)) %>%
+  kable("latex",
+        booktabs = TRUE,
+        col.names = c("Group", 
+                      "Population", 
+                      "Total PHT", 
+                      "Hours per person", 
+                      "Total PHT",
+                      "Hours per person"),
+        digits = 3,
+        align = c("lccccc"),
+        caption = "\\label{tab:distribution-results}Distribution of person hours travelled (PHT) by median total household income and region: pilot locations only, and scenario with three urban locations added") %>%
+  add_header_above(c(" " = 2, "Pilot Program" = 2, "Scenario" = 2)) %>%
+  kable_styling(latex_options = c("striped", "scale_down")) %>%
+  pack_rows("Income Quintile", 1, 5) %>%
+  pack_rows("Region", 6, 8) %>%
+  footnote(general = "The population totals differ due to small differences in the classification of the regions")
+```
 

### 2021-04-05 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.tex
Migrated analysis to paper, drafted abstract and introductory section

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{Transport Findings} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={An examination of the accessibility implications of a pilot COVID-19 vaccination program in Hamilton, Ontario},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

% Pandoc citation processing

% Pandoc header
\usepackage{float} \usepackage{xcolor} \floatplacement{figure}{H}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\begin{frontmatter}

  \title{An examination of the accessibility implications of a pilot COVID-19
vaccination program in Hamilton, Ontario}
    \author[McMaster University]{Antonio Paez\corref{1}}
   \ead{paezha@mcmaster.ca} 
    \author[University of Toronto Scarborough]{Christopher D. Higgins}
   \ead{cd.higgins@utoronto.ca} 
      \address[McMaster University]{School of Earth, Environment and Society, McMaster University, Hamilton,
ON, L8S 4K1, Canada}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, University of Toronto Scarborough,
1265 Military Trail, Toronto, ON M1C1A4}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  The province of Ontario in Canada announced the pilot for a new
  vaccination program, with designated pharmacies across the province now
  able to offer COVID-19 vaccines. The accessibility of this program
  raises questions about the cost of travel and the distribution of the
  cost among the population. In our examination of the City of Hamilton we
  find that selected sites do not serve well the rural and urban
  population of Hamilton, and that the associated cost of travel is
  expected to be disproportionally borne by lower income populations.
  Modest additions to the list of pilot sites in the city can
  substantially alleviate this inequity.
  \end{abstract}

 \end{frontmatter}

\hypertarget{research-questions-and-hypotheses}{%
\section{Research Questions and
Hypotheses}\label{research-questions-and-hypotheses}}

Along with the provision of health care facilities to treat severe cases
of COVID-19 (Pereira et al., 2021), another front in the fight against
the pandemic is the rolling out of vaccination programs. The Province of
Ontario, in Canada, announced on April 1st 2021 the expansion of a pilot
program to offer vaccines in pharmacies in the City of
Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
This program is in addition to dedicated vaccination centers for people
aged 70+. Twenty pharmacies in Hamilton were added to an earlier list of
325 locations in other cities across the province, and the program was
extended to people aged 55 years old and over.

Critics were quick to point out that the list of pharmacies approved for
Hamilton were mostly located in lower density parts of the city that are
not well serviced by transit and are difficult to reach by
foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}.
Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast
majority of the pharmacies are in suburban Hamilton. The issue is
somewhat less clear cut when we consider that Hamilton's suburbs tend to
be older (see Figure \ref{fig:population-map}). The population aged 55
to 69 in Hamilton is approximately 59,095 suburban, 35,704 urban, and
only 35,704 rural. Given the target demographic for the program, it is
possible that suburban sites could be convenient for mature and older
adults. Nevertheless, the selection of sites by the province raises some
important
questions\footnote{The decision-making process to select these sites appears to have been opaque, and the normally inert Major of the city was caught flat footed by the announcement; see:\url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
As Yu et al. (2021) note, good geographical coverage is a key element
for a successful vaccination campaign; at the same time, siting
vaccinations sites in car-oriented locations may introduce inequities in
access.

In this research, we investigate the accessibility implications of the
sites selected for the pilot vaccination program. Concretely, we ask:

\begin{itemize}
\tightlist
\item
  What is the estimated cost of travel to reach the vaccination sites,
  assuming that every person requires a vaccine?
\item
  What is the distribution of this cost across the population of the
  city?
\item
  How does the cost and its distribution change with the addition of
  candidate sites in urban Hamilton?
\end{itemize}

We concentrate on the 55 to 69 years old population segment because the
older 70+ group have access to other dedicated facilities besides those
in the provincial pilot.

\begin{figure}

{\centering \includegraphics{Accessibility-Vaccination-Sites-Hamilton_files/figure-latex/pharmacies-map-1}

}

\caption{\label{fig:pharmacies-and-regions}Location of pharmacies in pilot and regions with the City of Hamilton}\label{fig:pharmacies-map}
\end{figure}

\begin{figure}

{\centering \includegraphics{Accessibility-Vaccination-Sites-Hamilton_files/figure-latex/population-map-1}

}

\caption{\label{fig:population-map}Distribution of population age 55+ in the City of Hamilton}\label{fig:population-map}
\end{figure}

\hypertarget{methods-and-data}{%
\section{Methods and Data}\label{methods-and-data}}

This paper is a reproducible research document (see Brunsdon and Comber,
2020) conducted using open source tools for transportation analysis
(Lovelace, 2021). All code and data necessary to reproduce the tables
and figures are available in a public
repository\footnote{\url{https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines}}.

\hypertarget{findings}{%
\section{Findings}\label{findings}}

Words.

\begin{figure}

{\centering \includegraphics{Accessibility-Vaccination-Sites-Hamilton_files/figure-latex/figure-results-1}

}

\caption{\label{fig:results}Distribution of travel time (weighted by mode) for different population groups}\label{fig:figure-results}
\end{figure}

\begin{table}

\caption{\label{tab:table-results}\label{tab:distribution-results}Distribution of person hours travelled (PHT) by median total household income and region: pilot locations only, and scenario with three urban locations added}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lccccc}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{2}{c}{Pilot Program} & \multicolumn{2}{c}{Scenario} \\
\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6}
Group & Population & Total PHT & Hours per person & Total PHT & Hours per person\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Income Quintile}}\\
\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{23297.315} & \cellcolor{gray!6}{2243.857} & \cellcolor{gray!6}{0.096} & \cellcolor{gray!6}{2146.558} & \cellcolor{gray!6}{0.092}\\
\hspace{1em}Second 20\% & 22356.413 & 2471.952 & 0.111 & 2351.858 & 0.105\\
\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{19570.061} & \cellcolor{gray!6}{1749.497} & \cellcolor{gray!6}{0.089} & \cellcolor{gray!6}{1563.978} & \cellcolor{gray!6}{0.080}\\
\hspace{1em}Fourth 20\% & 17729.139 & 2928.959 & 0.165 & 1950.312 & 0.110\\
\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{19629.952} & \cellcolor{gray!6}{4068.548} & \cellcolor{gray!6}{0.207} & \cellcolor{gray!6}{2388.422} & \cellcolor{gray!6}{0.122}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Region}}\\
\hspace{1em}Rural & 8356.963 & 1730.268 & 0.207 & 1730.242 & 0.207\\
\hspace{1em}\cellcolor{gray!6}{Suburban} & \cellcolor{gray!6}{58711.629} & \cellcolor{gray!6}{4138.482} & \cellcolor{gray!6}{0.070} & \cellcolor{gray!6}{4138.392} & \cellcolor{gray!6}{0.070}\\
\hspace{1em}Urban & 35491.942 & 7588.590 & 0.214 & 4527.021 & 0.128\\
\bottomrule
\multicolumn{6}{l}{\rule{0pt}{1em}\textit{Note: }}\\
\multicolumn{6}{l}{\rule{0pt}{1em}The population totals differ due to small differences in the classification of the regions}\\
\end{tabular}}
\end{table}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Brunsdon2020opening}{}%
Brunsdon, C., Comber, A., 2020. Opening practice: Supporting
reproducibility and critical spatial data science. Journal of
Geographical Systems 1--20.

\leavevmode\hypertarget{ref-Lovelace2021open}{}%
Lovelace, R., 2021. Open source tools for geographic analysis in
transport planning. Journal of Geographical Systems.
doi:\href{https://doi.org/10.1007/s10109-020-00342-2}{10.1007/s10109-020-00342-2}

\leavevmode\hypertarget{ref-Pereira2021geographic}{}%
Pereira, R.H.M., Braga, C.K.V., Servo, L.M., Serra, B., Amaral, P.,
Gouveia, N., Paez, A., 2021. Geographic access to covid-19 healthcare in
brazil using a balanced float catchment area approach. Social Science \&
Medicine 273, 113773.
doi:\href{https://doi.org/https://doi.org/10.1016/j.socscimed.2021.113773}{https://doi.org/10.1016/j.socscimed.2021.113773}

\leavevmode\hypertarget{ref-Yu2021sustained}{}%
Yu, J.H., Jeong, H.J., Kim, S.J., Lee, J.Y., Choe, Y.J., Choi, E.H.,
Cho, E.H., 2021. Sustained vaccination coverage during the coronavirus
disease 2019 epidemic in the republic of korea. Vaccines 9, 8.
doi:\href{https://doi.org/10.3390/vaccines9010002}{10.3390/vaccines9010002}

\end{document}

### 2021-04-05 README.Rmd
Migrated analysis to paper, drafted abstract and introductory section

@@
 
-# An examination of accessibility to pilot COVID-19 vaccination sites in Hamilton, Ontario
+# An examination of the accessibility implications of a pilot COVID-19 vaccination program in Hamilton, Ontario
 
@@
 
-Words.
+The province of Ontario in Canada announced the pilot for a new vaccination program, with designated pharmacies across the province now able to offer COVID-19 vaccines.
+The accessibility of this program raises questions about the cost of travel and the distribution of the cost among the population.
+In our examination of the City of Hamilton we find that selected sites do not serve well the rural and urban population of Hamilton, and that the associated cost of travel is expected to be disproportionally borne by lower income populations.
+Modest additions to the list of pilot sites in the city can substantially alleviate this inequity.
 
@@
 - Reproducible research
+
+```{r install-data-package, include = FALSE}
+# Run only once if needed to install data package `vaccHamilton`
+if (!require("vaccHamilton", character.only = TRUE)) {
+      devtools::install_github("https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines", 
+                               subdir = "vaccHamilton")
+  }
+```
+
+```{r load-packages, include=FALSE}
+#Load the packages needed to read and work with the data
+library(grid)
+library(gridExtra)
+library(kableExtra)
+library(png)
+library(sf)
+library(tidyverse)
+library(vaccHamilton)
+```
+
+```{r invoke-data, include=FALSE}
+data("data_da_2016")
+data("dwelling_network_points_2016")
+data("hamilton_cma")
+data("modes_55to69")
+data("pharmacy_locations")
+data("ttm_car")
+data("ttm_transit")
+data("ttm_walk")
+data("urban_types")
+```
+
+```{r population-age-55-plus, include=FALSE}
+# Data preparation
+## Calculate number of people age 55+ per DA:
+data_da_2016 <- data_da_2016 %>%
+  filter(`Region Name` == "Hamilton") %>%
+  mutate(Population_55to69 = (`v_CA16_208: 55 to 59 years` + 
+                  `v_CA16_226: 60 to 64 years` + 
+                  `v_CA16_247: 65 to 69 years`),
+         Population_70plus = (`v_CA16_265: 70 to 74 years` +
+                  `v_CA16_283: 75 to 79 years` +
+                  `v_CA16_301: 80 to 84 years` + 
+                  `v_CA16_319: 85 years and over`))
+```
+
+```{r select-pharmacies, include=FALSE}
+# Data preparation
+## Select pharmacy locations:
+pharmacy_locations <- pharmacy_locations %>%
+  slice(1:23)
+
+## Also filter selected pharmacies from travel time tables:
+ttm_car <- ttm_car %>%
+  dplyr::filter(OBJECTID <= 23)
+
+ttm_transit <- ttm_transit %>%
+  dplyr::filter(OBJECTID <= 23)
+
+ttm_walk <- ttm_walk %>%
+  dplyr::filter(OBJECTID <= 23)
+```
+
+```{r clip-da-geometry, include=FALSE}
+# Data preparation
+## This is purely cosmetic: clip `data_da_2016`:
+data_da_2016 <- data_da_2016 %>%
+  st_intersection(hamilton_cma)
+```
+
+```{r number-of-residential-units-per-DA, include=FALSE}
+# Data preparation
+## Find number of residential units in each DA. This is needed to assign the population to parcels
+residential_units <- dwelling_network_points_2016 %>%
+  st_drop_geometry() %>%
+  group_by(DAUID) %>%
+  summarize(residential_units = sum(Dwellings))
+```
+
+```{r join-population-to-parcels, include=FALSE}
+# Data preparation
+## Join number of people age 55+ per DA to dwellings and total number of residential units:
+dwelling_network_points_2016 <- dwelling_network_points_2016 %>%
+  left_join(data_da_2016 %>%
+              st_drop_geometry() %>%
+              dplyr::transmute(DAUID = GeoUID, 
+                               Population_55to69,
+                               Population_70plus),
+            by = "DAUID") %>%
+  left_join(residential_units,
+            by = "DAUID")
+```
+
+```{r calculate-population-per-parcel, include=FALSE}
+# Data preparation
+##Calculate population per point on the network.
+This is the population in the DA divided by the total number of residential units.
+This gives the average number of people per residential unit.
+Then, multiply by the number of dwellings (residential units) at the point:
+dwelling_network_points_2016 <- dwelling_network_points_2016 %>%
+  mutate(Population_55to69 = (Population_55to69/residential_units) * Dwellings,
+         Population_70plus = (Population_70plus/residential_units) * Dwellings)
+```
+
+```{r join-region-to-parcels, include=FALSE}
+dwelling_network_points_2016 <- dwelling_network_points_2016 %>%
+  st_join(urban_types)
+```
+
+```{r calculate-population-statistics, include=FALSE}
+population_statistics <- dwelling_network_points_2016 %>% 
+  st_drop_geometry() %>% 
+  dplyr::filter(!is.na(Type)) %>% 
+  group_by(Type) %>% 
+  summarize(Population_55to69 = sum(Population_55to69, 
+                                    na.rm = TRUE), 
+            Population_70plus = sum(Population_70plus, 
+                                    na.rm = TRUE))
+```
+
+```{r join-population-data-to-travel-time-tables, include=FALSE}
+# Data preparation
+## Join the population data to the travel time matrices:
+ttm_car <- ttm_car %>%
+  left_join(dwelling_network_points_2016 %>%
+              st_drop_geometry() %>%
+              dplyr::transmute(ID, 
+                            Population_55to69,
+                            Population_70plus,
+                            Urban_type = Type,
+                            DAUID, 
+                            TAZUID),
+            by = c("UID" = "ID"))
+
+ttm_transit <- ttm_transit %>%
+  left_join(dwelling_network_points_2016 %>%
+              st_drop_geometry() %>%
+              dplyr::transmute(ID, 
+                            Population_55to69,
+                            Population_70plus,
+                            Urban_type = Type,
+                            DAUID, 
+                            TAZUID),
+            by = c("UID" = "ID"))
+
+ttm_walk <- ttm_walk %>%
+  left_join(dwelling_network_points_2016 %>%
+              st_drop_geometry() %>%
+              dplyr::transmute(ID, 
+                            Population_55to69,
+                            Population_70plus,
+                            Urban_type = Type,
+                            DAUID, 
+                            TAZUID),
+            by = c("UID" = "ID"))
+```
+
+```{r join-pharmacy-data-to-travel-time-tables, include=FALSE}
+# Data preparation
+## Join the pharmacy data to the travel time matrices:
+ttm_car <- ttm_car %>%
+  left_join(pharmacy_locations %>%
+              st_drop_geometry() %>%
+              dplyr::select(ID, 
+                            Type),
+            by = c("OBJECTID" = "ID"))
+
+ttm_transit <- ttm_transit %>%
+  left_join(pharmacy_locations %>%
+              st_drop_geometry() %>%
+              dplyr::select(ID, 
+                            Type),
+            by = c("OBJECTID" = "ID"))
+
+ttm_walk <- ttm_walk %>%
+  left_join(pharmacy_locations %>%
+              st_drop_geometry() %>%
+              dplyr::select(ID, 
+                            Type),
+            by = c("OBJECTID" = "ID"))
+```
+
+```{r drop-missing-values, include=FALSE}
+# Data preparation
+## Drop parcels with missing values:
+ttm_car <- ttm_car %>%
+  drop_na(TAZUID,
+          Population_55to69)
+
+ttm_transit <- ttm_transit %>%
+  drop_na(TAZUID,
+          Population_55to69)
+
+ttm_walk <- ttm_walk %>%
+  drop_na(TAZUID,
+          Population_55to69)
+```
+
+```{r proportion-of-trips-by-mode, include=FALSE}
+# Data preparation
+## Calculate proportion of trips by mode per TAZ:
+modes <- modes_55to69 %>%
+  mutate(p_car = Driver/(Driver + Transit + Walk),
+         p_transit = Transit/(Driver + Transit + Walk),
+         p_walk = Walk/(Driver + Transit + Walk))
+```
+
+```{r complete-proportion-of-trips-by-mode, include=FALSE}
+# Data preparation
+## Assume that TAZ with NAs have a proportion of car trips of one:
+modes <- modes %>%
+  dplyr::mutate(p_car = replace_na(p_car, 1),
+                p_transit = replace_na(p_transit, 0),
+                p_walk = replace_na(p_walk, 0))
+```
+
+```{r join-proportion-of-trips-to-travel-time-tables, include=FALSE}
+# Data preparation
+## Join proportion of trips by mode to travel time tables:
+ttm_car <- ttm_car %>%
+  left_join(modes %>%
+              st_drop_geometry() %>%
+              transmute(TAZUID,
+                        p_car,
+                        p_transit,
+                        p_walk),
+            by = "TAZUID")
+
+ttm_transit <- ttm_transit %>%
+  left_join(modes %>%
+              st_drop_geometry() %>%
+              transmute(TAZUID,
+                        p_car,
+                        p_transit,
+                        p_walk),
+            by = "TAZUID")
+
+ttm_walk <- ttm_walk %>%
+  left_join(modes %>%
+              st_drop_geometry() %>%
+              transmute(TAZUID,
+                        p_car,
+                        p_transit,
+                        p_walk),
+            by = "TAZUID")
+```
+
+```{r income-quintiles, include=FALSE}
+# Data preparation
+## Calculate quintiles of median household income in Hamilton:
+mhi_q <- data_da_2016 %>%
+  dplyr::filter(`Region Name` == "Hamilton") %>%
+  pull(`v_CA16_2397: Median total income of households in 2015 ($)`) %>%
+  quantile(c(0, 0.2, 0.4, 0.6, 0.8, 1), 
+           na.rm = TRUE)
+```
+
+```{r label-income-quintiles, include=FALSE}
+# Data preparation
+## Label the DAs according to income quintiles
+data_da_2016 <- data_da_2016 %>%
+  rename(Median_household_income = `v_CA16_2397: Median total income of households in 2015 ($)`) %>%
+  mutate(income_quintile = case_when(Median_household_income < mhi_q[2] ~ "Bottom 20%",
+                                     Median_household_income >= mhi_q[2] & Median_household_income < mhi_q[3] ~ "Fourth 20%",
+                                     Median_household_income >= mhi_q[3] & Median_household_income < mhi_q[4] ~ "Third 20%",
+                                     Median_household_income >= mhi_q[4] & Median_household_income < mhi_q[5] ~ "Second 20%",
+                                     Median_household_income >= mhi_q[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile,
+                                  levels = c("Top 20%",
+                                             "Second 20%",
+                                             "Third 20%",
+                                             "Fourth 20%",
+                                             "Bottom 20%"),
+                                  ordered = TRUE))
+```
+
+## Research Questions and Hypotheses
+
+Along with the provision of health care facilities to treat severe cases of COVID-19 [@Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs.
+The Province of Ontario, in Canada, announced on April 1st 2021 the expansion of a pilot program to offer vaccines in pharmacies in the City of Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
+This program is in addition to dedicated vaccination centers for people aged 70+.
+Twenty pharmacies in Hamilton were added to an earlier list of 325 locations in other cities across the province, and the program was extended to people aged 55 years old and over.
+
+Critics were quick to point out that the list of pharmacies approved for Hamilton were mostly located in lower density parts of the city that are not well serviced by transit and are difficult to reach by foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}.
+Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast majority of the pharmacies are in suburban Hamilton.
+The issue is somewhat less clear cut when we consider that Hamilton's suburbs tend to be older (see Figure \ref{fig:population-map}).
+The population aged 55 to 69 in Hamilton is approximately `r prettyNum(round(population_statistics$Population_55to69[2]), big.mark = ",")` suburban, `r prettyNum(round(population_statistics$Population_55to69[3]), big.mark = ",")` urban, and only `r prettyNum(round(population_statistics$Population_55to69[3]), big.mark = ",")` rural.
+Given the target demographic for the program, it is possible that suburban sites could be convenient for mature and older adults.
+Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the normally inert Major of the city was caught flat footed by the announcement; see:\url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
+As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may introduce inequities in access.
+
+In this research, we investigate the accessibility implications of the sites selected for the pilot vaccination program. Concretely, we ask:
+
+- What is the estimated cost of travel to reach the vaccination sites, assuming that every person requires a vaccine?
+- What is the distribution of this cost across the population of the city?
+- How does the cost and its distribution change with the addition of candidate sites in urban Hamilton?
+
+We concentrate on the 55 to 69 years old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pilot.
+
+```{r pharmacies-map, echo=FALSE, fig.height=5, fig.align="center", fig.cap="\\label{fig:pharmacies-and-regions}Location of pharmacies in pilot and regions with the City of Hamilton"}
+#map_pharmacies <-
+ggplot() +
+  geom_sf(data = urban_types,
+          aes(fill = Type),
+          color = "lightgray") +
+  geom_sf(data = pharmacy_locations %>%
+            filter(Type == "Pilot"),
+          shape = 17,
+          size = 2.5) +
+  scale_color_manual(values = c("Candidate" = "black",
+                                "Pilot" = "black")) +
+  scale_fill_manual(values = c("Urban" = "firebrick2",
+                               "Suburban" = "cornflowerblue",
+                               "Rural" = "forestgreen")) +
+  guides(fill = guide_legend(title = "")) + 
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+```
+
+```{r population-map, echo=FALSE, warning=FALSE, fig.height=7, fig.align="center", fig.cap="\\label{fig:population-map}Distribution of population age 55+ in the City of Hamilton"}
+ggplot() +
+  geom_sf(data = data_da_2016 %>%
+            filter(`Region Name` == "Hamilton") %>%
+            dplyr::select(Population_55to69, Population_70plus) %>%
+            pivot_longer(cols = -geometry, names_to = "Age", values_to = "Population") %>%
+            mutate(Age = factor(Age,
+                                levels = c("Population_55to69",
+                                           "Population_70plus"),
+                                labels = c("Age: 55 to 69",
+                                           "Age: 70+"),
+                                ordered = TRUE)) %>%
+            st_as_sf(),
+          aes(fill = Population),
+          color = NA) +
+  geom_sf(data = urban_types,
+          aes(color = Type),
+          fill = NA) +
+  scale_fill_distiller(name = "Population",
+                       breaks = c(20, 50, 150, 400),
+                       palette = "OrRd", 
+                       trans = "log",
+                       direction = 1) +
+  scale_color_manual(name = "",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen"))  +
+  facet_wrap(~ Age, nrow = 2) +
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+```
+
+## Methods and Data
+
+This paper is a reproducible research document [see @Brunsdon2020opening] conducted using open source tools for transportation analysis [@Lovelace2021open].
+All code and data necessary to reproduce the tables and figures are available in a public repository\footnote{\url{https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines}}.
+
+## Findings
+
+```{r minimum-travel-time-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Retrieve the minimum travel time from a point to a pharmacy by each mode for people aged 55 to 69:
+min_ttm_car <- ttm_car %>%
+  dplyr::filter(Type == "Pilot") %>%
+  group_by(UID) %>%
+  summarize(travel_time = min(travel_time),
+            Population = first(Population_55to69),
+            Urban_type = first(Urban_type),
+            DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            p_car = first(p_car),
+            p_transit = first(p_transit),
+            p_walk = first(p_walk),
+            .groups = "drop")
+
+min_ttm_transit <- ttm_transit %>%
+  dplyr::filter(Type == "Pilot") %>%
+  group_by(UID) %>%
+  summarize(travel_time = min(travel_time),
+            Population = first(Population_55to69),
+            Urban_type = first(Urban_type),
+            DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            p_car = first(p_car),
+            p_transit = first(p_transit),
+            p_walk = first(p_walk),
+            .groups = "drop")
+
+min_ttm_walk <- ttm_walk %>%
+  dplyr::filter(Type == "Pilot") %>%
+  group_by(UID) %>%
+  summarize(travel_time = min(travel_time),
+            Population = first(Population_55to69),
+            Urban_type = first(Urban_type),
+            DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            p_car = first(p_car),
+            p_transit = first(p_transit),
+            p_walk = first(p_walk),
+            .groups = "drop")
+```
+
+```{r weighted-travel-time-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Calculate weighted travel time:
+min_ttm_car <- min_ttm_car %>%
+  mutate(weighted_travel_time = travel_time * p_car)
+
+min_ttm_transit <- min_ttm_transit %>%
+  mutate(weighted_travel_time = travel_time * p_transit)
+
+min_ttm_walk <- min_ttm_walk %>%
+  mutate(weighted_travel_time = travel_time * p_walk)
+```
+
+```{r bind-travel-time-tables-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Bind the three tables:
+min_ttm <- rbind(min_ttm_car,
+                 min_ttm_transit,
+                 min_ttm_walk)
+```
+
+```{r aggregate-weighted-travel-times-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Aggregate weighted travel times by mode:
+min_ttm <- min_ttm %>%
+  group_by(UID) %>%
+  summarize(DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            Population = first(Population),
+            Urban_type = first(Urban_type),
+            weighted_travel_time = sum(weighted_travel_time))
+```
+
+```{r calculate-person-hours-travel-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Calculate person-hours of travel:
+min_ttm <- min_ttm %>%
+  mutate(weighted_person_hours = Population * weighted_travel_time/60)
+```
+
+```{r summarize-results-by-taz-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Summarize by TAZ and join geometry:
+min_ttm_taz <- min_ttm %>%
+  group_by(TAZUID) %>%
+  summarize(weighted_travel_time = mean(weighted_travel_time),
+            weighted_person_hours = sum(weighted_person_hours)) %>%
+  left_join(modes %>%
+              dplyr::select(TAZUID),
+            by = "TAZUID") %>%
+  st_as_sf()
+```
+
+```{r include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Map weighted travel time:
+ggplot() +
+  geom_sf(data = min_ttm_taz,
+          aes(fill = weighted_travel_time),
+          color = NA) +
+  geom_sf(data = hamilton_cma,
+          fill = NA) +
+  scale_fill_distiller(palette = "OrRd", 
+                       direction = 1)
+```
+
+```{r include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Map weighted person-hours of travel:
+ggplot() +
+  geom_sf(data = min_ttm_taz,
+          aes(fill = weighted_person_hours),
+          color = NA) +
+  geom_sf(data = hamilton_cma,
+          fill = NA) +
+  scale_fill_distiller(palette = "OrRd", 
+                       direction = 1)
+```
+
+```{r join-income-information-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Join median household information to table:
+min_ttm <- min_ttm %>%
+  left_join(data_da_2016 %>%
+              st_drop_geometry() %>%
+              dplyr::transmute(DAUID = GeoUID,
+                               income_quintile),
+            by = "DAUID")
+```
+
+```{r plot-distribution-travel-time-by-income-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Distribution of weighted travel time by income:
+ttm_income_baseline <- ggplot(data = min_ttm, 
+       aes(x = weighted_travel_time, 
+           color= income_quintile)) +
+  geom_density(size = 0.5, adjust = 2.5) +
+  labs(x = "Travel time (in minutes; weighted by mode)") +
+  scale_color_manual(values = c("Top 20%" = "blue",
+                                "Second 20%" = "steelblue1",
+                                "Third 20%" = "gray40", 
+                                "Fourth 20%" = "tomato",
+                                "Bottom 20%" = "red")) +
+  theme_minimal() +
+  theme(text = element_text(size = 7),
+        legend.text = element_text(size = 7))
+```
+
+```{r plot-distribution-travel-time-by-region-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Distribution of weighted travel time by region:
+ttm_region_baseline <- ggplot(data = min_ttm %>%
+                                drop_na(Urban_type), 
+       aes(x = weighted_travel_time, 
+           color= Urban_type)) +
+  geom_density(size = 0.5, adjust = 3) +
+  labs(x = "Travel time (in minutes; weighted by mode)") +
+  scale_color_manual(name = "",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen"))  +
+  theme_minimal() +
+  theme(text = element_text(size = 7),
+        legend.text = element_text(size = 7))
+```
+
+```{r total-weighted-person-hours-by-group-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Total weighted person-hours by income:
+person_hours_income_baseline <- min_ttm %>%
+  group_by(income_quintile) %>%
+  summarize(total_weighted_person_hours = sum(weighted_person_hours),
+            total_population = sum(Population),
+            .groups = "drop") %>%
+  mutate(ratio = total_weighted_person_hours/total_population) %>%
+  rename(group = income_quintile)
+
+## Total weighted person-hours by region:
+person_hours_region_baseline <- min_ttm %>%
+  group_by(Urban_type) %>%
+  summarize(total_weighted_person_hours = sum(weighted_person_hours),
+            total_population = sum(Population),
+            .groups = "drop") %>%
+  drop_na() %>%
+  mutate(ratio = total_weighted_person_hours/total_population) %>%
+  rename(group = Urban_type)
+
+person_hours_baseline <- rbind(person_hours_income_baseline,
+                               person_hours_region_baseline)
+```
+
+```{r minimum-travel-time, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Retrieve the minimum travel time from a point to a pharmacy by each mode for people aged 55 to 69:
+min_ttm_car <- ttm_car %>%
+  #dplyr::filter(Type == "Pilot") %>%
+  group_by(UID) %>%
+  summarize(travel_time = min(travel_time),
+            Population = first(Population_55to69),
+            Urban_type = first(Urban_type),
+            DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            p_car = first(p_car),
+            p_transit = first(p_transit),
+            p_walk = first(p_walk),
+            .groups = "drop")
+
+min_ttm_transit <- ttm_transit %>%
+  #dplyr::filter(Type == "Pilot") %>%
+  group_by(UID) %>%
+  summarize(travel_time = min(travel_time),
+            Population = first(Population_55to69),
+            Urban_type = first(Urban_type),
+            DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            p_car = first(p_car),
+            p_transit = first(p_transit),
+            p_walk = first(p_walk),
+            .groups = "drop")
+
+min_ttm_walk <- ttm_walk %>%
+  #dplyr::filter(Type == "Pilot") %>%
+  group_by(UID) %>%
+  summarize(travel_time = min(travel_time),
+            Population = first(Population_55to69),
+            Urban_type = first(Urban_type),
+            DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            p_car = first(p_car),
+            p_transit = first(p_transit),
+            p_walk = first(p_walk),
+            .groups = "drop")
+```
+
+```{r weighted-travel-time, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Calculate weighted travel time:
+min_ttm_car <- min_ttm_car %>%
+  mutate(weighted_travel_time = travel_time * p_car)
+
+min_ttm_transit <- min_ttm_transit %>%
+  mutate(weighted_travel_time = travel_time * p_transit)
+
+min_ttm_walk <- min_ttm_walk %>%
+  mutate(weighted_travel_time = travel_time * p_walk)
+```
+
+```{r bind-travel-time-tables, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Bind the three tables:
+min_ttm <- rbind(min_ttm_car,
+                 min_ttm_transit,
+                 min_ttm_walk)
+```
+
+```{r aggregate-weighted-travel-times, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Aggregate weighted travel times by mode:
+min_ttm <- min_ttm %>%
+  group_by(UID) %>%
+  summarize(DAUID = first(DAUID),
+            TAZUID = first(TAZUID),
+            Population = first(Population),
+            Urban_type = first(Urban_type),
+            weighted_travel_time = sum(weighted_travel_time))
+```
+
+```{r calculate-person-hours-travel, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Calculate person-hours of travel:
+min_ttm <- min_ttm %>%
+  mutate(weighted_person_hours = Population * weighted_travel_time/60)
+```
+
+```{r summarize-results-by-taz, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Summarize by TAZ and join geometry:
+min_ttm_taz <- min_ttm %>%
+  group_by(TAZUID) %>%
+  summarize(weighted_travel_time = mean(weighted_travel_time),
+            weighted_person_hours = sum(weighted_person_hours)) %>%
+  left_join(modes %>%
+              dplyr::select(TAZUID),
+            by = "TAZUID") %>%
+  st_as_sf()
+```
+
+```{r include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Map weighted travel time:
+ggplot() +
+  geom_sf(data = min_ttm_taz,
+          aes(fill = weighted_travel_time),
+          color = NA) +
+  geom_sf(data = hamilton_cma,
+          fill = NA) +
+  scale_fill_distiller(palette = "OrRd", 
+                       direction = 1)
+```
+
+```{r include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Map weighted person-hours of travel:
+ggplot() +
+  geom_sf(data = min_ttm_taz,
+          aes(fill = weighted_person_hours),
+          color = NA) +
+  geom_sf(data = hamilton_cma,
+          fill = NA) +
+  scale_fill_distiller(palette = "OrRd", 
+                       direction = 1)
+```
+
+```{r join-income-information, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Join median household information to table:
+min_ttm <- min_ttm %>%
+  left_join(data_da_2016 %>%
+              st_drop_geometry() %>%
+              dplyr::transmute(DAUID = GeoUID,
+                               income_quintile),
+            by = "DAUID")
+```
+
+```{r plot-distribution-travel-time-by-income, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Distribution of weighted travel time by income:
+ttm_income <- ggplot(data = min_ttm, 
+       aes(x = weighted_travel_time, 
+           color= income_quintile)) +
+  geom_density(size = 0.5, adjust = 2.5) +
+  labs(x = "Travel time (in minutes; weighted by mode)") +
+  scale_color_manual(name = "Income quintile",
+                     values = c("Top 20%" = "blue",
+                                "Second 20%" = "steelblue1",
+                                "Third 20%" = "gray40", 
+                                "Fourth 20%" = "tomato",
+                                "Bottom 20%" = "red")) +
+  theme_minimal() +
+  theme(text = element_text(size = 7),
+        legend.text = element_text(size = 7))
+```
+
+```{r plot-distribution-travel-time-by-region, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Distribution of weighted travel time by region:
+ttm_region <- ggplot(data = min_ttm %>%
+                                drop_na(Urban_type), 
+       aes(x = weighted_travel_time, 
+           color= Urban_type)) +
+  geom_density(size = 0.5, adjust = 3) +
+  labs(x = "Travel time (in minutes; weighted by mode)") +
+  scale_color_manual(name = "Region",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen"))  +
+  theme_minimal() +
+  theme(text = element_text(size = 7),
+        legend.text = element_text(size = 7))
+```
+
+```{r total-weighted-person-hours-by-group, include=FALSE}
+# Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
+## Total weighted person-hours by income:
+person_hours_income <- min_ttm %>%
+  group_by(income_quintile) %>%
+  summarize(total_weighted_person_hours = sum(weighted_person_hours),
+            total_population = sum(Population),
+            .groups = "drop") %>%
+  mutate(ratio = total_weighted_person_hours/total_population) %>%
+  rename(group = income_quintile)
+
+## Total weighted person-hours by region:
+person_hours_region <- min_ttm %>%
+  group_by(Urban_type) %>%
+  summarize(total_weighted_person_hours = sum(weighted_person_hours),
+            total_population = sum(Population),
+            .groups = "drop") %>%
+  drop_na() %>%
+  mutate(ratio = total_weighted_person_hours/total_population) %>%
+  rename(group = Urban_type)
+
+person_hours <- rbind(person_hours_income,
+                      person_hours_region)
+```
+
+Words.
+
+```{r figure-results, echo=FALSE, warning=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:results}Distribution of travel time (weighted by mode) for different population groups"}
+grid.arrange(ttm_income_baseline + 
+               ggtitle("By income: Pilot Program") + 
+               labs(x = "") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8), 
+                     axis.text = element_text(size = 7), 
+                     legend.position = "none"), 
+             ttm_income + 
+               ggtitle("By income: Scenario") + 
+               labs(x = "", y = "") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8), 
+                     axis.text = element_text(size = 7)), 
+             ttm_region_baseline + 
+               ggtitle("By region: Pilot Program") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8),
+                     axis.title = element_text(size = 8), 
+                     axis.text = element_text(size = 7), 
+                     legend.position = "none"), 
+             ttm_region + 
+               ggtitle("By region: Scenario") + 
+               labs(y = "") +
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8),
+                     axis.title = element_text(size = 8), 
+                     axis.text = element_text(size = 7)), 
+             nrow = 2)
+```
+
+```{r table-results, echo=FALSE}
+cbind(person_hours_baseline %>%
+        dplyr::select(group, total_population, total_weighted_person_hours, ratio), 
+      person_hours %>%
+        dplyr::select(total_weighted_person_hours, ratio)) %>%
+  kable("html",
+        booktabs = TRUE,
+        col.names = c("Group", 
+                      "Population", 
+                      "Total PHT", 
+                      "Hours per person", 
+                      "Total PHT",
+                      "Hours per person"),
+        digits = 3,
+        align = c("lccccc"),
+        caption = "\\label{tab:distribution-results}Distribution of person hours travelled (PHT) by median total household income and region: pilot locations only, and scenario with three urban locations added") %>%
+  add_header_above(c(" " = 2, "Pilot Program" = 2, "Scenario" = 2)) %>%
+  kable_styling(latex_options = c("striped", "scale_down")) %>%
+  pack_rows("Income Quintile", 1, 5) %>%
+  pack_rows("Region", 6, 8) %>%
+  footnote(general = "The population totals differ due to small differences in the classification of the regions")
+```
+
+References {#references .unnumbered}
+==========

### 2021-04-05 vaccHamilton/R/vaccHamilton.R

@@ -208,3 +208,21 @@ NULL
 #' @name ttm_walk
 #' @usage data(ttm_walk)
 "ttm_walk"
+
+#' Urban types
+#'
+#' Classification of regions by type: urban, suburban, and rural. This is based on the rural boundaries of the city, and the planning regions.
+#'
+#'#' @format A simple features data frame with 2 rows and 2 variables:
+#' \describe{
+#'   \item{Type}{Classification of the region: "Urban", "Suburban", "Rural"}
+#'   \item{geometry}{Geometry of polygons}
+#' }
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords transportation
+#' @name urban_types
+#' @usage data(urban_types)
+"urban_types"
+

### 2021-04-05 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.Rmd
First draft of paper

@@
  \floatplacement{figure}{H}
+ \usepackage[margin=1.25in]{geometry}
 ---
@@
 library(tidyverse)
-library(vaccHamilton)
+library(units)
+library(vaccHamilton) # Data package
 ```
@@
 
-Research Questions and Hypotheses
-==========================
-
-Along with the provision of health care facilities to treat severe cases of COVID-19 [@Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs.
-The Province of Ontario, in Canada, announced on April 1st 2021 the expansion of a pilot program to offer vaccines in pharmacies in the City of Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
-This program is in addition to dedicated vaccination centers for people aged 70+.
-Twenty pharmacies in Hamilton were added to an earlier list of 325 locations in other cities across the province, and the program was extended to people aged 55 years old and over.
-
-Critics were quick to point out that the list of pharmacies approved for Hamilton were mostly located in lower density parts of the city that are not well serviced by transit and are difficult to reach by foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}.
-Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast majority of the pharmacies are in suburban Hamilton.
-The issue is somewhat less clear cut when we consider that Hamilton's suburbs tend to be older (see Figure \ref{fig:population-map}).
-The population aged 55 to 69 in Hamilton is approximately `r prettyNum(round(population_statistics$Population_55to69[2]), big.mark = ",")` suburban, `r prettyNum(round(population_statistics$Population_55to69[3]), big.mark = ",")` urban, and only `r prettyNum(round(population_statistics$Population_55to69[3]), big.mark = ",")` rural.
-Given the target demographic for the program, it is possible that suburban sites could be convenient for mature and older adults.
-Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the normally inert Major of the city was caught flat footed by the announcement; see:\url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
-As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may introduce inequities in access.
-
-In this research, we investigate the accessibility implications of the sites selected for the pilot vaccination program. Concretely, we ask:
-
-- What is the estimated cost of travel to reach the vaccination sites, assuming that every person requires a vaccine?
-- What is the distribution of this cost across the population of the city?
-- How does the cost and its distribution change with the addition of candidate sites in urban Hamilton?
-
-We concentrate on the 55 to 69 years old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pilot.
-
-```{r pharmacies-map, echo=FALSE, fig.height=5, fig.align="center", fig.cap="\\label{fig:pharmacies-and-regions}Location of pharmacies in pilot and regions with the City of Hamilton"}
-#map_pharmacies <-
-ggplot() +
-  geom_sf(data = urban_types,
-          aes(fill = Type),
-          color = "lightgray") +
-  geom_sf(data = pharmacy_locations %>%
-            filter(Type == "Pilot"),
-          shape = 17,
-          size = 2.5) +
-  scale_color_manual(values = c("Candidate" = "black",
-                                "Pilot" = "black")) +
-  scale_fill_manual(values = c("Urban" = "firebrick2",
-                               "Suburban" = "cornflowerblue",
-                               "Rural" = "forestgreen")) +
-  guides(fill = guide_legend(title = "")) + 
-  theme_minimal() +
-  theme(panel.grid = element_blank(),
-        axis.text = element_text(size = 7),
-        legend.position = "bottom",
-        legend.text = element_text(size = 8))
-
-```
-
-```{r population-map, echo=FALSE, warning=FALSE, fig.height=7, fig.align="center", fig.cap="\\label{fig:population-map}Distribution of population age 55+ in the City of Hamilton"}
-ggplot() +
-  geom_sf(data = data_da_2016 %>%
-            filter(`Region Name` == "Hamilton") %>%
-            dplyr::select(Population_55to69, Population_70plus) %>%
-            pivot_longer(cols = -geometry, names_to = "Age", values_to = "Population") %>%
-            mutate(Age = factor(Age,
-                                levels = c("Population_55to69",
-                                           "Population_70plus"),
-                                labels = c("Age: 55 to 69",
-                                           "Age: 70+"),
-                                ordered = TRUE)) %>%
-            st_as_sf(),
-          aes(fill = Population),
-          color = NA) +
-  geom_sf(data = urban_types,
-          aes(color = Type),
-          fill = NA) +
-  scale_fill_distiller(name = "Population",
-                       breaks = c(20, 50, 150, 400),
-                       palette = "OrRd", 
-                       trans = "log",
-                       direction = 1) +
-  scale_color_manual(name = "",
-                     values = c("Urban" = "firebrick2",
-                                "Suburban" = "cornflowerblue",
-                                "Rural" = "forestgreen"))  +
-  facet_wrap(~ Age, nrow = 2) +
-  theme_minimal() +
-  theme(panel.grid = element_blank(),
-        axis.text = element_text(size = 7),
-        legend.position = "bottom",
-        legend.text = element_text(size = 8))
-```
-
-Methods and Data
-============
-
-This paper is a reproducible research document [see @Brunsdon2020opening] conducted using open source tools for transportation analysis [@Lovelace2021open].
-All code and data necessary to reproduce the tables and figures are available in a public repository\footnote{\url{https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines}}.
-
-Findings
-===================
-
 ```{r minimum-travel-time-baseline, include=FALSE}
@@
 
-```{r include=FALSE}
+```{r weighted-travel-time-map-baseline, include=FALSE}
 # Baseline analysis (PILOT LOCATIONS ONLY)
 ## Map weighted travel time:
-ggplot() +
+ttm_map_baseline <- ggplot() +
   geom_sf(data = min_ttm_taz,
@@
           color = NA) +
-  geom_sf(data = hamilton_cma,
+  scale_fill_distiller(name = "Expected travel time (min)",
+                       palette = "OrRd", 
+                       direction = 1) +
+  geom_sf(data = urban_types,
+          aes(color = Type),
           fill = NA) +
-  scale_fill_distiller(palette = "OrRd", 
-                       direction = 1)
+  scale_color_manual(name = "",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen"))  +
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
 ```
 
-```{r include=FALSE}
+```{r person-hours-traveled-map-baseline, include=FALSE}
 # Baseline analysis (PILOT LOCATIONS ONLY)
 ## Map weighted person-hours of travel:
-ggplot() +
+pht_map_baseline <- ggplot() +
   geom_sf(data = min_ttm_taz,
@@
           color = NA) +
-  geom_sf(data = hamilton_cma,
+  scale_fill_distiller(name = "Person-hours of travel",
+                       palette = "OrRd", 
+                       direction = 1) +
+  geom_sf(data = urban_types,
+          aes(color = Type),
           fill = NA) +
-  scale_fill_distiller(palette = "OrRd", 
-                       direction = 1)
+  scale_color_manual(name = "",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen"))  +
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
 ```
@@
 
-```{r include=FALSE}
+```{r weighted-travel-time-map, include=FALSE}
 # Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
-## Map weighted travel time:
-ggplot() +
+## Map weighted travel time. This is the average of the weighted travel time per TAZ:
+ttm_map <- ggplot() +
   geom_sf(data = min_ttm_taz,
@@
 
-```{r include=FALSE}
+```{r total-person-hours-map, include=FALSE}
 # Scenario analysis (PILOT AND CANDIDATE LOCATIONS)
-## Map weighted person-hours of travel:
-ggplot() +
+## Map weighted person-hours of travel. This is the total person-hours traveled per TAZ:
+pht_map <- ggplot() +
   geom_sf(data = min_ttm_taz,
@@
 
-Words.
+```{r table-comparison, echo=FALSE}
+table_comparison <- cbind(person_hours_baseline %>%
+        dplyr::select(group, total_population, total_weighted_person_hours, ratio), 
+      person_hours %>%
+        dplyr::select(total_weighted_person_hours, ratio))
+```
 
-```{r figure-results, echo=FALSE, warning=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:results}Distribution of travel time (weighted by mode) for different population groups"}
+Research Questions and Hypotheses
+==========================
+
+Along with the provision of health care facilities to treat severe cases of COVID-19 [@Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs.
+The Province of Ontario, in Canada, announced on April 1st 2021 the expansion of a pilot program to offer vaccines in pharmacies in the City of Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
+This program is in addition to dedicated vaccination centers for people aged 70+.
+Twenty pharmacies in Hamilton were added to an earlier list of 325 locations in other cities across the province, and the program was extended to people aged 55 years old and over.
+
+Critics were swift to point out that the list of pharmacies approved for Hamilton by the province were mostly located in lower density parts of the city that are not well serviced by transit and are difficult to reach by foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}.
+Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast majority of the pharmacies are in suburban Hamilton.
+The issue is somewhat less clear-cut when we consider that Hamilton's population skews suburban (see Figure \ref{fig:population-map}).
+Given the target demographic for the program, it is possible that suburban sites could be convenient for mature adults and young old: the population aged 55 to 69 in Hamilton is approximately `r prettyNum(round((table_comparison[7,2]/1000), 2) * 1000, big.mark = ",")` suburban, `r prettyNum(round((table_comparison[8,2]/1000), 2) * 1000, big.mark = ",")` urban, and only `r prettyNum(round((table_comparison[6,2]/1000), 2) * 1000, big.mark = ",")` rural.
+Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the typically inert Major of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
+As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may introduce inequities in access.
+
+In this research, we investigate the accessibility implications of the sites selected for the pilot vaccination program. Concretely, we ask:
+
+- What is the estimated cost of travel to reach the vaccination sites, assuming that every person requires a vaccine?
+- What is the distribution of this cost across the population of the city?
+- How does the cost and its distribution change with the addition of candidate sites in urban Hamilton?
+
+We concentrate on the 55 to 69 years old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pilot.
+
+```{r pharmacies-map, echo=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:pharmacies-and-regions}Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) and urban locations for scenario analysis (white circles)"}
+#map_pharmacies <-
+ggplot() +
+  geom_sf(data = urban_types,
+          aes(fill = Type),
+          color = "lightgray") +
+  geom_sf(data = pharmacy_locations,
+          aes(shape = Type,
+              color = Type),
+          size = 2.5,
+          show.legend = FALSE) +
+  scale_color_manual(values = c("Candidate" = "white",
+                                "Pilot" = "black")) +
+  scale_fill_manual(values = c("Urban" = "firebrick2",
+                               "Suburban" = "cornflowerblue",
+                               "Rural" = "forestgreen")) +
+  guides(fill = guide_legend(title = "")) + 
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+```
+
+```{r population-map, echo=FALSE, warning=FALSE, fig.height=8, fig.align="center", fig.cap="\\label{fig:population-map}Distribution of population age 55+ in the City of Hamilton"}
+ggplot() +
+  geom_sf(data = data_da_2016 %>%
+            filter(`Region Name` == "Hamilton") %>%
+            dplyr::select(Population_55to69, Population_70plus) %>%
+            pivot_longer(cols = -geometry, names_to = "Age", values_to = "Population") %>%
+            mutate(Age = factor(Age,
+                                levels = c("Population_55to69",
+                                           "Population_70plus"),
+                                labels = c("Age: 55 to 69",
+                                           "Age: 70+"),
+                                ordered = TRUE)) %>%
+            st_as_sf(),
+          aes(fill = Population),
+          color = NA) +
+  geom_sf(data = urban_types,
+          aes(color = Type),
+          fill = NA) +
+  scale_fill_distiller(name = "Population",
+                       breaks = c(20, 50, 150, 400),
+                       palette = "OrRd", 
+                       trans = "log",
+                       direction = 1) +
+  scale_color_manual(name = "",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen"))  +
+  facet_wrap(~ Age, nrow = 2) +
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+```
+
+Methods and Data
+============
+
+# Data
+
+We use data from the following sources.
+
+### Open Hamilton
+
+From the open data portal of the City of Hamilton\footnote{\url{https://open.hamilton.ca/}} we obtained boundaries for the city's various regions (the definition of urban, suburban, and rural regions follows the classification of development applications).
+
+### Statistics Canada
+
+Population and income statistics at the level of Dissemination Areas (DAs) were retrieved using the package `cancensus` [@vonBergmann2021cancensus].
+Dissemination Areas are the smallest publicly available census geography in Canada.
+We use data from the 2016 Population Census.
+
+### Transportation Tomorrow Survey
+
+Data about modal split by age by place of residence were downloaded from the Data Retrieval System of the Transportation Tomorrow Survey (TTS)\footnote{\url{http://dmg.utoronto.ca/}}.
+The data are geocoded at the level of Traffic Analysis Zones (TAZ).
+
+### Other
+
+The locations of pharmacies in the pilot were obtained from public records and geocoded.
+Three urban sites not in the program were also identified and geocoded for comparison purposes.
+In addition, we converted all recorded residential parcels in the City of Hamilton to points on the road network.
+Each point includes information about the number of residential units in the parcel.
+
+# Methods
+
+We used the population aged 55 to 59 y.o.
+in each Dissemination Area to calculate the average number of people per dwelling.
+This value was then assigned proportionally to the number of dwellings per parcel.
+The median total household income of the corresponding DAs was joined to the parcels.
+In addition, we calculated the proportion of trips by mode from the total number of trips by each of the three modes retrieved from the TTS data.
+These proportions were joined to the parcel data based on their corresponding TAZ.
+The package `r5r` [@Pereira2021r5r] was used to calculate the travel time from each parcel to all pharmacies by three modes: car, transit, and walking.
+For routing purposes we used a cutoff value of 180 min and a maximum walking distance of 10,000 m.
+
+Once we obtained travel time tables with population, proportion of trips by mode, and income information, we calculated the expected travel time $ett$ from each parcel $i$ to a pharmacy $j$ as follows:
+$$
+ett_i = p^c_i \min(tt^c_{ij}) + p^t_i \min(tt^t_{ij}) + p^w_i \min(tt^w_{ij})
+$$
+\noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the vector of travel times from parcel $i$ to the pharmacies.
+In other words, the expected travel time is the weighted sum of travel times to the nearest pharmacy, with the weights given by the expected modal split in the TAZ.
+
+The expected travel time $i$ was multiplied by the population in parcel $i$ to obtain a measure of person-hours of travel ($PHT$) as follows:
+$$
+PHT_i = P_i\cdot ett_i
+$$
+
+Please note that this paper is a reproducible research document [see @Brunsdon2020opening] conducted using open source tools for transportation analysis [@Lovelace2021open].
+The code and data necessary to reproduce the analysis are available in a public repository\footnote{\url{https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines}}.
+
+Findings
+===================
+
+The top panel of Figure \ref{fig:maps-baseline} shows the average expected travel time by TAZ in Hamilton.
+It is apparent that travel times tend to be lower in much of suburban Hamilton, and higher in the urban core and some rural parts of the city, particularly to the west.
+This is unsurprising, given the higher probability of travel by car and the predominantly suburban character of the vaccination sites.
+However, even accounting for the distribution of population, this leads to large disparities in the number of person-hours of travel across the city, with a concentration of the burden of travel in the urban core and the rural west (see bottom panel of Figure \ref{fig:maps-baseline}).
+
+The disparities are not trivial.
+
+As seen in Table \ref{tab:distribution-results}, under the pilot program approximately `r round((table_comparison[5,2] + table_comparison[4,2]) / (sum(table_comparison[1:5,2])) * 100, 2)`\% of people live in DAs in the bottom 40% of the median household income scale, but they account for `r round((table_comparison[5,3] + table_comparison[4,3]) / (sum(table_comparison[1:5,3])) * 100, 2)`\% of the total person-hours of travel.
+In contrast, `r round((table_comparison[1,2] + table_comparison[2,2]) / (sum(table_comparison[1:5,2])) * 100, 2)`\% of people aged 55 to 69 in DAs in the top 40% of the median household income scale accrue only `r round((table_comparison[1,3] + table_comparison[2,3]) / (sum(table_comparison[1:5,3])) * 100, 2)`\% of the total person-hours of travel.
+Where the mean travel time of residents of DAs with high median household income is `r table_comparison[1, 4] %>% set_units(h) %>% set_units(min) %>% round() %>% drop_units()` minutes, residents of lower income DAs average `r table_comparison[5, 4] %>% set_units(h) %>% set_units(min) %>% round() %>% drop_units()` minutes in travel time.
+In addition to longer average travel time, residents in lower income DAs also see substantially larger variations in travel times, and some may face considerably longer travel times (see top-left panel in Figure \ref{fig:results}).
+
+There are also important disparities by region.
+As shown in Table \ref{tab:distribution-results}, the urban and rural populations in Hamilton are approximately `r round((table_comparison[6,2] + table_comparison[8,2])/ (sum(table_comparison[6:8,2])) * 100, 2)`\% of the population but they bear `r round((table_comparison[6,3] + table_comparison[8,3])/ (sum(table_comparison[6:8,3])) * 100, 2)`\% of the total person-hours of travel, with also much greater variability in expected travel times (Figure \ref{fig:results}, bottom-left panel).
+
+For comparison purposes we consider a scenario with some modest additions to the list of pharmacies in the provincial pilot.
+We repeat the analysis, but include the three urban sites shown in white circles in Figure \ref{fig:pharmacies-and-regions}.
+The results of this scenario appear in the last two columns of Table \ref{tab:distribution-results} and the two right panels of Figure \ref{fig:results}.
+We begin by noting that all income groups benefit from the addition of these three sites with shorter mean trip durations; the most remarkable difference is the large reduction in the disparities between residents in DAs with different levels of income.
+The top-right panel of Figure \ref{fig:results} shows that the distribution of expected travel time is now more in line for all income groups, even if the bottom two income quintiles still have somewhat wider spreads.
+With respect to the distribution of expected travel time and person-hours of travel by region, unsurprisingly the addition of three urban vaccination sites in the scenario makes a difference for urban residents but not for rural residents.
+
+The results indicate that the locations chosen by the province for the pilot vaccination program do not serve well urban or rural residents of the city, and there are some important questions regarding equity of access to the program, with a disproportionate burden in the cost of travel falling on lower income urban populations and rural populations.
+A scenario did not consider candidate sites in a systematic way.
+Nonetheless, selection of three sensible urban locations does much to alleviate disparities in the burden of transportation.
+On the other hand, unlike the urban context where there are numerous candidate locations that could be chosen for a vaccination program, there are not many candidate locations in rural regions of the city.
+Increasing access in rural Hamilton likely will involve an expansion of existing mobile vaccination pop-up clinics\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
+
+```{r figure-maps-baseline, echo=FALSE, warning=FALSE, fig.height=7, fig.align="center", fig.cap="\\label{fig:maps-baseline}Average expected travel time by TAZ (in minutes) and total person-hours of travel by TAZ."}
+grid.arrange(ttm_map_baseline +
+               theme(legend.title = element_text(size = 8),
+                     legend.text = element_text(size = 7), 
+                     axis.text = element_text(size = 7)),
+             pht_map_baseline +
+               theme(legend.title = element_text(size = 8),
+                     legend.text = element_text(size = 7), 
+                     axis.text = element_text(size = 7)),
+             nrow = 2)
+```
+
+```{r table-results, echo=FALSE}
+table_comparison  %>%
+  kable("latex",
+        booktabs = TRUE,
+        col.names = c("Group", 
+                      "Population", 
+                      "Total PHT", 
+                      "Hours per person", 
+                      "Total PHT",
+                      "Hours per person"),
+        digits = 3,
+        align = c("lccccc"),
+        caption = "\\label{tab:distribution-results}Distribution of person-hours of travel (PHT) by median total household income and region: pilot locations only, and scenario with three urban locations added") %>%
+  add_header_above(c(" " = 2, "Pilot Program" = 2, "Scenario" = 2)) %>%
+  kable_styling(latex_options = c("striped", "scale_down")) %>%
+  pack_rows("Income Quintile", 1, 5) %>%
+  pack_rows("Region", 6, 8) %>%
+  footnote(general = "The population totals differ due to small differences in the classification of the regions")
+```
+
+```{r figure-results, echo=FALSE, warning=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:results}Distribution of expected travel time for different population groups"}
 grid.arrange(ttm_income_baseline + 
@@
 
-```{r table-results, echo=FALSE}
-cbind(person_hours_baseline %>%
-        dplyr::select(group, total_population, total_weighted_person_hours, ratio), 
-      person_hours %>%
-        dplyr::select(total_weighted_person_hours, ratio)) %>%
-  kable("latex",
-        booktabs = TRUE,
-        col.names = c("Group", 
-                      "Population", 
-                      "Total PHT", 
-                      "Hours per person", 
-                      "Total PHT",
-                      "Hours per person"),
-        digits = 3,
-        align = c("lccccc"),
-        caption = "\\label{tab:distribution-results}Distribution of person hours travelled (PHT) by median total household income and region: pilot locations only, and scenario with three urban locations added") %>%
-  add_header_above(c(" " = 2, "Pilot Program" = 2, "Scenario" = 2)) %>%
-  kable_styling(latex_options = c("striped", "scale_down")) %>%
-  pack_rows("Income Quintile", 1, 5) %>%
-  pack_rows("Region", 6, 8) %>%
-  footnote(general = "The population totals differ due to small differences in the classification of the regions")
-```
-
 References {#references .unnumbered}

### 2021-04-05 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.tex
First draft of paper

@@
 % Pandoc header
-\usepackage{float} \usepackage{xcolor} \floatplacement{figure}{H}
+\usepackage{float} \usepackage{xcolor} \floatplacement{figure}{H} \usepackage[margin=1.25in]{geometry}
 \usepackage{booktabs}
@@
 Along with the provision of health care facilities to treat severe cases
-of COVID-19 (Pereira et al., 2021), another front in the fight against
-the pandemic is the rolling out of vaccination programs. The Province of
-Ontario, in Canada, announced on April 1st 2021 the expansion of a pilot
-program to offer vaccines in pharmacies in the City of
+of COVID-19 (Pereira, C. K. V. S. Braga, et al., 2021), another front in
+the fight against the pandemic is the rolling out of vaccination
+programs. The Province of Ontario, in Canada, announced on April 1st
+2021 the expansion of a pilot program to offer vaccines in pharmacies in
+the City of
 Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
@@
 
-Critics were quick to point out that the list of pharmacies approved for
-Hamilton were mostly located in lower density parts of the city that are
-not well serviced by transit and are difficult to reach by
+Critics were swift to point out that the list of pharmacies approved for
+Hamilton by the province were mostly located in lower density parts of
+the city that are not well serviced by transit and are difficult to
+reach by
 foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}.
@@
 majority of the pharmacies are in suburban Hamilton. The issue is
-somewhat less clear cut when we consider that Hamilton's suburbs tend to
-be older (see Figure \ref{fig:population-map}). The population aged 55
-to 69 in Hamilton is approximately 59,095 suburban, 35,704 urban, and
-only 35,704 rural. Given the target demographic for the program, it is
-possible that suburban sites could be convenient for mature and older
-adults. Nevertheless, the selection of sites by the province raises some
-important
-questions\footnote{The decision-making process to select these sites appears to have been opaque, and the normally inert Major of the city was caught flat footed by the announcement; see:\url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
+somewhat less clear-cut when we consider that Hamilton's population
+skews suburban (see Figure \ref{fig:population-map}). Given the target
+demographic for the program, it is possible that suburban sites could be
+convenient for mature adults and young old: the population aged 55 to 69
+in Hamilton is approximately 58,710 suburban, 35,490 urban, and only
+8,360 rural. Nevertheless, the selection of sites by the province raises
+some important
+questions\footnote{The decision-making process to select these sites appears to have been opaque, and the typically inert Major of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
 As Yu et al. (2021) note, good geographical coverage is a key element
@@
 
-\caption{\label{fig:pharmacies-and-regions}Location of pharmacies in pilot and regions with the City of Hamilton}\label{fig:pharmacies-map}
+\caption{\label{fig:pharmacies-and-regions}Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) and urban locations for scenario analysis (white circles)}\label{fig:pharmacies-map}
 \end{figure}
@@
 
-This paper is a reproducible research document (see Brunsdon and Comber,
-2020) conducted using open source tools for transportation analysis
-(Lovelace, 2021). All code and data necessary to reproduce the tables
-and figures are available in a public
+\hypertarget{data}{%
+\section{Data}\label{data}}
+
+We use data from the following sources.
+
+\hypertarget{open-hamilton}{%
+\subsubsection{Open Hamilton}\label{open-hamilton}}
+
+From the open data portal of the City of
+Hamilton\footnote{\url{https://open.hamilton.ca/}} we obtained
+boundaries for the city's various regions (the definition of urban,
+suburban, and rural regions follows the classification of development
+applications).
+
+\hypertarget{statistics-canada}{%
+\subsubsection{Statistics Canada}\label{statistics-canada}}
+
+Population and income statistics at the level of Dissemination Areas
+(DAs) were retrieved using the package \texttt{cancensus} (von Bergmann
+et al., 2021). Dissemination Areas are the smallest publicly available
+census geography in Canada. We use data from the 2016 Population Census.
+
+\hypertarget{transportation-tomorrow-survey}{%
+\subsubsection{Transportation Tomorrow
+Survey}\label{transportation-tomorrow-survey}}
+
+Data about modal split by age by place of residence were downloaded from
+the Data Retrieval System of the Transportation Tomorrow Survey
+(TTS)\footnote{\url{http://dmg.utoronto.ca/}}. The data are geocoded at
+the level of Traffic Analysis Zones (TAZ).
+
+\hypertarget{other}{%
+\subsubsection{Other}\label{other}}
+
+The locations of pharmacies in the pilot were obtained from public
+records and geocoded. Three urban sites not in the program were also
+identified and geocoded for comparison purposes. In addition, we
+converted all recorded residential parcels in the City of Hamilton to
+points on the road network. Each point includes information about the
+number of residential units in the parcel.
+
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
+
+We used the population aged 55 to 59 y.o. in each Dissemination Area to
+calculate the average number of people per dwelling. This value was then
+assigned proportionally to the number of dwellings per parcel. The
+median total household income of the corresponding DAs was joined to the
+parcels. In addition, we calculated the proportion of trips by mode from
+the total number of trips by each of the three modes retrieved from the
+TTS data. These proportions were joined to the parcel data based on
+their corresponding TAZ. The package \texttt{r5r} (Pereira, Saraiva, et
+al., 2021) was used to calculate the travel time from each parcel to all
+pharmacies by three modes: car, transit, and walking. For routing
+purposes we used a cutoff value of 180 min and a maximum walking
+distance of 10,000 m.
+
+Once we obtained travel time tables with population, proportion of trips
+by mode, and income information, we calculated the expected travel time
+\(ett\) from each parcel \(i\) to a pharmacy \(j\) as follows: \[
+ett_i = p^c_i \min(tt^c_{ij}) + p^t_i \min(tt^t_{ij}) + p^w_i \min(tt^w_{ij})
+\] \noindent where \(p^k_i\) is the proportion of trips by mode \(k\) in
+the TAZ of parcel \(i\), and \(tt^k_{ij}\) is the vector of travel times
+from parcel \(i\) to the pharmacies. In other words, the expected travel
+time is the weighted sum of travel times to the nearest pharmacy, with
+the weights given by the expected modal split in the TAZ.
+
+The expected travel time \(i\) was multiplied by the population in
+parcel \(i\) to obtain a measure of person-hours of travel (\(PHT\)) as
+follows: \[
+PHT_i = P_i\cdot ett_i
+\]
+
+Please note that this paper is a reproducible research document (see
+Brunsdon and Comber, 2020) conducted using open source tools for
+transportation analysis (Lovelace, 2021). The code and data necessary to
+reproduce the analysis are available in a public
 repository\footnote{\url{https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines}}.
@@
 
-Words.
+The top panel of Figure \ref{fig:maps-baseline} shows the average
+expected travel time by TAZ in Hamilton. It is apparent that travel
+times tend to be lower in much of suburban Hamilton, and higher in the
+urban core and some rural parts of the city, particularly to the west.
+This is unsurprising, given the higher probability of travel by car and
+the predominantly suburban character of the vaccination sites. However,
+even accounting for the distribution of population, this leads to large
+disparities in the number of person-hours of travel across the city,
+with a concentration of the burden of travel in the urban core and the
+rural west (see bottom panel of Figure \ref{fig:maps-baseline}).
+
+The disparities are not trivial.
+
+As seen in Table \ref{tab:distribution-results}, under the pilot program
+approximately 36.42\% of people live in DAs in the bottom 40\% of the
+median household income scale, but they account for 51.98\% of the total
+person-hours of travel. In contrast, 44.5\% of people aged 55 to 69 in
+DAs in the top 40\% of the median household income scale accrue only
+35.03\% of the total person-hours of travel. Where the mean travel time
+of residents of DAs with high median household income is 6 minutes,
+residents of lower income DAs average 12 minutes in travel time. In
+addition to longer average travel time, residents in lower income DAs
+also see substantially larger variations in travel times, and some may
+face considerably longer travel times (see top-left panel in Figure
+\ref{fig:results}).
+
+There are also important disparities by region. As shown in Table
+\ref{tab:distribution-results}, the urban and rural populations in
+Hamilton are approximately 42.75\% of the population but they bear
+69.25\% of the total person-hours of travel, with also much greater
+variability in expected travel times (Figure \ref{fig:results},
+bottom-left panel).
+
+For comparison purposes we consider a scenario with some modest
+additions to the list of pharmacies in the provincial pilot. We repeat
+the analysis, but include the three urban sites shown in white circles
+in Figure \ref{fig:pharmacies-and-regions}. The results of this scenario
+appear in the last two columns of Table \ref{tab:distribution-results}
+and the two right panels of Figure \ref{fig:results}. We begin by noting
+that all income groups benefit from the addition of these three sites
+with shorter mean trip durations; the most remarkable difference is the
+large reduction in the disparities between residents in DAs with
+different levels of income. The top-right panel of Figure
+\ref{fig:results} shows that the distribution of expected travel time is
+now more in line for all income groups, even if the bottom two income
+quintiles still have somewhat wider spreads. With respect to the
+distribution of expected travel time and person-hours of travel by
+region, unsurprisingly the addition of three urban vaccination sites in
+the scenario makes a difference for urban residents but not for rural
+residents.
+
+The results indicate that the locations chosen by the province for the
+pilot vaccination program do not serve well urban or rural residents of
+the city, and there are some important questions regarding equity of
+access to the program, with a disproportionate burden in the cost of
+travel falling on lower income urban populations and rural populations.
+A scenario did not consider candidate sites in a systematic way.
+Nonetheless, selection of three sensible urban locations does much to
+alleviate disparities in the burden of transportation. On the other
+hand, unlike the urban context where there are numerous candidate
+locations that could be chosen for a vaccination program, there are not
+many candidate locations in rural regions of the city. Increasing access
+in rural Hamilton likely will involve an expansion of existing mobile
+vaccination pop-up
+clinics\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
 
@@
 
-{\centering \includegraphics{Accessibility-Vaccination-Sites-Hamilton_files/figure-latex/figure-results-1}
+{\centering \includegraphics{Accessibility-Vaccination-Sites-Hamilton_files/figure-latex/figure-maps-baseline-1}
 
@@
 
-\caption{\label{fig:results}Distribution of travel time (weighted by mode) for different population groups}\label{fig:figure-results}
+\caption{\label{fig:maps-baseline}Average expected travel time by TAZ (in minutes) and total person-hours of travel by TAZ.}\label{fig:figure-maps-baseline}
 \end{figure}
@@
 
-\caption{\label{tab:table-results}\label{tab:distribution-results}Distribution of person hours travelled (PHT) by median total household income and region: pilot locations only, and scenario with three urban locations added}
+\caption{\label{tab:table-results}\label{tab:distribution-results}Distribution of person-hours of travel (PHT) by median total household income and region: pilot locations only, and scenario with three urban locations added}
 \centering
@@
 
+\begin{figure}
+
+{\centering \includegraphics{Accessibility-Vaccination-Sites-Hamilton_files/figure-latex/figure-results-1}
+
+}
+
+\caption{\label{fig:results}Distribution of expected travel time for different population groups}\label{fig:figure-results}
+\end{figure}
+
 \hypertarget{references}{%
@@
 Geographical Systems 1--20.
+doi:\href{https://doi.org/10.1007/s10109-020-00334-2}{10.1007/s10109-020-00334-2}
 
@@
 \leavevmode\hypertarget{ref-Pereira2021geographic}{}%
-Pereira, R.H.M., Braga, C.K.V., Servo, L.M., Serra, B., Amaral, P.,
+Pereira, R.H.M., Braga, C.K.V.S., Mendes and Serra, L., Amaral, P.B.,
 Gouveia, N., Paez, A., 2021. Geographic access to covid-19 healthcare in
@@
 doi:\href{https://doi.org/https://doi.org/10.1016/j.socscimed.2021.113773}{https://doi.org/10.1016/j.socscimed.2021.113773}
+
+\leavevmode\hypertarget{ref-Pereira2021r5r}{}%
+Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga, C.K.V., Conway,
+M.W., 2021. R5r: Rapid realistic routing on multimodal transport
+networks with r\textsuperscript{5} in r. Findings.
+doi:\href{https://doi.org/10.32866/001c.21262}{10.32866/001c.21262}
+
+\leavevmode\hypertarget{ref-vonBergmann2021cancensus}{}%
+von Bergmann, J., Shkolnik, D., Jacobs, A., 2021. Cancensus: R package
+to access, retrieve, and work with canadian census data and geography.
 

### 2021-04-05 README.Rmd
First draft of paper

@@
 
-## Research Questions and Hypotheses
-
-Along with the provision of health care facilities to treat severe cases of COVID-19 [@Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs.
-The Province of Ontario, in Canada, announced on April 1st 2021 the expansion of a pilot program to offer vaccines in pharmacies in the City of Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
-This program is in addition to dedicated vaccination centers for people aged 70+.
-Twenty pharmacies in Hamilton were added to an earlier list of 325 locations in other cities across the province, and the program was extended to people aged 55 years old and over.
-
-Critics were quick to point out that the list of pharmacies approved for Hamilton were mostly located in lower density parts of the city that are not well serviced by transit and are difficult to reach by foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}.
-Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast majority of the pharmacies are in suburban Hamilton.
-The issue is somewhat less clear cut when we consider that Hamilton's suburbs tend to be older (see Figure \ref{fig:population-map}).
-The population aged 55 to 69 in Hamilton is approximately `r prettyNum(round(population_statistics$Population_55to69[2]), big.mark = ",")` suburban, `r prettyNum(round(population_statistics$Population_55to69[3]), big.mark = ",")` urban, and only `r prettyNum(round(population_statistics$Population_55to69[3]), big.mark = ",")` rural.
-Given the target demographic for the program, it is possible that suburban sites could be convenient for mature and older adults.
-Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the normally inert Major of the city was caught flat footed by the announcement; see:\url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
-As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may introduce inequities in access.
-
-In this research, we investigate the accessibility implications of the sites selected for the pilot vaccination program. Concretely, we ask:
-
-- What is the estimated cost of travel to reach the vaccination sites, assuming that every person requires a vaccine?
-- What is the distribution of this cost across the population of the city?
-- How does the cost and its distribution change with the addition of candidate sites in urban Hamilton?
-
-We concentrate on the 55 to 69 years old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pilot.
-
-```{r pharmacies-map, echo=FALSE, fig.height=5, fig.align="center", fig.cap="\\label{fig:pharmacies-and-regions}Location of pharmacies in pilot and regions with the City of Hamilton"}
-#map_pharmacies <-
-ggplot() +
-  geom_sf(data = urban_types,
-          aes(fill = Type),
-          color = "lightgray") +
-  geom_sf(data = pharmacy_locations %>%
-            filter(Type == "Pilot"),
-          shape = 17,
-          size = 2.5) +
-  scale_color_manual(values = c("Candidate" = "black",
-                                "Pilot" = "black")) +
-  scale_fill_manual(values = c("Urban" = "firebrick2",
-                               "Suburban" = "cornflowerblue",
-                               "Rural" = "forestgreen")) +
-  guides(fill = guide_legend(title = "")) + 
-  theme_minimal() +
-  theme(panel.grid = element_blank(),
-        axis.text = element_text(size = 7),
-        legend.position = "bottom",
-        legend.text = element_text(size = 8))
-
-```
-
-```{r population-map, echo=FALSE, warning=FALSE, fig.height=7, fig.align="center", fig.cap="\\label{fig:population-map}Distribution of population age 55+ in the City of Hamilton"}
-ggplot() +
-  geom_sf(data = data_da_2016 %>%
-            filter(`Region Name` == "Hamilton") %>%
-            dplyr::select(Population_55to69, Population_70plus) %>%
-            pivot_longer(cols = -geometry, names_to = "Age", values_to = "Population") %>%
-            mutate(Age = factor(Age,
-                                levels = c("Population_55to69",
-                                           "Population_70plus"),
-                                labels = c("Age: 55 to 69",
-                                           "Age: 70+"),
-                                ordered = TRUE)) %>%
-            st_as_sf(),
-          aes(fill = Population),
-          color = NA) +
-  geom_sf(data = urban_types,
-          aes(color = Type),
-          fill = NA) +
-  scale_fill_distiller(name = "Population",
-                       breaks = c(20, 50, 150, 400),
-                       palette = "OrRd", 
-                       trans = "log",
-                       direction = 1) +
-  scale_color_manual(name = "",
-                     values = c("Urban" = "firebrick2",
-                                "Suburban" = "cornflowerblue",
-                                "Rural" = "forestgreen"))  +
-  facet_wrap(~ Age, nrow = 2) +
-  theme_minimal() +
-  theme(panel.grid = element_blank(),
-        axis.text = element_text(size = 7),
-        legend.position = "bottom",
-        legend.text = element_text(size = 8))
-```
-
-## Methods and Data
-
-This paper is a reproducible research document [see @Brunsdon2020opening] conducted using open source tools for transportation analysis [@Lovelace2021open].
-All code and data necessary to reproduce the tables and figures are available in a public repository\footnote{\url{https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines}}.
-
-## Findings
-
 ```{r minimum-travel-time-baseline, include=FALSE}
@@
 
-Words.
+```{r table-comparison, echo=FALSE}
+table_comparison <- cbind(person_hours_baseline %>%
+        dplyr::select(group, total_population, total_weighted_person_hours, ratio), 
+      person_hours %>%
+        dplyr::select(total_weighted_person_hours, ratio))
+```
 
-```{r figure-results, echo=FALSE, warning=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:results}Distribution of travel time (weighted by mode) for different population groups"}
+## Research Questions and Hypotheses
+
+Along with the provision of health care facilities to treat severe cases of COVID-19 [@Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs.
+The Province of Ontario, in Canada, announced on April 1st 2021 the expansion of a pilot program to offer vaccines in pharmacies in the City of Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
+This program is in addition to dedicated vaccination centers for people aged 70+.
+Twenty pharmacies in Hamilton were added to an earlier list of 325 locations in other cities across the province, and the program was extended to people aged 55 years old and over.
+
+Critics were swift to point out that the list of pharmacies approved for Hamilton by the province were mostly located in lower density parts of the city that are not well serviced by transit and are difficult to reach by foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}.
+Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast majority of the pharmacies are in suburban Hamilton.
+The issue is somewhat less clear-cut when we consider that Hamilton's population skews suburban (see Figure \ref{fig:population-map}).
+Given the target demographic for the program, it is possible that suburban sites could be convenient for mature adults and young old: the population aged 55 to 69 in Hamilton is approximately `r prettyNum(round((table_comparison[7,2]/1000), 2) * 1000, big.mark = ",")` suburban, `r prettyNum(round((table_comparison[8,2]/1000), 2) * 1000, big.mark = ",")` urban, and only `r prettyNum(round((table_comparison[6,2]/1000), 2) * 1000, big.mark = ",")` rural.
+Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the typically inert Major of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
+As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may introduce inequities in access.
+
+In this research, we investigate the accessibility implications of the sites selected for the pilot vaccination program. Concretely, we ask:
+
+- What is the estimated cost of travel to reach the vaccination sites, assuming that every person requires a vaccine?
+- What is the distribution of this cost across the population of the city?
+- How does the cost and its distribution change with the addition of candidate sites in urban Hamilton?
+
+We concentrate on the 55 to 69 years old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pilot.
+
+```{r pharmacies-map, echo=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:pharmacies-and-regions}Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) and urban locations for scenario analysis (white circles)"}
+#map_pharmacies <-
+ggplot() +
+  geom_sf(data = urban_types,
+          aes(fill = Type),
+          color = "lightgray") +
+  geom_sf(data = pharmacy_locations,
+          aes(shape = Type,
+              color = Type),
+          size = 2.5,
+          show.legend = FALSE) +
+  scale_color_manual(values = c("Candidate" = "white",
+                                "Pilot" = "black")) +
+  scale_fill_manual(values = c("Urban" = "firebrick2",
+                               "Suburban" = "cornflowerblue",
+                               "Rural" = "forestgreen")) +
+  guides(fill = guide_legend(title = "")) + 
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+```
+
+```{r population-map, echo=FALSE, warning=FALSE, fig.height=8, fig.align="center", fig.cap="\\label{fig:population-map}Distribution of population age 55+ in the City of Hamilton"}
+ggplot() +
+  geom_sf(data = data_da_2016 %>%
+            filter(`Region Name` == "Hamilton") %>%
+            dplyr::select(Population_55to69, Population_70plus) %>%
+            pivot_longer(cols = -geometry, names_to = "Age", values_to = "Population") %>%
+            mutate(Age = factor(Age,
+                                levels = c("Population_55to69",
+                                           "Population_70plus"),
+                                labels = c("Age: 55 to 69",
+                                           "Age: 70+"),
+                                ordered = TRUE)) %>%
+            st_as_sf(),
+          aes(fill = Population),
+          color = NA) +
+  geom_sf(data = urban_types,
+          aes(color = Type),
+          fill = NA) +
+  scale_fill_distiller(name = "Population",
+                       breaks = c(20, 50, 150, 400),
+                       palette = "OrRd", 
+                       trans = "log",
+                       direction = 1) +
+  scale_color_manual(name = "",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen"))  +
+  facet_wrap(~ Age, nrow = 2) +
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+```
+
+## Methods and Data
+
+## Data
+
+We use data from the following sources.
+
+### Open Hamilton
+
+From the open data portal of the City of Hamilton\footnote{\url{https://open.hamilton.ca/}} we obtained boundaries for the city's various regions (the definition of urban, suburban, and rural regions follows the classification of development applications).
+
+### Statistics Canada
+
+Population and income statistics at the level of Dissemination Areas (DAs) were retrieved using the package `cancensus` [@vonBergmann2021cancensus].
+Dissemination Areas are the smallest publicly available census geography in Canada.
+We use data from the 2016 Population Census.
+
+### Transportation Tomorrow Survey
+
+Data about modal split by age by place of residence were downloaded from the Data Retrieval System of the Transportation Tomorrow Survey (TTS)\footnote{\url{http://dmg.utoronto.ca/}}.
+The data are geocoded at the level of Traffic Analysis Zones (TAZ).
+
+### Other
+
+The locations of pharmacies in the pilot were obtained from public records and geocoded.
+Three urban sites not in the program were also identified and geocoded for comparison purposes.
+In addition, we converted all recorded residential parcels in the City of Hamilton to points on the road network.
+Each point includes information about the number of residential units in the parcel.
+
+## Methods
+
+We used the population aged 55 to 59 y.o.
+in each Dissemination Area to calculate the average number of people per dwelling.
+This value was then assigned proportionally to the number of dwellings per parcel.
+The median total household income of the corresponding DAs was joined to the parcels.
+In addition, we calculated the proportion of trips by mode from the total number of trips by each of the three modes retrieved from the TTS data.
+These proportions were joined to the parcel data based on their corresponding TAZ.
+The package `r5r` [@Pereira2021r5r] was used to calculate the travel time from each parcel to all pharmacies by three modes: car, transit, and walking.
+For routing purposes we used a cutoff value of 180 min and a maximum walking distance of 10,000 m.
+
+Once we obtained travel time tables with population, proportion of trips by mode, and income information, we calculated the expected travel time $ett$ from each parcel $i$ to a pharmacy $j$ as follows:
+$$
+ett_i = p^c_i \min(tt^c_{ij}) + p^t_i \min(tt^t_{ij}) + p^w_i \min(tt^w_{ij})
+$$
+\noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the vector of travel times from parcel $i$ to the pharmacies.
+In other words, the expected travel time is the weighted sum of travel times to the nearest pharmacy, with the weights given by the expected modal split in the TAZ.
+
+The expected travel time $i$ was multiplied by the population in parcel $i$ to obtain a measure of person-hours of travel ($PHT$) as follows:
+$$
+PHT_i = P_i\cdot ett_i
+$$
+
+Please note that this paper is a reproducible research document [see @Brunsdon2020opening] conducted using open source tools for transportation analysis [@Lovelace2021open].
+The code and data necessary to reproduce the analysis are available in a public repository\footnote{\url{https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines}}.
+
+## Findings
+
+The top panel of Figure \ref{fig:maps-baseline} shows the average expected travel time by TAZ in Hamilton.
+It is apparent that travel times tend to be lower in much of suburban Hamilton, and higher in the urban core and some rural parts of the city, particularly to the west.
+This is unsurprising, given the higher probability of travel by car and the predominantly suburban character of the vaccination sites.
+However, even accounting for the distribution of population, this leads to large disparities in the number of person-hours of travel across the city, with a concentration of the burden of travel in the urban core and the rural west (see bottom panel of Figure \ref{fig:maps-baseline}).
+
+The disparities are not trivial.
+
+As seen in Table \ref{tab:distribution-results}, under the pilot program approximately `r round((table_comparison[5,2] + table_comparison[4,2]) / (sum(table_comparison[1:5,2])) * 100, 2)`\% of people live in DAs in the bottom 40% of the median household income scale, but they account for `r round((table_comparison[5,3] + table_comparison[4,3]) / (sum(table_comparison[1:5,3])) * 100, 2)`\% of the total person-hours of travel.
+In contrast, `r round((table_comparison[1,2] + table_comparison[2,2]) / (sum(table_comparison[1:5,2])) * 100, 2)`\% of people aged 55 to 69 in DAs in the top 40% of the median household income scale accrue only `r round((table_comparison[1,3] + table_comparison[2,3]) / (sum(table_comparison[1:5,3])) * 100, 2)`\% of the total person-hours of travel.
+Where the mean travel time of residents of DAs with high median household income is `r table_comparison[1, 4] %>% set_units(h) %>% set_units(min) %>% round() %>% drop_units()` minutes, residents of lower income DAs average `r table_comparison[5, 4] %>% set_units(h) %>% set_units(min) %>% round() %>% drop_units()` minutes in travel time.
+In addition to longer average travel time, residents in lower income DAs also see substantially larger variations in travel times, and some may face considerably longer travel times (see top-left panel in Figure \ref{fig:results}).
+
+There are also important disparities by region.
+As shown in Table \ref{tab:distribution-results}, the urban and rural populations in Hamilton are approximately `r round((table_comparison[6,2] + table_comparison[8,2])/ (sum(table_comparison[6:8,2])) * 100, 2)`\% of the population but they bear `r round((table_comparison[6,3] + table_comparison[8,3])/ (sum(table_comparison[6:8,3])) * 100, 2)`\% of the total person-hours of travel, with also much greater variability in expected travel times (Figure \ref{fig:results}, bottom-left panel).
+
+For comparison purposes we consider a scenario with some modest additions to the list of pharmacies in the provincial pilot.
+We repeat the analysis, but include the three urban sites shown in white circles in Figure \ref{fig:pharmacies-and-regions}.
+The results of this scenario appear in the last two columns of Table \ref{tab:distribution-results} and the two right panels of Figure \ref{fig:results}.
+We begin by noting that all income groups benefit from the addition of these three sites with shorter mean trip durations; the most remarkable difference is the large reduction in the disparities between residents in DAs with different levels of income.
+The top-right panel of Figure \ref{fig:results} shows that the distribution of expected travel time is now more in line for all income groups, even if the bottom two income quintiles still have somewhat wider spreads.
+With respect to the distribution of expected travel time and person-hours of travel by region, unsurprisingly the addition of three urban vaccination sites in the scenario makes a difference for urban residents but not for rural residents.
+
+The results indicate that the locations chosen by the province for the pilot vaccination program do not serve well urban or rural residents of the city, and there are some important questions regarding equity of access to the program, with a disproportionate burden in the cost of travel falling on lower income urban populations and rural populations.
+A scenario did not consider candidate sites in a systematic way.
+Nonetheless, selection of three sensible urban locations does much to alleviate disparities in the burden of transportation.
+On the other hand, unlike the urban context where there are numerous candidate locations that could be chosen for a vaccination program, there are not many candidate locations in rural regions of the city.
+Increasing access in rural Hamilton likely will involve an expansion of existing mobile vaccination pop-up clinics\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
+
+```{r figure-maps-baseline, echo=FALSE, warning=FALSE, fig.height=7, fig.align="center", fig.cap="\\label{fig:maps-baseline}Average expected travel time by TAZ (in minutes) and total person-hours of travel by TAZ."}
+grid.arrange(ttm_map_baseline +
+               theme(legend.title = element_text(size = 8),
+                     legend.text = element_text(size = 7), 
+                     axis.text = element_text(size = 7)),
+             pht_map_baseline +
+               theme(legend.title = element_text(size = 8),
+                     legend.text = element_text(size = 7), 
+                     axis.text = element_text(size = 7)),
+             nrow = 2)
+```
+
+```{r table-results, echo=FALSE}
+table_comparison  %>%
+  kable("html",
+        booktabs = TRUE,
+        col.names = c("Group", 
+                      "Population", 
+                      "Total PHT", 
+                      "Hours per person", 
+                      "Total PHT",
+                      "Hours per person"),
+        digits = 3,
+        align = c("lccccc"),
+        caption = "\\label{tab:distribution-results}Distribution of person-hours of travel (PHT) by median total household income and region: pilot locations only, and scenario with three urban locations added") %>%
+  add_header_above(c(" " = 2, "Pilot Program" = 2, "Scenario" = 2)) %>%
+  kable_styling(latex_options = c("striped", "scale_down")) %>%
+  pack_rows("Income Quintile", 1, 5) %>%
+  pack_rows("Region", 6, 8) %>%
+  footnote(general = "The population totals differ due to small differences in the classification of the regions")
+```
+
+```{r figure-results, echo=FALSE, warning=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:results}Distribution of expected travel time for different population groups"}
 grid.arrange(ttm_income_baseline + 
@@
 
-```{r table-results, echo=FALSE}
-cbind(person_hours_baseline %>%
-        dplyr::select(group, total_population, total_weighted_person_hours, ratio), 
-      person_hours %>%
-        dplyr::select(total_weighted_person_hours, ratio)) %>%
-  kable("html",
-        booktabs = TRUE,
-        col.names = c("Group", 
-                      "Population", 
-                      "Total PHT", 
-                      "Hours per person", 
-                      "Total PHT",
-                      "Hours per person"),
-        digits = 3,
-        align = c("lccccc"),
-        caption = "\\label{tab:distribution-results}Distribution of person hours travelled (PHT) by median total household income and region: pilot locations only, and scenario with three urban locations added") %>%
-  add_header_above(c(" " = 2, "Pilot Program" = 2, "Scenario" = 2)) %>%
-  kable_styling(latex_options = c("striped", "scale_down")) %>%
-  pack_rows("Income Quintile", 1, 5) %>%
-  pack_rows("Region", 6, 8) %>%
-  footnote(general = "The population totals differ due to small differences in the classification of the regions")
-```
-
-References {#references .unnumbered}
-==========
+## References {#references .unnumbered}

### 2021-04-06 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.Rmd
Update Accessibility-Vaccination-Sites-Hamilton.Rmd
edits to text

@@
 abstract: |
-  The province of Ontario in Canada announced the pilot for a new vaccination program, with designated pharmacies across the province now able to offer COVID-19 vaccines.
+  The Government of Ontario in Canada announced the pilot for a new vaccination program, with designated pharmacies across the province now able to offer COVID-19 vaccines.
 The accessibility of this program raises questions about the cost of travel and the distribution of the cost among the population.
-In our examination of the City of Hamilton we find that selected sites do not serve well the rural and urban population of Hamilton, and that the associated cost of travel is expected to be disproportionally borne by lower income populations.
+In our examination of the City of Hamilton we find that selected sites do not serve well rural and urban residents and that the associated cost of travel is expected to be disproportionally borne by lower income populations.
 Modest additions to the list of pilot sites in the city can substantially alleviate this inequity.
@@
 Along with the provision of health care facilities to treat severe cases of COVID-19 [@Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs.
-The Province of Ontario, in Canada, announced on April 1st 2021 the expansion of a pilot program to offer vaccines in pharmacies in the City of Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
+The Government of Ontario, in Canada, announced on April 1st 2021 the expansion of a pilot program to offer vaccines in pharmacies in the City of Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
 This program is in addition to dedicated vaccination centers for people aged 70+.
@@
 Given the target demographic for the program, it is possible that suburban sites could be convenient for mature adults and young old: the population aged 55 to 69 in Hamilton is approximately `r prettyNum(round((table_comparison[7,2]/1000), 2) * 1000, big.mark = ",")` suburban, `r prettyNum(round((table_comparison[8,2]/1000), 2) * 1000, big.mark = ",")` urban, and only `r prettyNum(round((table_comparison[6,2]/1000), 2) * 1000, big.mark = ",")` rural.
-Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the typically inert Major of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
-As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may introduce inequities in access.
+Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the Mayor of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
+As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may introduce significant inequities in access.
 
@@
         legend.text = element_text(size = 8))
-
 ```
@@
 
-### Open Hamilton
-
-From the open data portal of the City of Hamilton\footnote{\url{https://open.hamilton.ca/}} we obtained boundaries for the city's various regions (the definition of urban, suburban, and rural regions follows the classification of development applications).
-
-### Statistics Canada
-
-Population and income statistics at the level of Dissemination Areas (DAs) were retrieved using the package `cancensus` [@vonBergmann2021cancensus].
-Dissemination Areas are the smallest publicly available census geography in Canada.
-We use data from the 2016 Population Census.
-
-### Transportation Tomorrow Survey
-
-Data about modal split by age by place of residence were downloaded from the Data Retrieval System of the Transportation Tomorrow Survey (TTS)\footnote{\url{http://dmg.utoronto.ca/}}.
-The data are geocoded at the level of Traffic Analysis Zones (TAZ).
-
-### Other
-
-The locations of pharmacies in the pilot were obtained from public records and geocoded.
-Three urban sites not in the program were also identified and geocoded for comparison purposes.
-In addition, we converted all recorded residential parcels in the City of Hamilton to points on the road network.
-Each point includes information about the number of residential units in the parcel.
+- Urban, suburban, and rural boundary definitions from the City of Hamilton\footnote{\url{https://open.hamilton.ca/}}
+- Population and median total household income for 2016 by Dissemination Area (DA) boundary using the `cancensus` [@vonBergmann2021cancensus] package
+- Modal split by age for traffic analysis zones (TAZ) from the 2016 Transportation Tomorrow Survey (TTS)\footnote{\url{http://dmg.utoronto.ca/}}
+- Locations of pilot pharmacies from public records
+- Locations of three additional major chain pharmacies for scenario testing
+- Residential units by land parcel
 
@@
 
-We used the population aged 55 to 59 y.o.
-in each Dissemination Area to calculate the average number of people per dwelling.
-This value was then assigned proportionally to the number of dwellings per parcel.
-The median total household income of the corresponding DAs was joined to the parcels.
-In addition, we calculated the proportion of trips by mode from the total number of trips by each of the three modes retrieved from the TTS data.
-These proportions were joined to the parcel data based on their corresponding TAZ.
-The package `r5r` [@Pereira2021r5r] was used to calculate the travel time from each parcel to all pharmacies by three modes: car, transit, and walking.
-For routing purposes we used a cutoff value of 180 min and a maximum walking distance of 10,000 m.
-
+Using the population aged 55 to 59 y.o.
+we first calculate the average number of people per dwelling and assign them proportionally to the dwellings by parcel.
+Second, median incomes and modal splits across walking, transit, and car are joined to the parcels.
+Although our data are from 2016 (the most recent available), we assume they are largely representative of current spatial and demographic trends in the city.
+Third, we use the `r5r` [@Pereira2021r5r] package to calculate the travel time from each parcel to all pharmacies by three modes using a cutoff value of 180 min and a maximum walking distance of 10,000 m.
 Once we obtained travel time tables with population, proportion of trips by mode, and income information, we calculated the expected travel time $ett$ from each parcel $i$ to a pharmacy $j$ as follows:
@@
 
-The expected travel time $i$ was multiplied by the population in parcel $i$ to obtain a measure of person-hours of travel ($PHT$) as follows:
+The expected travel time $i$ was multiplied by the assigned population in parcel $i$ to obtain a measure of person-hours of travel ($PHT$) as follows:
 $$
@@
 
-The top panel of Figure \ref{fig:maps-baseline} shows the average expected travel time by TAZ in Hamilton.
-It is apparent that travel times tend to be lower in much of suburban Hamilton, and higher in the urban core and some rural parts of the city, particularly to the west.
+The top panel of Figure \ref{fig:maps-baseline} shows the average expected use-weighted multi-modal travel times by TAZ in Hamilton.
+It is apparent that travel times tend to be lower in much of suburban Hamilton and higher in the urban core and some rural parts of the city, particularly to the west.
 This is unsurprising, given the higher probability of travel by car and the predominantly suburban character of the vaccination sites.
@@
 The results of this scenario appear in the last two columns of Table \ref{tab:distribution-results} and the two right panels of Figure \ref{fig:results}.
-We begin by noting that all income groups benefit from the addition of these three sites with shorter mean trip durations; the most remarkable difference is the large reduction in the disparities between residents in DAs with different levels of income.
+While all income groups benefit from the addition of these three sites with shorter mean trip durations, the most remarkable difference is the large reduction in the disparities between residents in DAs with different levels of income.
 The top-right panel of Figure \ref{fig:results} shows that the distribution of expected travel time is now more in line for all income groups, even if the bottom two income quintiles still have somewhat wider spreads.
-With respect to the distribution of expected travel time and person-hours of travel by region, unsurprisingly the addition of three urban vaccination sites in the scenario makes a difference for urban residents but not for rural residents.
+Unsurprisingly, the addition of three urban vaccination sites does not have a large impact for rural residents.
 
 The results indicate that the locations chosen by the province for the pilot vaccination program do not serve well urban or rural residents of the city, and there are some important questions regarding equity of access to the program, with a disproportionate burden in the cost of travel falling on lower income urban populations and rural populations.
-A scenario did not consider candidate sites in a systematic way.
-Nonetheless, selection of three sensible urban locations does much to alleviate disparities in the burden of transportation.
-On the other hand, unlike the urban context where there are numerous candidate locations that could be chosen for a vaccination program, there are not many candidate locations in rural regions of the city.
-Increasing access in rural Hamilton likely will involve an expansion of existing mobile vaccination pop-up clinics\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
+Selection of three sensible urban locations does much to alleviate disparities in the burden of transportation.
+On the other hand, because there are not many candidate locations in rural parts of the city, increasing access for these residents likely will involve an expansion of the existing mobile vaccination pop-up clinic program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
 

### 2021-04-06 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.Rmd
v1 Final edits before submission to Transport Findings

@@
 
-Research Questions and Hypotheses
+Questions
 ==========================
@@
 Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast majority of the pharmacies are in suburban Hamilton.
-The issue is somewhat less clear-cut when we consider that Hamilton's population skews suburban (see Figure \ref{fig:population-map}).
-Given the target demographic for the program, it is possible that suburban sites could be convenient for mature adults and young old: the population aged 55 to 69 in Hamilton is approximately `r prettyNum(round((table_comparison[7,2]/1000), 2) * 1000, big.mark = ",")` suburban, `r prettyNum(round((table_comparison[8,2]/1000), 2) * 1000, big.mark = ",")` urban, and only `r prettyNum(round((table_comparison[6,2]/1000), 2) * 1000, big.mark = ",")` rural.
+The issue is less clear-cut when we consider that Hamilton's older population skews suburban (see Figure \ref{fig:population-map}).
+Given the target demographic for the program, it is possible that suburban sites could be convenient for mature adults and the young old: the population aged 55 to 69 in Hamilton is approximately `r prettyNum(round((table_comparison[7,2]/1000), 2) * 1000, big.mark = ",")` suburban, `r prettyNum(round((table_comparison[8,2]/1000), 2) * 1000, big.mark = ",")` urban, and only `r prettyNum(round((table_comparison[6,2]/1000), 2) * 1000, big.mark = ",")` rural.
 Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the Mayor of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
-As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may introduce significant inequities in access.
+As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may lead to inequities in access.
 
@@
 
-```{r pharmacies-map, echo=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:pharmacies-and-regions}Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) and urban locations for scenario analysis (white circles)"}
+```{r pharmacies-map, echo=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:pharmacies-and-regions}Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) along with urban locations for scenario analysis (white circles)."}
 #map_pharmacies <-
@@
 
-```{r population-map, echo=FALSE, warning=FALSE, fig.height=8, fig.align="center", fig.cap="\\label{fig:population-map}Distribution of population age 55+ in the City of Hamilton"}
+```{r save-pharmacies-map-1, include=FALSE}
+ggplot() +
+  geom_sf(data = urban_types,
+          aes(fill = Type),
+          color = "lightgray") +
+  geom_sf(data = pharmacy_locations %>%
+            filter(Type == "Pilot"),
+          aes(shape = Type,
+              color = Type),
+          size = 2.5,
+          show.legend = FALSE) +
+  scale_shape_manual(values = c("Candidate" = 16,
+                                "Pilot" = 17)) +
+  scale_color_manual(values = c("Candidate" = "white",
+                                "Pilot" = "black")) +
+  scale_fill_manual(values = c("Urban" = "firebrick2",
+                               "Suburban" = "cornflowerblue",
+                               "Rural" = "forestgreen")) +
+  guides(fill = guide_legend(title = "")) + 
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+ggsave("images/pharmacies-map-1.jpeg")
+```
+
+```{r save-pharmacies-map-2, include=FALSE}
+ggplot() +
+  geom_sf(data = urban_types,
+          aes(fill = Type),
+          color = "lightgray") +
+  geom_sf(data = pharmacy_locations,
+          aes(shape = Type,
+              color = Type),
+          size = 2.5,
+          show.legend = FALSE) +
+  scale_shape_manual(values = c("Candidate" = 16,
+                                "Pilot" = 17)) +
+  scale_color_manual(values = c("Candidate" = "white",
+                                "Pilot" = "black")) +
+  scale_fill_manual(values = c("Urban" = "firebrick2",
+                               "Suburban" = "cornflowerblue",
+                               "Rural" = "forestgreen")) +
+  guides(fill = guide_legend(title = "")) + 
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+ggsave("images/pharmacies-map-2.jpeg")
+```
+
+```{r population-map, echo=FALSE, warning=FALSE, fig.height=8, fig.align="center", fig.cap="\\label{fig:population-map}Distribution of population aged 55+ in the City of Hamilton."}
 ggplot() +
@@
 
-Methods and Data
+```{r save-population-map, include=FALSE}
+ggplot() +
+  geom_sf(data = data_da_2016 %>%
+            filter(`Region Name` == "Hamilton") %>%
+            dplyr::select(Population_55to69, Population_70plus) %>%
+            pivot_longer(cols = -geometry, names_to = "Age", values_to = "Population") %>%
+            mutate(Age = factor(Age,
+                                levels = c("Population_55to69",
+                                           "Population_70plus"),
+                                labels = c("Age: 55 to 69",
+                                           "Age: 70+"),
+                                ordered = TRUE)) %>%
+            st_as_sf(),
+          aes(fill = Population),
+          color = NA) +
+  geom_sf(data = urban_types,
+          aes(color = Type),
+          fill = NA) +
+  scale_fill_distiller(name = "Population",
+                       breaks = c(20, 50, 150, 400),
+                       palette = "OrRd", 
+                       trans = "log",
+                       direction = 1) +
+  scale_color_manual(name = "",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen"))  +
+  facet_wrap(~ Age, nrow = 2) +
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+ggsave("images/population-map.jpeg")
+```
+
+Methods
 ============
 
-# Data
-
 We use data from the following sources.
@@
 - Urban, suburban, and rural boundary definitions from the City of Hamilton\footnote{\url{https://open.hamilton.ca/}}
-- Population and median total household income for 2016 by Dissemination Area (DA) boundary using the `cancensus` [@vonBergmann2021cancensus] package
+- Population and median total household income for 2016 by Dissemination Area (DA) boundary using the `cancensus` package [@vonBergmann2021cancensus]
 - Modal split by age for traffic analysis zones (TAZ) from the 2016 Transportation Tomorrow Survey (TTS)\footnote{\url{http://dmg.utoronto.ca/}}
@@
 - Residential units by land parcel
-
-# Methods
 
@@
 \noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the vector of travel times from parcel $i$ to the pharmacies.
-In other words, the expected travel time is the weighted sum of travel times to the nearest pharmacy, with the weights given by the expected modal split in the TAZ.
+The expected travel time is thus the weighted sum of travel times to the nearest pharmacy, with the weights given by the expected modal split in the TAZ.
 
@@
 The results of this scenario appear in the last two columns of Table \ref{tab:distribution-results} and the two right panels of Figure \ref{fig:results}.
-While all income groups benefit from the addition of these three sites with shorter mean trip durations, the most remarkable difference is the large reduction in the disparities between residents in DAs with different levels of income.
+While all income groups benefit from the addition of these three sites with shorter mean trip durations, the most remarkable difference is the large reduction in the disparities between residents in DAs with lower levels of income.
 The top-right panel of Figure \ref{fig:results} shows that the distribution of expected travel time is now more in line for all income groups, even if the bottom two income quintiles still have somewhat wider spreads.
@@
 Selection of three sensible urban locations does much to alleviate disparities in the burden of transportation.
-On the other hand, because there are not many candidate locations in rural parts of the city, increasing access for these residents likely will involve an expansion of the existing mobile vaccination pop-up clinic program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
+On the other hand, because there are not many candidate locations in rural parts of the city, increasing access for residents there likely necessitates an expansion of the existing mobile vaccination pop-up clinic program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
 
@@
              nrow = 2)
+```
+
+```{r save-figure-maps-baseline, include=FALSE}
+jpeg(filename = "images/baseline-maps.jpeg")
+grid.arrange(ttm_map_baseline +
+               theme(legend.title = element_text(size = 8),
+                     legend.text = element_text(size = 7), 
+                     axis.text = element_text(size = 7)),
+             pht_map_baseline +
+               theme(legend.title = element_text(size = 8),
+                     legend.text = element_text(size = 7), 
+                     axis.text = element_text(size = 7)),
+             nrow = 2)
+dev.off()
+
+ttm_map_baseline
+ggsave("images/ttm-baseline-map.jpeg")
+
+pht_map_baseline
+ggsave("images/pht-baseline-map.jpeg")
 ```
@@
 
-```{r figure-results, echo=FALSE, warning=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:results}Distribution of expected travel time for different population groups"}
+```{r figure-results, echo=FALSE, warning=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:results}Distribution of expected travel time for different population groups."}
 grid.arrange(ttm_income_baseline + 
@@
 
+```{r save-figure-results, include=FALSE}
+jpeg(filename = "images/figure-results.jpeg")
+grid.arrange(ttm_income_baseline + 
+               ggtitle("By income: Pilot Program") + 
+               labs(x = "") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8), 
+                     axis.text = element_text(size = 7), 
+                     legend.position = "none"), 
+             ttm_income + 
+               ggtitle("By income: Scenario") + 
+               labs(x = "", y = "") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8), 
+                     axis.text = element_text(size = 7)), 
+             ttm_region_baseline + 
+               ggtitle("By region: Pilot Program") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8),
+                     axis.title = element_text(size = 8), 
+                     axis.text = element_text(size = 7), 
+                     legend.position = "none"), 
+             ttm_region + 
+               ggtitle("By region: Scenario") + 
+               labs(y = "") +
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8),
+                     axis.title = element_text(size = 8), 
+                     axis.text = element_text(size = 7)), 
+             nrow = 2)
+dev.off()
+
+ttm_income_baseline + 
+               ggtitle("By income: Pilot Program") + 
+               labs(x = "") + 
+               xlim(c(0,40))
+ggsave("images/ttm-income-pilot.jpeg")
+
+ttm_region_baseline + 
+  ggtitle("By region: Pilot Program") + 
+  xlim(c(0,40))
+ggsave("images/ttm-region-pilot.jpeg")
+
+ttm_income +
+  ggtitle("By income: Scenario") + 
+  labs(x = "", y = "") + 
+  xlim(c(0,40))
+ggsave("images/ttm-income.jpeg")
+
+ttm_region + 
+  ggtitle("By region: Scenario") + 
+  labs(y = "") +
+  xlim(c(0,40))
+ggsave("images/ttm-region.jpeg")
+```
+
 References {#references .unnumbered}

### 2021-04-06 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.tex
v1 Final edits before submission to Transport Findings

@@
   \begin{abstract}
-  The province of Ontario in Canada announced the pilot for a new
+  The Government of Ontario in Canada announced the pilot for a new
   vaccination program, with designated pharmacies across the province now
@@
   cost among the population. In our examination of the City of Hamilton we
-  find that selected sites do not serve well the rural and urban
-  population of Hamilton, and that the associated cost of travel is
-  expected to be disproportionally borne by lower income populations.
-  Modest additions to the list of pilot sites in the city can
-  substantially alleviate this inequity.
+  find that selected sites do not serve well rural and urban residents and
+  that the associated cost of travel is expected to be disproportionally
+  borne by lower income populations. Modest additions to the list of pilot
+  sites in the city can substantially alleviate this inequity.
   \end{abstract}
@@
 
-\hypertarget{research-questions-and-hypotheses}{%
-\section{Research Questions and
-Hypotheses}\label{research-questions-and-hypotheses}}
+\hypertarget{questions}{%
+\section{Questions}\label{questions}}
 
@@
 the fight against the pandemic is the rolling out of vaccination
-programs. The Province of Ontario, in Canada, announced on April 1st
+programs. The Government of Ontario, in Canada, announced on April 1st
 2021 the expansion of a pilot program to offer vaccines in pharmacies in
@@
 Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast
-majority of the pharmacies are in suburban Hamilton. The issue is
-somewhat less clear-cut when we consider that Hamilton's population
-skews suburban (see Figure \ref{fig:population-map}). Given the target
+majority of the pharmacies are in suburban Hamilton. The issue is less
+clear-cut when we consider that Hamilton's older population skews
+suburban (see Figure \ref{fig:population-map}). Given the target
 demographic for the program, it is possible that suburban sites could be
-convenient for mature adults and young old: the population aged 55 to 69
-in Hamilton is approximately 58,710 suburban, 35,490 urban, and only
-8,360 rural. Nevertheless, the selection of sites by the province raises
-some important
-questions\footnote{The decision-making process to select these sites appears to have been opaque, and the typically inert Major of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
+convenient for mature adults and the young old: the population aged 55
+to 69 in Hamilton is approximately 58,710 suburban, 35,490 urban, and
+only 8,360 rural. Nevertheless, the selection of sites by the province
+raises some important
+questions\footnote{The decision-making process to select these sites appears to have been opaque, and the Mayor of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
 As Yu et al. (2021) note, good geographical coverage is a key element
 for a successful vaccination campaign; at the same time, siting
-vaccinations sites in car-oriented locations may introduce inequities in
+vaccinations sites in car-oriented locations may lead to inequities in
 access.
@@
 
-\caption{\label{fig:pharmacies-and-regions}Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) and urban locations for scenario analysis (white circles)}\label{fig:pharmacies-map}
+\caption{\label{fig:pharmacies-and-regions}Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) along with urban locations for scenario analysis (white circles).}\label{fig:pharmacies-map}
 \end{figure}
@@
 
-\caption{\label{fig:population-map}Distribution of population age 55+ in the City of Hamilton}\label{fig:population-map}
+\caption{\label{fig:population-map}Distribution of population aged 55+ in the City of Hamilton.}\label{fig:population-map}
 \end{figure}
-
-\hypertarget{methods-and-data}{%
-\section{Methods and Data}\label{methods-and-data}}
-
-\hypertarget{data}{%
-\section{Data}\label{data}}
-
-We use data from the following sources.
-
-\hypertarget{open-hamilton}{%
-\subsubsection{Open Hamilton}\label{open-hamilton}}
-
-From the open data portal of the City of
-Hamilton\footnote{\url{https://open.hamilton.ca/}} we obtained
-boundaries for the city's various regions (the definition of urban,
-suburban, and rural regions follows the classification of development
-applications).
-
-\hypertarget{statistics-canada}{%
-\subsubsection{Statistics Canada}\label{statistics-canada}}
-
-Population and income statistics at the level of Dissemination Areas
-(DAs) were retrieved using the package \texttt{cancensus} (von Bergmann
-et al., 2021). Dissemination Areas are the smallest publicly available
-census geography in Canada. We use data from the 2016 Population Census.
-
-\hypertarget{transportation-tomorrow-survey}{%
-\subsubsection{Transportation Tomorrow
-Survey}\label{transportation-tomorrow-survey}}
-
-Data about modal split by age by place of residence were downloaded from
-the Data Retrieval System of the Transportation Tomorrow Survey
-(TTS)\footnote{\url{http://dmg.utoronto.ca/}}. The data are geocoded at
-the level of Traffic Analysis Zones (TAZ).
-
-\hypertarget{other}{%
-\subsubsection{Other}\label{other}}
-
-The locations of pharmacies in the pilot were obtained from public
-records and geocoded. Three urban sites not in the program were also
-identified and geocoded for comparison purposes. In addition, we
-converted all recorded residential parcels in the City of Hamilton to
-points on the road network. Each point includes information about the
-number of residential units in the parcel.
 
@@
 
-We used the population aged 55 to 59 y.o. in each Dissemination Area to
-calculate the average number of people per dwelling. This value was then
-assigned proportionally to the number of dwellings per parcel. The
-median total household income of the corresponding DAs was joined to the
-parcels. In addition, we calculated the proportion of trips by mode from
-the total number of trips by each of the three modes retrieved from the
-TTS data. These proportions were joined to the parcel data based on
-their corresponding TAZ. The package \texttt{r5r} (Pereira, Saraiva, et
-al., 2021) was used to calculate the travel time from each parcel to all
-pharmacies by three modes: car, transit, and walking. For routing
-purposes we used a cutoff value of 180 min and a maximum walking
-distance of 10,000 m.
+We use data from the following sources.
 
-Once we obtained travel time tables with population, proportion of trips
-by mode, and income information, we calculated the expected travel time
-\(ett\) from each parcel \(i\) to a pharmacy \(j\) as follows: \[
+\begin{itemize}
+\tightlist
+\item
+  Urban, suburban, and rural boundary definitions from the City of
+  Hamilton\footnote{\url{https://open.hamilton.ca/}}
+\item
+  Population and median total household income for 2016 by Dissemination
+  Area (DA) boundary using the \texttt{cancensus} package (von Bergmann
+  et al., 2021)
+\item
+  Modal split by age for traffic analysis zones (TAZ) from the 2016
+  Transportation Tomorrow Survey
+  (TTS)\footnote{\url{http://dmg.utoronto.ca/}}
+\item
+  Locations of pilot pharmacies from public records
+\item
+  Locations of three additional major chain pharmacies for scenario
+  testing
+\item
+  Residential units by land parcel
+\end{itemize}
+
+Using the population aged 55 to 59 y.o. we first calculate the average
+number of people per dwelling and assign them proportionally to the
+dwellings by parcel. Second, median incomes and modal splits across
+walking, transit, and car are joined to the parcels. Although our data
+are from 2016 (the most recent available), we assume they are largely
+representative of current spatial and demographic trends in the city.
+Third, we use the \texttt{r5r} (Pereira, Saraiva, et al., 2021) package
+to calculate the travel time from each parcel to all pharmacies by three
+modes using a cutoff value of 180 min and a maximum walking distance of
+10,000 m. Once we obtained travel time tables with population,
+proportion of trips by mode, and income information, we calculated the
+expected travel time \(ett\) from each parcel \(i\) to a pharmacy \(j\)
+as follows: \[
 ett_i = p^c_i \min(tt^c_{ij}) + p^t_i \min(tt^t_{ij}) + p^w_i \min(tt^w_{ij})
@@
 the TAZ of parcel \(i\), and \(tt^k_{ij}\) is the vector of travel times
-from parcel \(i\) to the pharmacies. In other words, the expected travel
-time is the weighted sum of travel times to the nearest pharmacy, with
-the weights given by the expected modal split in the TAZ.
+from parcel \(i\) to the pharmacies. The expected travel time is thus
+the weighted sum of travel times to the nearest pharmacy, with the
+weights given by the expected modal split in the TAZ.
 
-The expected travel time \(i\) was multiplied by the population in
-parcel \(i\) to obtain a measure of person-hours of travel (\(PHT\)) as
-follows: \[
+The expected travel time \(i\) was multiplied by the assigned population
+in parcel \(i\) to obtain a measure of person-hours of travel (\(PHT\))
+as follows: \[
 PHT_i = P_i\cdot ett_i
@@
 The top panel of Figure \ref{fig:maps-baseline} shows the average
-expected travel time by TAZ in Hamilton. It is apparent that travel
-times tend to be lower in much of suburban Hamilton, and higher in the
-urban core and some rural parts of the city, particularly to the west.
-This is unsurprising, given the higher probability of travel by car and
-the predominantly suburban character of the vaccination sites. However,
-even accounting for the distribution of population, this leads to large
-disparities in the number of person-hours of travel across the city,
-with a concentration of the burden of travel in the urban core and the
-rural west (see bottom panel of Figure \ref{fig:maps-baseline}).
+expected use-weighted multi-modal travel times by TAZ in Hamilton. It is
+apparent that travel times tend to be lower in much of suburban Hamilton
+and higher in the urban core and some rural parts of the city,
+particularly to the west. This is unsurprising, given the higher
+probability of travel by car and the predominantly suburban character of
+the vaccination sites. However, even accounting for the distribution of
+population, this leads to large disparities in the number of
+person-hours of travel across the city, with a concentration of the
+burden of travel in the urban core and the rural west (see bottom panel
+of Figure \ref{fig:maps-baseline}).
 
@@
 appear in the last two columns of Table \ref{tab:distribution-results}
-and the two right panels of Figure \ref{fig:results}. We begin by noting
-that all income groups benefit from the addition of these three sites
-with shorter mean trip durations; the most remarkable difference is the
-large reduction in the disparities between residents in DAs with
-different levels of income. The top-right panel of Figure
-\ref{fig:results} shows that the distribution of expected travel time is
-now more in line for all income groups, even if the bottom two income
-quintiles still have somewhat wider spreads. With respect to the
-distribution of expected travel time and person-hours of travel by
-region, unsurprisingly the addition of three urban vaccination sites in
-the scenario makes a difference for urban residents but not for rural
-residents.
+and the two right panels of Figure \ref{fig:results}. While all income
+groups benefit from the addition of these three sites with shorter mean
+trip durations, the most remarkable difference is the large reduction in
+the disparities between residents in DAs with lower levels of income.
+The top-right panel of Figure \ref{fig:results} shows that the
+distribution of expected travel time is now more in line for all income
+groups, even if the bottom two income quintiles still have somewhat
+wider spreads. Unsurprisingly, the addition of three urban vaccination
+sites does not have a large impact for rural residents.
 
@@
 travel falling on lower income urban populations and rural populations.
-A scenario did not consider candidate sites in a systematic way.
-Nonetheless, selection of three sensible urban locations does much to
-alleviate disparities in the burden of transportation. On the other
-hand, unlike the urban context where there are numerous candidate
-locations that could be chosen for a vaccination program, there are not
-many candidate locations in rural regions of the city. Increasing access
-in rural Hamilton likely will involve an expansion of existing mobile
-vaccination pop-up
-clinics\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
+Selection of three sensible urban locations does much to alleviate
+disparities in the burden of transportation. On the other hand, because
+there are not many candidate locations in rural parts of the city,
+increasing access for residents there likely necessitates an expansion
+of the existing mobile vaccination pop-up clinic
+program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
 
@@
 
-\caption{\label{fig:results}Distribution of expected travel time for different population groups}\label{fig:figure-results}
+\caption{\label{fig:results}Distribution of expected travel time for different population groups.}\label{fig:figure-results}
 \end{figure}

### 2021-04-06 README.Rmd
v1 Final edits before submission to Transport Findings

@@
 
-The province of Ontario in Canada announced the pilot for a new vaccination program, with designated pharmacies across the province now able to offer COVID-19 vaccines.
+The Government of Ontario in Canada announced the pilot for a new vaccination program, with designated pharmacies across the province now able to offer COVID-19 vaccines.
 The accessibility of this program raises questions about the cost of travel and the distribution of the cost among the population.
-In our examination of the City of Hamilton we find that selected sites do not serve well the rural and urban population of Hamilton, and that the associated cost of travel is expected to be disproportionally borne by lower income populations.
+In our examination of the City of Hamilton we find that selected sites do not serve well rural and urban residents and that the associated cost of travel is expected to be disproportionally borne by lower income populations.
 Modest additions to the list of pilot sites in the city can substantially alleviate this inequity.
@@
 
-## Research Questions and Hypotheses
+## Questions
 
 Along with the provision of health care facilities to treat severe cases of COVID-19 [@Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs.
-The Province of Ontario, in Canada, announced on April 1st 2021 the expansion of a pilot program to offer vaccines in pharmacies in the City of Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
+The Government of Ontario, in Canada, announced on April 1st 2021 the expansion of a [pilot program](https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704) to offer vaccines in pharmacies in the City of Hamilton.
 This program is in addition to dedicated vaccination centers for people aged 70+.
@@
 
-Critics were swift to point out that the list of pharmacies approved for Hamilton by the province were mostly located in lower density parts of the city that are not well serviced by transit and are difficult to reach by foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}.
-Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast majority of the pharmacies are in suburban Hamilton.
-The issue is somewhat less clear-cut when we consider that Hamilton's population skews suburban (see Figure \ref{fig:population-map}).
-Given the target demographic for the program, it is possible that suburban sites could be convenient for mature adults and young old: the population aged 55 to 69 in Hamilton is approximately `r prettyNum(round((table_comparison[7,2]/1000), 2) * 1000, big.mark = ",")` suburban, `r prettyNum(round((table_comparison[8,2]/1000), 2) * 1000, big.mark = ",")` urban, and only `r prettyNum(round((table_comparison[6,2]/1000), 2) * 1000, big.mark = ",")` rural.
-Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the typically inert Major of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
-As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may introduce inequities in access.
+[Critics](https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20) were swift to point out that the list of pharmacies approved for Hamilton by the province were mostly located in [lower density](https://twitter.com/NrinderWard3/status/1378679195514060801?s=20) parts of the city that are not well serviced by transit and are difficult to reach by foot.
+Indeed, as seen in Figure 1, a vast majority of the pharmacies are in suburban Hamilton.
+The issue is less clear-cut when we consider that Hamilton's older population skews suburban (see Figure 2).
+Given the target demographic for the program, it is possible that suburban sites could be convenient for mature adults and the young old: the population aged 55 to 69 in Hamilton is approximately `r prettyNum(round((table_comparison[7,2]/1000), 2) * 1000, big.mark = ",")` suburban, `r prettyNum(round((table_comparison[8,2]/1000), 2) * 1000, big.mark = ",")` urban, and only `r prettyNum(round((table_comparison[6,2]/1000), 2) * 1000, big.mark = ",")` rural.
+Nevertheless, the selection of sites by the province raises some important questions and the decision-making process to select these sites appears to have been opaque, with the Mayor of the city caught [flat footed](https://twitter.com/FredEisenberger/status/1378350123114242053?s=20) by the announcement.
+As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may lead to inequities in access.
 
@@
 
-```{r pharmacies-map, echo=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:pharmacies-and-regions}Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) and urban locations for scenario analysis (white circles)"}
+```{r pharmacies-map, echo=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:pharmacies-and-regions}Figure 1.
+Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) along with urban locations for scenario analysis (white circles)."}
 #map_pharmacies <-
@@
         legend.text = element_text(size = 8))
-
-```
-
-```{r population-map, echo=FALSE, warning=FALSE, fig.height=8, fig.align="center", fig.cap="\\label{fig:population-map}Distribution of population age 55+ in the City of Hamilton"}
+```
+
+```{r save-pharmacies-map-1, include=FALSE, eval=FALSE}
+ggplot() +
+  geom_sf(data = urban_types,
+          aes(fill = Type),
+          color = "lightgray") +
+  geom_sf(data = pharmacy_locations %>%
+            filter(Type == "Pilot"),
+          aes(shape = Type,
+              color = Type),
+          size = 2.5,
+          show.legend = FALSE) +
+  scale_shape_manual(values = c("Candidate" = 16,
+                                "Pilot" = 17)) +
+  scale_color_manual(values = c("Candidate" = "white",
+                                "Pilot" = "black")) +
+  scale_fill_manual(values = c("Urban" = "firebrick2",
+                               "Suburban" = "cornflowerblue",
+                               "Rural" = "forestgreen")) +
+  guides(fill = guide_legend(title = "")) + 
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+ggsave("images/pharmacies-map-1.jpeg")
+```
+
+```{r save-pharmacies-map-2, include=FALSE, eval=FALSE}
+ggplot() +
+  geom_sf(data = urban_types,
+          aes(fill = Type),
+          color = "lightgray") +
+  geom_sf(data = pharmacy_locations,
+          aes(shape = Type,
+              color = Type),
+          size = 2.5,
+          show.legend = FALSE) +
+  scale_shape_manual(values = c("Candidate" = 16,
+                                "Pilot" = 17)) +
+  scale_color_manual(values = c("Candidate" = "white",
+                                "Pilot" = "black")) +
+  scale_fill_manual(values = c("Urban" = "firebrick2",
+                               "Suburban" = "cornflowerblue",
+                               "Rural" = "forestgreen")) +
+  guides(fill = guide_legend(title = "")) + 
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+ggsave("images/pharmacies-map-2.jpeg")
+```
+
+```{r population-map, echo=FALSE, warning=FALSE, fig.height=8, fig.align="center", fig.cap="\\label{fig:population-map}Figure 2. Distribution of population aged 55+ in the City of Hamilton."}
 ggplot() +
@@
 
-## Methods and Data
-
-## Data
+```{r save-population-map, include=FALSE, eval=FALSE}
+ggplot() +
+  geom_sf(data = data_da_2016 %>%
+            filter(`Region Name` == "Hamilton") %>%
+            dplyr::select(Population_55to69, Population_70plus) %>%
+            pivot_longer(cols = -geometry, names_to = "Age", values_to = "Population") %>%
+            mutate(Age = factor(Age,
+                                levels = c("Population_55to69",
+                                           "Population_70plus"),
+                                labels = c("Age: 55 to 69",
+                                           "Age: 70+"),
+                                ordered = TRUE)) %>%
+            st_as_sf(),
+          aes(fill = Population),
+          color = NA) +
+  geom_sf(data = urban_types,
+          aes(color = Type),
+          fill = NA) +
+  scale_fill_distiller(name = "Population",
+                       breaks = c(20, 50, 150, 400),
+                       palette = "OrRd", 
+                       trans = "log",
+                       direction = 1) +
+  scale_color_manual(name = "",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen"))  +
+  facet_wrap(~ Age, nrow = 2) +
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+ggsave("images/population-map.jpeg")
+```
+
+## Methods
 
@@
 
-### Open Hamilton
-
-From the open data portal of the City of Hamilton\footnote{\url{https://open.hamilton.ca/}} we obtained boundaries for the city's various regions (the definition of urban, suburban, and rural regions follows the classification of development applications).
-
-### Statistics Canada
-
-Population and income statistics at the level of Dissemination Areas (DAs) were retrieved using the package `cancensus` [@vonBergmann2021cancensus].
-Dissemination Areas are the smallest publicly available census geography in Canada.
-We use data from the 2016 Population Census.
-
-### Transportation Tomorrow Survey
-
-Data about modal split by age by place of residence were downloaded from the Data Retrieval System of the Transportation Tomorrow Survey (TTS)\footnote{\url{http://dmg.utoronto.ca/}}.
-The data are geocoded at the level of Traffic Analysis Zones (TAZ).
-
-### Other
-
-The locations of pharmacies in the pilot were obtained from public records and geocoded.
-Three urban sites not in the program were also identified and geocoded for comparison purposes.
-In addition, we converted all recorded residential parcels in the City of Hamilton to points on the road network.
-Each point includes information about the number of residential units in the parcel.
-
-## Methods
-
-We used the population aged 55 to 59 y.o.
-in each Dissemination Area to calculate the average number of people per dwelling.
-This value was then assigned proportionally to the number of dwellings per parcel.
-The median total household income of the corresponding DAs was joined to the parcels.
-In addition, we calculated the proportion of trips by mode from the total number of trips by each of the three modes retrieved from the TTS data.
-These proportions were joined to the parcel data based on their corresponding TAZ.
-The package `r5r` [@Pereira2021r5r] was used to calculate the travel time from each parcel to all pharmacies by three modes: car, transit, and walking.
-For routing purposes we used a cutoff value of 180 min and a maximum walking distance of 10,000 m.
-
+- Urban, suburban, and rural boundary definitions from the [City of Hamilton](https://open.hamilton.ca/)
+- Population and median total household income for 2016 by Dissemination Area (DA) boundary using the `cancensus` package [@vonBergmann2021cancensus]
+- Modal split by age for traffic analysis zones (TAZ) from the 2016 Transportation Tomorrow Survey ([TTS](http://dmg.utoronto.ca/))
+- Locations of pilot pharmacies from public records
+- Locations of three additional major chain pharmacies for scenario testing
+- Residential units by land parcel
+
+Using the population aged 55 to 59 y.o.
+we first calculate the average number of people per dwelling and assign them proportionally to the dwellings by parcel.
+Second, median incomes and modal splits across walking, transit, and car are joined to the parcels.
+Although our data are from 2016 (the most recent available), we assume they are largely representative of current spatial and demographic trends in the city.
+Third, we use the `r5r` [@Pereira2021r5r] package to calculate the travel time from each parcel to all pharmacies by three modes using a cutoff value of 180 min and a maximum walking distance of 10,000 m.
 Once we obtained travel time tables with population, proportion of trips by mode, and income information, we calculated the expected travel time $ett$ from each parcel $i$ to a pharmacy $j$ as follows:
@@
 \noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the vector of travel times from parcel $i$ to the pharmacies.
-In other words, the expected travel time is the weighted sum of travel times to the nearest pharmacy, with the weights given by the expected modal split in the TAZ.
-
-The expected travel time $i$ was multiplied by the population in parcel $i$ to obtain a measure of person-hours of travel ($PHT$) as follows:
+The expected travel time is thus the weighted sum of travel times to the nearest pharmacy, with the weights given by the expected modal split in the TAZ.
+
+The expected travel time $i$ was multiplied by the assigned population in parcel $i$ to obtain a measure of person-hours of travel ($PHT$) as follows:
 $$
@@
 Please note that this paper is a reproducible research document [see @Brunsdon2020opening] conducted using open source tools for transportation analysis [@Lovelace2021open].
-The code and data necessary to reproduce the analysis are available in a public repository\footnote{\url{https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines}}.
+The code and data necessary to reproduce the analysis are available in a [public repository](https://github.com/paezha/Accessibility-Pharmacies-Hamilton-Vaccines).
 
@@
 
-The top panel of Figure \ref{fig:maps-baseline} shows the average expected travel time by TAZ in Hamilton.
-It is apparent that travel times tend to be lower in much of suburban Hamilton, and higher in the urban core and some rural parts of the city, particularly to the west.
+The top panel of Figure 3 shows the average expected use-weighted multi-modal travel times by TAZ in Hamilton.
+It is apparent that travel times tend to be lower in much of suburban Hamilton and higher in the urban core and some rural parts of the city, particularly to the west.
 This is unsurprising, given the higher probability of travel by car and the predominantly suburban character of the vaccination sites.
-However, even accounting for the distribution of population, this leads to large disparities in the number of person-hours of travel across the city, with a concentration of the burden of travel in the urban core and the rural west (see bottom panel of Figure \ref{fig:maps-baseline}).
+However, even accounting for the distribution of population, this leads to large disparities in the number of person-hours of travel across the city, with a concentration of the burden of travel in the urban core and the rural west (see bottom panel of Figure 3).
 
@@
 
-As seen in Table \ref{tab:distribution-results}, under the pilot program approximately `r round((table_comparison[5,2] + table_comparison[4,2]) / (sum(table_comparison[1:5,2])) * 100, 2)`\% of people live in DAs in the bottom 40% of the median household income scale, but they account for `r round((table_comparison[5,3] + table_comparison[4,3]) / (sum(table_comparison[1:5,3])) * 100, 2)`\% of the total person-hours of travel.
+As seen in Table 1, under the pilot program approximately `r round((table_comparison[5,2] + table_comparison[4,2]) / (sum(table_comparison[1:5,2])) * 100, 2)`\% of people live in DAs in the bottom 40% of the median household income scale, but they account for `r round((table_comparison[5,3] + table_comparison[4,3]) / (sum(table_comparison[1:5,3])) * 100, 2)`\% of the total person-hours of travel.
 In contrast, `r round((table_comparison[1,2] + table_comparison[2,2]) / (sum(table_comparison[1:5,2])) * 100, 2)`\% of people aged 55 to 69 in DAs in the top 40% of the median household income scale accrue only `r round((table_comparison[1,3] + table_comparison[2,3]) / (sum(table_comparison[1:5,3])) * 100, 2)`\% of the total person-hours of travel.
 Where the mean travel time of residents of DAs with high median household income is `r table_comparison[1, 4] %>% set_units(h) %>% set_units(min) %>% round() %>% drop_units()` minutes, residents of lower income DAs average `r table_comparison[5, 4] %>% set_units(h) %>% set_units(min) %>% round() %>% drop_units()` minutes in travel time.
-In addition to longer average travel time, residents in lower income DAs also see substantially larger variations in travel times, and some may face considerably longer travel times (see top-left panel in Figure \ref{fig:results}).
+In addition to longer average travel time, residents in lower income DAs also see substantially larger variations in travel times, and some may face considerably longer travel times (see top-left panel in Figure 4).
 
 There are also important disparities by region.
-As shown in Table \ref{tab:distribution-results}, the urban and rural populations in Hamilton are approximately `r round((table_comparison[6,2] + table_comparison[8,2])/ (sum(table_comparison[6:8,2])) * 100, 2)`\% of the population but they bear `r round((table_comparison[6,3] + table_comparison[8,3])/ (sum(table_comparison[6:8,3])) * 100, 2)`\% of the total person-hours of travel, with also much greater variability in expected travel times (Figure \ref{fig:results}, bottom-left panel).
+As shown in Table 1, the urban and rural populations in Hamilton are approximately `r round((table_comparison[6,2] + table_comparison[8,2])/ (sum(table_comparison[6:8,2])) * 100, 2)`\% of the population but they bear `r round((table_comparison[6,3] + table_comparison[8,3])/ (sum(table_comparison[6:8,3])) * 100, 2)`\% of the total person-hours of travel, with also much greater variability in expected travel times (Figure 4, bottom-left panel).
 
 For comparison purposes we consider a scenario with some modest additions to the list of pharmacies in the provincial pilot.
-We repeat the analysis, but include the three urban sites shown in white circles in Figure \ref{fig:pharmacies-and-regions}.
-The results of this scenario appear in the last two columns of Table \ref{tab:distribution-results} and the two right panels of Figure \ref{fig:results}.
-We begin by noting that all income groups benefit from the addition of these three sites with shorter mean trip durations; the most remarkable difference is the large reduction in the disparities between residents in DAs with different levels of income.
-The top-right panel of Figure \ref{fig:results} shows that the distribution of expected travel time is now more in line for all income groups, even if the bottom two income quintiles still have somewhat wider spreads.
-With respect to the distribution of expected travel time and person-hours of travel by region, unsurprisingly the addition of three urban vaccination sites in the scenario makes a difference for urban residents but not for rural residents.
+We repeat the analysis, but include the three urban sites shown in white circles in Figure 1.
+The results of this scenario appear in the last two columns of Table 1 and the two right panels of Figure 4.
+While all income groups benefit from the addition of these three sites with shorter mean trip durations, the most remarkable difference is the large reduction in the disparities between residents in DAs with lower levels of income.
+The top-right panel of Figure 4 shows that the distribution of expected travel time is now more in line for all income groups, even if the bottom two income quintiles still have somewhat wider spreads.
+Unsurprisingly, the addition of three urban vaccination sites does not have a large impact for rural residents.
 
 The results indicate that the locations chosen by the province for the pilot vaccination program do not serve well urban or rural residents of the city, and there are some important questions regarding equity of access to the program, with a disproportionate burden in the cost of travel falling on lower income urban populations and rural populations.
-A scenario did not consider candidate sites in a systematic way.
-Nonetheless, selection of three sensible urban locations does much to alleviate disparities in the burden of transportation.
-On the other hand, unlike the urban context where there are numerous candidate locations that could be chosen for a vaccination program, there are not many candidate locations in rural regions of the city.
-Increasing access in rural Hamilton likely will involve an expansion of existing mobile vaccination pop-up clinics\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
-
-```{r figure-maps-baseline, echo=FALSE, warning=FALSE, fig.height=7, fig.align="center", fig.cap="\\label{fig:maps-baseline}Average expected travel time by TAZ (in minutes) and total person-hours of travel by TAZ."}
+Selection of three sensible urban locations does much to alleviate disparities in the burden of transportation.
+On the other hand, because there are not many candidate locations in rural parts of the city, increasing access for residents there likely necessitates an expansion of the existing mobile vaccination pop-up clinic program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
+
+```{r figure-maps-baseline, echo=FALSE, warning=FALSE, fig.height=7, fig.align="center", fig.cap="\\label{fig:maps-baseline}Figure 3.
+Average expected travel time by TAZ (in minutes) and total person-hours of travel by TAZ."}
 grid.arrange(ttm_map_baseline +
@@
 
+```{r save-figure-maps-baseline, include=FALSE, eval=FALSE}
+jpeg(filename = "images/baseline-maps.jpeg")
+grid.arrange(ttm_map_baseline +
+               theme(legend.title = element_text(size = 8),
+                     legend.text = element_text(size = 7), 
+                     axis.text = element_text(size = 7)),
+             pht_map_baseline +
+               theme(legend.title = element_text(size = 8),
+                     legend.text = element_text(size = 7), 
+                     axis.text = element_text(size = 7)),
+             nrow = 2)
+dev.off()
+
+ttm_map_baseline
+ggsave("images/ttm-baseline-map.jpeg")
+
+pht_map_baseline
+ggsave("images/pht-baseline-map.jpeg")
+```
+
 ```{r table-results, echo=FALSE}
 table_comparison  %>%
-  kable("html",
+  kable("latex",
         booktabs = TRUE,
@@
 
-```{r figure-results, echo=FALSE, warning=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:results}Distribution of expected travel time for different population groups"}
+```{r figure-results, echo=FALSE, warning=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:results}Figure 4. Distribution of expected travel time for different population groups."}
 grid.arrange(ttm_income_baseline + 
@@
 
+```{r save-figure-results, include=FALSE, eval=FALSE}
+jpeg(filename = "images/figure-results.jpeg")
+grid.arrange(ttm_income_baseline + 
+               ggtitle("By income: Pilot Program") + 
+               labs(x = "") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8), 
+                     axis.text = element_text(size = 7), 
+                     legend.position = "none"), 
+             ttm_income + 
+               ggtitle("By income: Scenario") + 
+               labs(x = "", y = "") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8), 
+                     axis.text = element_text(size = 7)), 
+             ttm_region_baseline + 
+               ggtitle("By region: Pilot Program") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8),
+                     axis.title = element_text(size = 8), 
+                     axis.text = element_text(size = 7), 
+                     legend.position = "none"), 
+             ttm_region + 
+               ggtitle("By region: Scenario") + 
+               labs(y = "") +
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8),
+                     axis.title = element_text(size = 8), 
+                     axis.text = element_text(size = 7)), 
+             nrow = 2)
+dev.off()
+
+ttm_income_baseline + 
+               ggtitle("By income: Pilot Program") + 
+               labs(x = "") + 
+               xlim(c(0,40))
+ggsave("images/ttm-income-pilot.jpeg")
+
+ttm_region_baseline + 
+  ggtitle("By region: Pilot Program") + 
+  xlim(c(0,40))
+ggsave("images/ttm-region-pilot.jpeg")
+
+ttm_income +
+  ggtitle("By income: Scenario") + 
+  labs(x = "", y = "") + 
+  xlim(c(0,40))
+ggsave("images/ttm-income.jpeg")
+
+ttm_region + 
+  ggtitle("By region: Scenario") + 
+  labs(y = "") +
+  xlim(c(0,40))
+ggsave("images/ttm-region.jpeg")
+```
+
 ## References {#references .unnumbered}

### 2021-04-06 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.Rmd
v1 Created map of population 55-69

@@
 ggsave("images/population-map.jpeg")
+
+ggplot() +
+  geom_sf(data = data_da_2016 %>%
+            filter(`Region Name` == "Hamilton"),
+          aes(fill = Population_55to69),
+          color = NA) +
+  geom_sf(data = urban_types,
+          aes(color = Type),
+          fill = NA) +
+  scale_fill_distiller(name = "Population aged 55 to 69",
+                       breaks = c(20, 50, 150, 400),
+                       palette = "OrRd", 
+                       trans = "log",
+                       direction = 1) +
+  scale_color_manual(name = "",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen")) +
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+ggsave("images/population-55-to-69-map.jpeg")
 ```

### 2021-04-07 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.Rmd
v1 Calculated person-hours of travel savings by mode

@@
 
+```{r weighted-travel-time-by-mode-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Bind the three tables and label them by mode:
+min_ttm_mode_baseline <- rbind(data.frame(min_ttm_car, Mode = "Car"),
+                               data.frame(min_ttm_transit, Mode = "Transit"),
+                               data.frame(min_ttm_walk, Mode = "Walking"))
+```
+
 ```{r aggregate-weighted-travel-times-baseline, include=FALSE}
@@
 
+```{r weighted-travel-time-by-mode, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Bind the three tables and label them by mode:
+min_ttm_mode <- rbind(data.frame(min_ttm_car, Mode = "Car"),
+                      data.frame(min_ttm_transit, Mode = "Transit"),
+                      data.frame(min_ttm_walk, Mode = "Walking"))
+```
+
 ```{r aggregate-weighted-travel-times, include=FALSE}
@@
         dplyr::select(total_weighted_person_hours, ratio))
+```
+
+```{r comparison-travel-time-by-mode, echo=FALSE}
+# Comparison of weighted travel times by mode between pilot and scenario
+min_ttm_mode_baseline %>%
+  group_by(Mode) %>%
+  summarize(pht_mode = sum(Population * weighted_travel_time/60),
+            .groups = "drop")
+
+min_ttm_mode %>%
+  group_by(Mode) %>%
+  summarize(pht_mode = sum(Population * weighted_travel_time/60),
+            .groups = "drop")
 ```

### 2021-05-03 Accessibility-Vaccination-Sites-Hamilton/Response-to-Reviewers-v2.Rmd
First revisions after reviewer reports from Findings

---
title: "Response to Reviewers"
author: "Antonio Paez & Christopher D. Higgins"
date: "5/11/2020"
output: pdf_document
---

We would like to thank two anonymous reviewers for their comments and suggestions for revisions. In this letter we respond to those comments and explain how the paper was revised in response.

\newpage

# Reviewer 1

This article examines the accessibility implications of a pilot COVID-19 vaccination program for adults age 55+ (focusing on adults age 55–69) in Hamilton, Ontario. It makes a timely contribution to accessibility literature, generally, and to a small but growing body of scholarship more specifically related to accessing the COVID-19 vaccine. This article is certainly publishable, but would benefit from minor revisions. I present some suggestions and comments following that the authors might consider as they revise the piece:

>\textcolor{blue}{Thank you for your positive assessment of the paper and suggestions for improvements. We feel that the paper is better with the revisions suggested.}

• Please clarify from the outset of the article that you all are measuring “cost of travel” in travel time.

>\textcolor{blue}{We have done this, starting from the abstract.}

• The initial description of the pilot program is a bit unclear. By my understanding, 20 pharmacies in Hamilton were added as vaccination sites, and people age 55+ qualified to be vaccinated at these 20 pharmacies. People age 55–69 do not qualify at the other 325 locations mentioned, correct? I suggest you all clarify this. Also, the authors very briefly mention mobile vaccination pop-up clinics at the end of the article. Who’s eligible for these? Where are they? How many are there? I think the analysis in your article is scoped appropriately, but I do wonder whether the accessibility “gains” garnered by this pilot program matter too much in the grand scheme, particularly if so many other stationary/mobile clinics are available or are likely to become available to the 55+ age group (or even more age groups) soon. I suppose I’m asking that the authors clarify the importance of understanding and analyzing this pilot program in light of the greater vaccine rollout in the area.

>\textcolor{blue}{We are grateful for this comment. In trying to be concise we did not manage to be clear, and we appreciate the opportunity to clarify. In this respect, it is important to note that the City of Hamilton operated dedicated vaccination sites for people 70+ and mobile pop-up clinics for people 75+\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}; however, it did not have any province-approved pharmacies as part of the pharmacy vaccination program until April 1st, when we began this research. The 325 pharmacies mentioned were all in other cities in Ontario, the closest of which is Toronto. In this way, the pilot greatly expanded access to the vaccine in the City of Hamilton, however our analysis suggests that the selection of sites was inequitable, requiring lower income populations in the city center and rural populations to travel much longer to reach a vaccination site. We revised the introduction and hope that the revisions will provide sufficient clarity with respect to the program and its relevance.}

• The authors write, “As Yu et al. (2021) note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may lead to inequities in access.” I understand word count is limited, but I encourage the authors to provide more discussion here of why this “may lead to inequities in access” and who, exactly, is likely to reap the benefits and bear the costs of these inequities.

>\textcolor{blue}{We provide more discussion about this in the introduction, in effect previewing our hypotheses for the paper.}

• The travel time threshold/walking distance cutoff values seem rather large. Did you test or consider other, smaller values?

>\textcolor{blue}{The thresholds were used to calculate the travel time tables. Since we don't know the distribution of travel times a priori, a conservative approach to ensure that every parcel can reach at least one pharmacy is to use large thresholds when running the routing algorithm. After the travel time tables have been processed the thresholds are not relevant, because the analysis is based on travel time to the nearest pharmacy. As seen in Figure 4 of the original paper, the longest trip to the nearest pharmacy from any location is approximately 40 minutes (i.e., there are lots of superfluous origin-destination pairs in our time tables).}

• The expected travel time incorporating weighted sums given the modal split in a TAZ is interesting. Did you consider analyzing/presenting findings for different modes discretely, which might highlight intramode inequities more clearly? I’m mostly curious—I do think the authors’ approach using the weighted sums capturing, in theory, travel time by all modes is a valid and interesting approach.

>\textcolor{blue}{Thank you for this excellent suggestion. We now report the estimated travel times by mode with the twenty pharmacies and with the proposed additions. The additions reduce travel time for every mode, but the largest reductions are for transit and walking. At a time when mobility is limited to essential travel, it clearly is in the public interest to reduce the need to travel longer by any mode.}

• I find it more interpretable and impactful when you all present travel time results measured in minutes rather than in person-hours. I understand needing the person-hours results as a cumulative, individual-based measure to compare across income and region classes. However, as a reader, I do find the person-hours analysis to be harder to understand than the travel time analysis. I suggest the authors put some more thought towards explaining, particularly, the person-hours analysis results clearly.

>\textcolor{blue}{We agree. The cumulative values are easier to understand in hours, but at the individual level it is less intuitive to say "0.09 hours". We have adjusted this in the revised paper.}

• I like the idea of the comparison scenario, but wonder how the scenario sites were chosen. Could the three sites feasibly serve as vaccination sites? Were these locations “optimized” in any way to lead to greater reductions in travel time disparities? Were they randomly chosen? I think more is needed to clarify this scenario experiment/analysis for the reader.

>\textcolor{blue}{The three sites we use for the scenario are existing pharmacies and of chains used by the provine elsewhere for the pilot, so they are definitely feasible for the program. We did not use any systematic approach to select these sites, we just went with three reasonable sites. The fact that three sites chosen in a reasonable but not systematic way can lead to such important reductions is telling about the effort that the province put in the selection of sites for the program. There are certainly better ways of approaching this issue (for instance, by designing optimal accessibility landscapes, see Paez et al., 2013) but we feel that designing optimal landscapes is beyond the scope of a short findings paper, since the finding that we wish to report is the inequity in the roll-out of the program.}

• The article would benefit from proofreading for spelling and grammar errors, as well as some issues with general fluency and, especially, clarity of language. For instance, I suggest “do not serve well urban and rural residents,” be rephrased as, “do not serve urban and rural residents well.” Also, a (potentially important) typo in Methods: “Using the population aged 55 to 59 y.o.”—do you all mean 55 to 69 y.o.?

>\textcolor{blue}{We proofread the paper and hopefully the overall writing is improved. In the methods we did indeed mean 55 to 69 y.o.; this has been fixed in the revision.}

• The authors should note that the Findings guidelines state, “All articles have a maximum of 1000 words of text, exclusive of References. In addition, there is a maximum figure count of 3, and maximum table count of 3.” I am not sure of the article’s current word count, but I do notice the article includes 4 figures. The authors might consider combining Figures 1 and 2 by layering the vaccination site locations atop the population distribution maps (although, this might look busy). Figure 2 might, alternatively, be included as supplemental information.

>\textcolor{blue}{We tried layering the locations of the pharmacies on the population maps, but legibility is severely compromised. We are open to including some of the figures as supplemental information at the pleasure of the editor. Perhaps the editor is willing to trade one figure for one table? (we only have two tables in the paper and both are relatively compact).}

I am glad to have had the opportunity to review this interesting paper!

>\textcolor{blue}{Thank you again for you suggestions, and the opportunity to revise and improve the paper.}

\newpage

# Reviewer 2

This paper looks at a location allocation issue that has been studied before in geograohy for several decades, where to place new facilities ti reduce travel and increase access to heakthcare in a region. The study finds that rural and low income are poorly srved and big burden of travel falls on these groups and they offer the location of new sites for vaccinations to decrease such burden. The paper depnds ona major assumotion that those who commuted by mode X will commute by the same mode for vaccination, which is not true, if COVID has shiwn us something is new travel behaviour emerge and peopel are switching modes so this assumption need to be highlighted as a major limitation. it could have been better to see a map with the old sites alone, with new sites added and have the background as the collective travel time this will enable the reader to visualize the impact spatially.

>>\textcolor{blue}{Thank you for these comments. Both authors are familiar with the literature on accessibility to healthcare, having contributed to it as recently as 2019. The findings that we wish to report here concern the specific situation of the vaccination pilot in Ontario. This is a policy-relevant issue for which there is no empirical information available besides this paper. We agree that a limitation of the analysis is that we assume all demographic/travel behavior patterns are stable since 2016. This assumption was already highlighted in the original paper (and aren't all assumptions limitations by definition?); in the revision we  redouble our effort to highlight this issue more explicitly as a limitation. We are aware that travel behavior for routine activities has changed during the pandemic (the first author has researched this in, for example, Paez, 2020, and Jamal and Paez, 2020, both papers published in Findings). But travel to get an essential vaccine is not necessarily a routine, quotidiane sort of travel behavior. We do not really know (at least for the region under study) how behavior for non-routine travel may have changed during the pandemic since we know relatively little about non-routine travel to begin with. In principle, it seems possible to do scenario analysis, but for this we would need to make up numbers regarding modal shares during the pandemic, and such an endeavor is not without problems either - besides probably requiring a substantially longer paper. We plotted the locations of the pharmacies in the travel time maps for ease of visualization.}

### 2021-05-03 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.Rmd
First revisions after reviewer reports from Findings

@@
   The Government of Ontario in Canada announced the pilot for a new vaccination program, with designated pharmacies across the province now able to offer COVID-19 vaccines.
-The accessibility of this program raises questions about the cost of travel and the distribution of the cost among the population.
-In our examination of the City of Hamilton we find that selected sites do not serve well rural and urban residents and that the associated cost of travel is expected to be disproportionally borne by lower income populations.
+The accessibility of this program raises questions about travel times to vaccination sites and the distribution of these times among the population.
+In our examination of the City of Hamilton we find that selected sites do not serve rural and urban residents well; particularly, the associated cost of travel (in terms of travel time) is expected to be disproportionally borne by lower income urban populations and rural residents.
 Modest additions to the list of pilot sites in the city can substantially alleviate this inequity.
@@
 
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(cache = TRUE)
+```
+
 ```{r install-data-package, include = FALSE}
@@
 
-```{r load-packages, include=FALSE}
+```{r load-packages, include=FALSE, cache=FALSE}
 #Load the packages needed to read and work with the data
@@
         dplyr::select(total_weighted_person_hours, ratio))
+
+colnames(table_comparison) <- c("group", "total_population", "total_weighted_person_hours_baseline", "ratio_baseline",
+                                "total_weighted_person_hours", "ratio")
+
+table_comparison <- table_comparison %>%
+  mutate(total_weighted_person_hours_baseline = round(total_weighted_person_hours_baseline, 2),
+         ratio_baseline = set_units(ratio_baseline, h) %>% 
+           set_units(min) %>% 
+           drop_units(),
+         ratio_baseline = round(ratio_baseline, 2),
+         total_weighted_person_hours = round(total_weighted_person_hours, 2),
+         ratio = set_units(ratio, h) %>% 
+           set_units(min) %>% 
+           drop_units(),
+         ratio = round(ratio, 2))
 ```
@@
 # Comparison of weighted travel times by mode between pilot and scenario
-min_ttm_mode_baseline %>%
+comparison_ttm_mode <- cbind(min_ttm_mode_baseline %>%
   group_by(Mode) %>%
-  summarize(pht_mode = sum(Population * weighted_travel_time/60),
-            .groups = "drop")
-
+  summarize(pht_mode_baseline = sum(Population * weighted_travel_time/60),
+            .groups = "drop"),
   min_ttm_mode %>%
@@
   summarize(pht_mode = sum(Population * weighted_travel_time/60),
-            .groups = "drop")
+            .groups = "drop") %>%
+    select(pht_mode)) %>% 
+  mutate(diff = paste0(round((pht_mode - pht_mode_baseline)/pht_mode_baseline * 100, 2), "%", sep = ""))
 ```
@@
 
-Along with the provision of health care facilities to treat severe cases of COVID-19 [@Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs.
-The Government of Ontario, in Canada, announced on April 1st 2021 the expansion of a pilot program to offer vaccines in pharmacies in the City of Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
-This program is in addition to dedicated vaccination centers for people aged 70+.
-Twenty pharmacies in Hamilton were added to an earlier list of 325 locations in other cities across the province, and the program was extended to people aged 55 years old and over.
+Along with the provision of health care facilities to treat severe cases of COVID-19 [e.g., @Ghorbanzadeh2021spatial; @Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs.
+The Government of Ontario, in Canada, implemented in 2021 a pilot program to offer vaccines in pharmacies for people 60 years and older.
+In the first incarnation of this program, the province approved 325 pharmacies in Toronto, Windsor-Essex, and Kingston (the latter including the region of Frontenac-Lennox and Addington).
+The City of Hamilton, the third most populous urban center in the province, operated dedicated vaccination centers for people aged 70+ and mobile pop-up clinics for people aged 75+; however, it did not have any approved pharmacies under the pilot for other populations until April 1st 2021, with the announcement of 20 pharmacies in the city \footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
+At the same time that the expansion of the pilot was announced, the province extended eligibility for this program to people 55 years and older.
 
@@
 Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast majority of the pharmacies are in suburban Hamilton.
-The issue is less clear-cut when we consider that Hamilton's older population skews suburban (see Figure \ref{fig:population-map}).
+Whether this is an issue is less clear-cut when we consider that Hamilton's older population skews suburban (see Figure \ref{fig:population-map}).
 Given the target demographic for the program, it is possible that suburban sites could be convenient for mature adults and the young old: the population aged 55 to 69 in Hamilton is approximately `r prettyNum(round((table_comparison[7,2]/1000), 2) * 1000, big.mark = ",")` suburban, `r prettyNum(round((table_comparison[8,2]/1000), 2) * 1000, big.mark = ",")` urban, and only `r prettyNum(round((table_comparison[6,2]/1000), 2) * 1000, big.mark = ",")` rural.
-Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the Mayor of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
-As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may lead to inequities in access.
+Nevertheless, the selection of sites by the province raised some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the Mayor of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
+As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may lead to inequities in access by requiring people who do not own cars or who typically travel by other modes (mostly lower income urban residents) to spend more of their time reaching an approved pharmacy.
 
@@
 
-- What is the estimated cost of travel to reach the vaccination sites, assuming that every person requires a vaccine?
-- What is the distribution of this cost across the population of the city?
-- How does the cost and its distribution change with the addition of candidate sites in urban Hamilton?
-
-We concentrate on the 55 to 69 years old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pilot.
+- What is the estimated travel time needed to reach the nearest approved pharmacy, assuming that every person requires a vaccine?
+- What is the distribution of travel times across the population of the city?
+- How does the cost of time needed for travel and its distribution change with the addition of candidate vaccination sites in urban Hamilton?
+
+We concentrate on the 55 to 69 years old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pilot (including pop-up clinics for people 75+).
 
@@
 
-Using the population aged 55 to 59 y.o.
+Using the population aged 55 to 69 y.o.
 we first calculate the average number of people per dwelling and assign them proportionally to the dwellings by parcel.
-Second, median incomes and modal splits across walking, transit, and car are joined to the parcels.
-Although our data are from 2016 (the most recent available), we assume they are largely representative of current spatial and demographic trends in the city.
+Second, median household incomes and modal splits across walking, transit, and car are joined to the parcels.
+Our data are from 2016 (the most recent available), but we assume they are largely representative of spatial, mobility, and demographic trends in the city.
+In the absence of travel behavior information during the pandemic we suggest this is an inevitable limitation of the analysis.
 Third, we use the `r5r` [@Pereira2021r5r] package to calculate the travel time from each parcel to all pharmacies by three modes using a cutoff value of 180 min and a maximum walking distance of 10,000 m.
+These thresholds are chosen to maximize the coverage of the travel time tables, but the analysis, as explained next, is based on travel to the nearest approved pharmacy.
 Once we obtained travel time tables with population, proportion of trips by mode, and income information, we calculated the expected travel time $ett$ from each parcel $i$ to a pharmacy $j$ as follows:
@@
 \noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the vector of travel times from parcel $i$ to the pharmacies.
-The expected travel time is thus the weighted sum of travel times to the nearest pharmacy, with the weights given by the expected modal split in the TAZ.
-
-The expected travel time $i$ was multiplied by the assigned population in parcel $i$ to obtain a measure of person-hours of travel ($PHT$) as follows:
+The expected travel time is thus the weighted sum of travel times to the _nearest_ pharmacy, with the weights given by the expected modal split in the TAZ.
+
+The expected travel time $i$ was multiplied by the assigned population in parcel $i$ to obtain a measure of person-hours of travel ($PHT$) that gives a cumulative cost (in travel time) for a population group.
+This is as follows:
 $$
@@
 
-The top panel of Figure \ref{fig:maps-baseline} shows the average expected use-weighted multi-modal travel times by TAZ in Hamilton.
+The top panel of Figure \ref{fig:maps-baseline} shows the average multi-modal travel times (weighted by modal shares) by TAZ in Hamilton.
 It is apparent that travel times tend to be lower in much of suburban Hamilton and higher in the urban core and some rural parts of the city, particularly to the west.
@@
 For comparison purposes we consider a scenario with some modest additions to the list of pharmacies in the provincial pilot.
-We repeat the analysis, but include the three urban sites shown in white circles in Figure \ref{fig:pharmacies-and-regions}.
+For the scenario we choose three plausible sites (i.e., pharmacies that belong to chains already used by the province for this program).
+We do not conduct a systematic search of candidate locations.
+We note that some papers propound the design optimal accessibility landscapes [e.g., @Horner2008optimal; @Paez2013exploring], but such an endeavor is beyond the scope of a short findings paper.
+Instead, we repeat the analysis after including the three urban sites shown in white circles in Figure \ref{fig:pharmacies-and-regions}.
+
 The results of this scenario appear in the last two columns of Table \ref{tab:distribution-results} and the two right panels of Figure \ref{fig:results}.
@@
 Unsurprisingly, the addition of three urban vaccination sites does not have a large impact for rural residents.
-
-The results indicate that the locations chosen by the province for the pilot vaccination program do not serve well urban or rural residents of the city, and there are some important questions regarding equity of access to the program, with a disproportionate burden in the cost of travel falling on lower income urban populations and rural populations.
-Selection of three sensible urban locations does much to alleviate disparities in the burden of transportation.
-On the other hand, because there are not many candidate locations in rural parts of the city, increasing access for residents there likely necessitates an expansion of the existing mobile vaccination pop-up clinic program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
+The scenario also suggests that important reductions in the need for travel can be achieved.
+While this is true for all three modes of transportation (see Table \ref{tab:distribution-results-by-mode}), the reductions are more substantial for travel by transit and walking.
+At a time that non-pharmaceutical interventions call for reductions in mobility, it is clearly in the public interest to reduce the need to travel even for this essential purpose.
+
+The results indicate that the locations chosen by the province for the pilot vaccination program do not serve urban or rural residents of the city well, and there are some important questions regarding equity of access to the program.
+As shown, a disproportionate burden in the cost of travel (in terms of travel time) fall on lower income urban populations and rural populations.
+Selection of three sensible urban locations does much to alleviate disparities in the burden of transportation - and the fact that this is so despite the lack of a systematic search for optimal locations does not speak well of the effort put by the province in the choice of vaccination sites.
+The situation is different for rural populations, mainly because there are not many candidate locations in rural parts of the city; in this case, it is likely that increasing access for residents there will necessitate an expansion of the existing mobile vaccination pop-up clinic program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
 
@@
 grid.arrange(ttm_map_baseline +
+               geom_sf(data = pharmacy_locations %>%
+                         dplyr::filter(Type == "Pilot"),
+                       shape = 17,
+                       size = 1,
+                       color = "black",
+                       show.legend = FALSE) +
+               geom_sf(data = pharmacy_locations %>%
+                         dplyr::filter(Type == "Candidate"),
+                       shape = 16,
+                       color = "white",
+                       size = 1,
+                       show.legend = FALSE) +
+               geom_sf(data = pharmacy_locations %>%
+                         dplyr::filter(Type == "Candidate"),
+                       shape = 1,
+                       color = "black",
+                       size = 1,
+                       show.legend = FALSE) +
                theme(legend.title = element_text(size = 8),
@@
              pht_map_baseline  +
+               geom_sf(data = pharmacy_locations %>%
+                         dplyr::filter(Type == "Pilot"),
+                       shape = 17,
+                       size = 1,
+                       color = "black",
+                       show.legend = FALSE) +
+               geom_sf(data = pharmacy_locations %>%
+                         dplyr::filter(Type == "Candidate"),
+                       shape = 16,
+                       color = "white",
+                       size = 1,
+                       show.legend = FALSE) +
+               geom_sf(data = pharmacy_locations %>%
+                         dplyr::filter(Type == "Candidate"),
+                       shape = 1,
+                       color = "black",
+                       size = 1,
+                       show.legend = FALSE) +
                theme(legend.title = element_text(size = 8),
@@
                       "Total PHT", 
-                      "Hours per person", 
+                      "Minutes per person", 
                       "Total PHT",
-                      "Hours per person"),
+                      "Minutes per person"),
         digits = 3,
         align = c("lccccc"),
-        caption = "\\label{tab:distribution-results}Distribution of person-hours of travel (PHT) by median total household income and region: pilot locations only, and scenario with three urban locations added") %>%
+        caption = "\\label{tab:distribution-results}Distribution of person-hours of travel (PHT) by median total household income and region: pilot locations only (baseline), and scenario with three urban locations added") %>%
   add_header_above(c(" " = 2, "Pilot Program" = 2, "Scenario" = 2)) %>%
@@
   footnote(general = "The population totals differ due to small differences in the classification of the regions")
+```
+
+```{r table-results-by-mode, echo=FALSE}
+comparison_ttm_mode %>%
+  kable("latex",
+        booktabs = TRUE,
+        col.names = c("Mode", 
+                      "Baseline PHT", 
+                      "Scenario PHT", 
+                      "Reduction in PHT"),
+        digits = 3,
+        align = c("lccc"),
+        caption = "\\label{tab:distribution-results-by-mode}Distribution of person-hours of travel (PHT) by mode of transportation: pilot locations only (baseline) and scenario with three urban locations added")  %>%
+  kable_styling(latex_options = c("striped"))
 ```

### 2021-05-03 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.tex
First revisions after reviewer reports from Findings

@@
 % Pandoc citation processing
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc} % for calculating minipage widths
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 
-  \title{An examination of the accessibility implications of a pilot COVID-19
-vaccination program in Hamilton, Ontario}
+  \title{An examination of the accessibility implications of a pilot
+COVID-19 vaccination program in Hamilton, Ontario}
     \author[McMaster University]{Antonio Paez\corref{1}}
@@
    \ead{cd.higgins@utoronto.ca} 
-      \address[McMaster University]{School of Earth, Environment and Society, McMaster University, Hamilton,
-ON, L8S 4K1, Canada}
-    \address[University of Toronto Scarborough]{Department of Geography \& Planning, University of Toronto Scarborough,
-1265 Military Trail, Toronto, ON M1C1A4}
+      \address[McMaster University]{School of Earth, Environment and
+Society, McMaster University, Hamilton, ON, L8S 4K1, Canada}
+    \address[University of Toronto Scarborough]{Department of Geography
+\& Planning, University of Toronto Scarborough, 1265 Military Trail,
+Toronto, ON M1C1A4}
       \cortext[1]{Corresponding Author}
@@
   The Government of Ontario in Canada announced the pilot for a new
-  vaccination program, with designated pharmacies across the province now
-  able to offer COVID-19 vaccines. The accessibility of this program
-  raises questions about the cost of travel and the distribution of the
-  cost among the population. In our examination of the City of Hamilton we
-  find that selected sites do not serve well rural and urban residents and
-  that the associated cost of travel is expected to be disproportionally
-  borne by lower income populations. Modest additions to the list of pilot
-  sites in the city can substantially alleviate this inequity.
+  vaccination program, with designated pharmacies across the province
+  now able to offer COVID-19 vaccines. The accessibility of this program
+  raises questions about travel times to vaccination sites and the
+  distribution of these times among the population. In our examination
+  of the City of Hamilton we find that selected sites do not serve rural
+  and urban residents well; particularly, the associated cost of travel
+  (in terms of travel time) is expected to be disproportionally borne by
+  lower income urban populations and rural residents. Modest additions
+  to the list of pilot sites in the city can substantially alleviate
+  this inequity.
   \end{abstract}
@@
 Along with the provision of health care facilities to treat severe cases
-of COVID-19 (Pereira, C. K. V. S. Braga, et al., 2021), another front in
-the fight against the pandemic is the rolling out of vaccination
-programs. The Government of Ontario, in Canada, announced on April 1st
-2021 the expansion of a pilot program to offer vaccines in pharmacies in
-the City of
-Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
-This program is in addition to dedicated vaccination centers for people
-aged 70+. Twenty pharmacies in Hamilton were added to an earlier list of
-325 locations in other cities across the province, and the program was
-extended to people aged 55 years old and over.
+of COVID-19 (e.g., Ghorbanzadeh et al., 2021; Pereira et al., 2021a),
+another front in the fight against the pandemic is the rolling out of
+vaccination programs. The Government of Ontario, in Canada, implemented
+in 2021 a pilot program to offer vaccines in pharmacies for people 60
+years and older. In the first incarnation of this program, the province
+approved 325 pharmacies in Toronto, Windsor-Essex, and Kingston (the
+latter including the region of Frontenac-Lennox and Addington). The City
+of Hamilton, the third most populous urban center in the province,
+operated dedicated vaccination centers for people aged 70+ and mobile
+pop-up clinics for people aged 75+; however, it did not have any
+approved pharmacies under the pilot for other populations until April
+1st 2021, with the announcement of 20 pharmacies in the city
+\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
+At the same time that the expansion of the pilot was announced, the
+province extended eligibility for this program to people 55 years and
+older.
 
@@
 Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast
-majority of the pharmacies are in suburban Hamilton. The issue is less
-clear-cut when we consider that Hamilton's older population skews
-suburban (see Figure \ref{fig:population-map}). Given the target
-demographic for the program, it is possible that suburban sites could be
-convenient for mature adults and the young old: the population aged 55
-to 69 in Hamilton is approximately 58,710 suburban, 35,490 urban, and
-only 8,360 rural. Nevertheless, the selection of sites by the province
-raises some important
+majority of the pharmacies are in suburban Hamilton. Whether this is an
+issue is less clear-cut when we consider that Hamilton's older
+population skews suburban (see Figure \ref{fig:population-map}). Given
+the target demographic for the program, it is possible that suburban
+sites could be convenient for mature adults and the young old: the
+population aged 55 to 69 in Hamilton is approximately 58,710 suburban,
+35,490 urban, and only 8,360 rural. Nevertheless, the selection of sites
+by the province raised some important
 questions\footnote{The decision-making process to select these sites appears to have been opaque, and the Mayor of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
@@
 vaccinations sites in car-oriented locations may lead to inequities in
-access.
+access by requiring people who do not own cars or who typically travel
+by other modes (mostly lower income urban residents) to spend more of
+their time reaching an approved pharmacy.
 
@@
 \item
-  What is the estimated cost of travel to reach the vaccination sites,
-  assuming that every person requires a vaccine?
+  What is the estimated travel time needed to reach the nearest approved
+  pharmacy, assuming that every person requires a vaccine?
 \item
-  What is the distribution of this cost across the population of the
+  What is the distribution of travel times across the population of the
   city?
 \item
-  How does the cost and its distribution change with the addition of
-  candidate sites in urban Hamilton?
+  How does the cost of time needed for travel and its distribution
+  change with the addition of candidate vaccination sites in urban
+  Hamilton?
 \end{itemize}
@@
 older 70+ group have access to other dedicated facilities besides those
-in the provincial pilot.
+in the provincial pilot (including pop-up clinics for people 75+).
 
@@
 
-Using the population aged 55 to 59 y.o. we first calculate the average
+Using the population aged 55 to 69 y.o. we first calculate the average
 number of people per dwelling and assign them proportionally to the
-dwellings by parcel. Second, median incomes and modal splits across
-walking, transit, and car are joined to the parcels. Although our data
-are from 2016 (the most recent available), we assume they are largely
-representative of current spatial and demographic trends in the city.
-Third, we use the \texttt{r5r} (Pereira, Saraiva, et al., 2021) package
-to calculate the travel time from each parcel to all pharmacies by three
-modes using a cutoff value of 180 min and a maximum walking distance of
-10,000 m. Once we obtained travel time tables with population,
-proportion of trips by mode, and income information, we calculated the
-expected travel time \(ett\) from each parcel \(i\) to a pharmacy \(j\)
-as follows: \[
+dwellings by parcel. Second, median household incomes and modal splits
+across walking, transit, and car are joined to the parcels. Our data are
+from 2016 (the most recent available), but we assume they are largely
+representative of spatial, mobility, and demographic trends in the city.
+In the absence of travel behavior information during the pandemic we
+suggest this is an inevitable limitation of the analysis. Third, we use
+the \texttt{r5r} (Pereira et al., 2021b) package to calculate the travel
+time from each parcel to all pharmacies by three modes using a cutoff
+value of 180 min and a maximum walking distance of 10,000 m. These
+thresholds are chosen to maximize the coverage of the travel time
+tables, but the analysis, as explained next, is based on travel to the
+nearest approved pharmacy. Once we obtained travel time tables with
+population, proportion of trips by mode, and income information, we
+calculated the expected travel time \(ett\) from each parcel \(i\) to a
+pharmacy \(j\) as follows: \[
 ett_i = p^c_i \min(tt^c_{ij}) + p^t_i \min(tt^t_{ij}) + p^w_i \min(tt^w_{ij})
@@
 from parcel \(i\) to the pharmacies. The expected travel time is thus
-the weighted sum of travel times to the nearest pharmacy, with the
-weights given by the expected modal split in the TAZ.
+the weighted sum of travel times to the \emph{nearest} pharmacy, with
+the weights given by the expected modal split in the TAZ.
 
@@
 in parcel \(i\) to obtain a measure of person-hours of travel (\(PHT\))
-as follows: \[
+that gives a cumulative cost (in travel time) for a population group.
+This is as follows: \[
 PHT_i = P_i\cdot ett_i
@@
 The top panel of Figure \ref{fig:maps-baseline} shows the average
-expected use-weighted multi-modal travel times by TAZ in Hamilton. It is
-apparent that travel times tend to be lower in much of suburban Hamilton
-and higher in the urban core and some rural parts of the city,
+multi-modal travel times (weighted by modal shares) by TAZ in Hamilton.
+It is apparent that travel times tend to be lower in much of suburban
+Hamilton and higher in the urban core and some rural parts of the city,
 particularly to the west. This is unsurprising, given the higher
@@
 35.03\% of the total person-hours of travel. Where the mean travel time
-of residents of DAs with high median household income is 6 minutes,
-residents of lower income DAs average 12 minutes in travel time. In
+of residents of DAs with high median household income is 347 minutes,
+residents of lower income DAs average 746 minutes in travel time. In
 addition to longer average travel time, residents in lower income DAs
@@
 For comparison purposes we consider a scenario with some modest
-additions to the list of pharmacies in the provincial pilot. We repeat
-the analysis, but include the three urban sites shown in white circles
-in Figure \ref{fig:pharmacies-and-regions}. The results of this scenario
-appear in the last two columns of Table \ref{tab:distribution-results}
-and the two right panels of Figure \ref{fig:results}. While all income
-groups benefit from the addition of these three sites with shorter mean
-trip durations, the most remarkable difference is the large reduction in
-the disparities between residents in DAs with lower levels of income.
-The top-right panel of Figure \ref{fig:results} shows that the
-distribution of expected travel time is now more in line for all income
-groups, even if the bottom two income quintiles still have somewhat
-wider spreads. Unsurprisingly, the addition of three urban vaccination
-sites does not have a large impact for rural residents.
+additions to the list of pharmacies in the provincial pilot. For the
+scenario we choose three plausible sites (i.e., pharmacies that belong
+to chains already used by the province for this program). We do not
+conduct a systematic search of candidate locations. We note that some
+papers propound the design optimal accessibility landscapes (e.g.,
+Horner, 2008; Páez et al., 2013), but such an endeavor is beyond the
+scope of a short findings paper. Instead, we repeat the analysis after
+including the three urban sites shown in white circles in Figure
+\ref{fig:pharmacies-and-regions}.
+
+The results of this scenario appear in the last two columns of Table
+\ref{tab:distribution-results} and the two right panels of Figure
+\ref{fig:results}. While all income groups benefit from the addition of
+these three sites with shorter mean trip durations, the most remarkable
+difference is the large reduction in the disparities between residents
+in DAs with lower levels of income. The top-right panel of Figure
+\ref{fig:results} shows that the distribution of expected travel time is
+now more in line for all income groups, even if the bottom two income
+quintiles still have somewhat wider spreads. Unsurprisingly, the
+addition of three urban vaccination sites does not have a large impact
+for rural residents. The scenario also suggests that important
+reductions in the need for travel can be achieved. While this is true
+for all three modes of transportation (see Table
+\ref{tab:distribution-results-by-mode}), the reductions are more
+substantial for travel by transit and walking. At a time that
+non-pharmaceutical interventions call for reductions in mobility, it is
+clearly in the public interest to reduce the need to travel even for
+this essential purpose.
 
 The results indicate that the locations chosen by the province for the
-pilot vaccination program do not serve well urban or rural residents of
-the city, and there are some important questions regarding equity of
-access to the program, with a disproportionate burden in the cost of
-travel falling on lower income urban populations and rural populations.
-Selection of three sensible urban locations does much to alleviate
-disparities in the burden of transportation. On the other hand, because
-there are not many candidate locations in rural parts of the city,
-increasing access for residents there likely necessitates an expansion
-of the existing mobile vaccination pop-up clinic
+pilot vaccination program do not serve urban or rural residents of the
+city well, and there are some important questions regarding equity of
+access to the program. As shown, a disproportionate burden in the cost
+of travel (in terms of travel time) fall on lower income urban
+populations and rural populations. Selection of three sensible urban
+locations does much to alleviate disparities in the burden of
+transportation - and the fact that this is so despite the lack of a
+systematic search for optimal locations does not speak well of the
+effort put by the province in the choice of vaccination sites. The
+situation is different for rural populations, mainly because there are
+not many candidate locations in rural parts of the city; in this case,
+it is likely that increasing access for residents there will necessitate
+an expansion of the existing mobile vaccination pop-up clinic
 program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
@@
 
-\caption{\label{tab:table-results}\label{tab:distribution-results}Distribution of person-hours of travel (PHT) by median total household income and region: pilot locations only, and scenario with three urban locations added}
+\caption{\label{tab:table-results}\label{tab:distribution-results}Distribution of person-hours of travel (PHT) by median total household income and region: pilot locations only (baseline), and scenario with three urban locations added}
 \centering
@@
 \cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6}
-Group & Population & Total PHT & Hours per person & Total PHT & Hours per person\\
+Group & Population & Total PHT & Minutes per person & Total PHT & Minutes per person\\
 \midrule
@@
 \multicolumn{6}{l}{\textbf{Income Quintile}}\\
-\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{23297.315} & \cellcolor{gray!6}{2243.857} & \cellcolor{gray!6}{0.096} & \cellcolor{gray!6}{2146.558} & \cellcolor{gray!6}{0.092}\\
-\hspace{1em}Second 20\% & 22356.413 & 2471.952 & 0.111 & 2351.858 & 0.105\\
-\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{19570.061} & \cellcolor{gray!6}{1749.497} & \cellcolor{gray!6}{0.089} & \cellcolor{gray!6}{1563.978} & \cellcolor{gray!6}{0.080}\\
-\hspace{1em}Fourth 20\% & 17729.139 & 2928.959 & 0.165 & 1950.312 & 0.110\\
-\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{19629.952} & \cellcolor{gray!6}{4068.548} & \cellcolor{gray!6}{0.207} & \cellcolor{gray!6}{2388.422} & \cellcolor{gray!6}{0.122}\\
+\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{23297.315} & \cellcolor{gray!6}{2243.86} & \cellcolor{gray!6}{5.78} & \cellcolor{gray!6}{2146.56} & \cellcolor{gray!6}{5.53}\\
+\hspace{1em}Second 20\% & 22356.413 & 2471.95 & 6.63 & 2351.86 & 6.31\\
+\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{19570.061} & \cellcolor{gray!6}{1749.50} & \cellcolor{gray!6}{5.36} & \cellcolor{gray!6}{1563.98} & \cellcolor{gray!6}{4.80}\\
+\hspace{1em}Fourth 20\% & 17729.139 & 2928.96 & 9.91 & 1950.31 & 6.60\\
+\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{19629.952} & \cellcolor{gray!6}{4068.55} & \cellcolor{gray!6}{12.44} & \cellcolor{gray!6}{2388.42} & \cellcolor{gray!6}{7.30}\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Region}}\\
-\hspace{1em}Rural & 8356.963 & 1730.268 & 0.207 & 1730.242 & 0.207\\
-\hspace{1em}\cellcolor{gray!6}{Suburban} & \cellcolor{gray!6}{58711.629} & \cellcolor{gray!6}{4138.482} & \cellcolor{gray!6}{0.070} & \cellcolor{gray!6}{4138.392} & \cellcolor{gray!6}{0.070}\\
-\hspace{1em}Urban & 35491.942 & 7588.590 & 0.214 & 4527.021 & 0.128\\
+\hspace{1em}Rural & 8356.963 & 1730.27 & 12.42 & 1730.24 & 12.42\\
+\hspace{1em}\cellcolor{gray!6}{Suburban} & \cellcolor{gray!6}{58711.629} & \cellcolor{gray!6}{4138.48} & \cellcolor{gray!6}{4.23} & \cellcolor{gray!6}{4138.39} & \cellcolor{gray!6}{4.23}\\
+\hspace{1em}Urban & 35491.942 & 7588.59 & 12.83 & 4527.02 & 7.65\\
 \bottomrule
@@
 \end{tabular}}
+\end{table}
+
+\begin{table}
+
+\caption{\label{tab:table-results-by-mode}\label{tab:distribution-results-by-mode}Distribution of person-hours of travel (PHT) by mode of transportation: pilot locations only (baseline) and scenario with three urban locations added}
+\centering
+\begin{tabular}[t]{lccc}
+\toprule
+Mode & Baseline PHT & Scenario PHT & Reduction in PHT\\
+\midrule
+\cellcolor{gray!6}{Car} & \cellcolor{gray!6}{8535.429} & \cellcolor{gray!6}{7092.220} & \cellcolor{gray!6}{-16.91\%}\\
+Transit & 2730.205 & 2107.684 & -22.8\%\\
+\cellcolor{gray!6}{Walking} & \cellcolor{gray!6}{2197.179} & \cellcolor{gray!6}{1201.223} & \cellcolor{gray!6}{-45.33\%}\\
+\bottomrule
+\end{tabular}
 \end{table}
@@
 \hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
 \leavevmode\hypertarget{ref-Brunsdon2020opening}{}%
@@
 
+\leavevmode\hypertarget{ref-Ghorbanzadeh2021spatial}{}%
+Ghorbanzadeh, M., Kim, K., Erman Ozguven, E., Horner, M.W., 2021.
+Spatial accessibility assessment of COVID-19 patients to healthcare
+facilities: A case study of florida. Travel Behaviour and Society 24,
+95--101. doi:\url{https://doi.org/10.1016/j.tbs.2021.03.004}
+
+\leavevmode\hypertarget{ref-Horner2008optimal}{}%
+Horner, M.W., 2008. 'Optimal' accessibility landscapes? Development of a
+new methodology for simulating and assessing jobs-housing relationships
+in urban regions. Urban Studies 45, 1583--1602.
+doi:\href{https://doi.org/10.1177/0042098008091492}{10.1177/0042098008091492}
+
 \leavevmode\hypertarget{ref-Lovelace2021open}{}%
@@
 
+\leavevmode\hypertarget{ref-Paez2013exploring}{}%
+Páez, A., Esita, J., Newbold, K.B., Heddle, N.M., Blake, J.T., 2013.
+Exploring resource allocation and alternate clinic accessibility
+landscapes for improved blood donor turnout. Applied Geography 45,
+89--97. doi:\url{http://dx.doi.org/10.1016/j.apgeog.2013.08.008}
+
 \leavevmode\hypertarget{ref-Pereira2021geographic}{}%
 Pereira, R.H.M., Braga, C.K.V.S., Mendes and Serra, L., Amaral, P.B.,
-Gouveia, N., Paez, A., 2021. Geographic access to covid-19 healthcare in
-brazil using a balanced float catchment area approach. Social Science \&
-Medicine 273, 113773.
-doi:\href{https://doi.org/https://doi.org/10.1016/j.socscimed.2021.113773}{https://doi.org/10.1016/j.socscimed.2021.113773}
+Gouveia, N., Paez, A., 2021a. Geographic access to COVID-19 healthcare
+in brazil using a balanced float catchment area approach. Social Science
+\& Medicine 273, 113773.
+doi:\url{https://doi.org/10.1016/j.socscimed.2021.113773}
 
@@
 Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga, C.K.V., Conway,
-M.W., 2021. R5r: Rapid realistic routing on multimodal transport
+M.W., 2021b. r5r: Rapid realistic routing on multimodal transport
 networks with r\textsuperscript{5} in r. Findings.
@@
 
+\end{CSLReferences}
+
 \end{document}

### 2021-05-04 Accessibility-Vaccination-Sites-Hamilton/Response-to-Reviewers-v2.Rmd
minor typographic edits

@@
 
->\textcolor{blue}{The three sites we use for the scenario are existing pharmacies and of chains used by the provine elsewhere for the pilot, so they are definitely feasible for the program.
+>\textcolor{blue}{The three sites we use for the scenario are existing pharmacies and of chains used by the province elsewhere for the pilot, so they are definitely feasible for the program.
 We did not use any systematic approach to select these sites, we just went with three reasonable sites.
@@
 We are aware that travel behavior for routine activities has changed during the pandemic (the first author has researched this in, for example, Paez, 2020, and Jamal and Paez, 2020, both papers published in Findings).
-But travel to get an essential vaccine is not necessarily a routine, quotidiane sort of travel behavior.
+But travel to get an essential vaccine is not necessarily a routine, quotidian sort of travel behavior.
 We do not really know (at least for the region under study) how behavior for non-routine travel may have changed during the pandemic since we know relatively little about non-routine travel to begin with.

### 2021-05-04 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.Rmd
minor typographic edits

@@
 Along with the provision of health care facilities to treat severe cases of COVID-19 [e.g., @Ghorbanzadeh2021spatial; @Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs.
-The Government of Ontario, in Canada, implemented in 2021 a pilot program to offer vaccines in pharmacies for people 60 years and older.
+The Government of Ontario, in Canada, implemented a pilot program to offer vaccines in pharmacies for people 60 years and older in early 2021.
 In the first incarnation of this program, the province approved 325 pharmacies in Toronto, Windsor-Essex, and Kingston (the latter including the region of Frontenac-Lennox and Addington).
@@
 
-We concentrate on the 55 to 69 years old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pilot (including pop-up clinics for people 75+).
+We concentrate on the 55 to 69 year old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pharmacy pilot (including pop-up clinics for people 75+).
 
@@
 
-The top panel of Figure \ref{fig:maps-baseline} shows the average multi-modal travel times (weighted by modal shares) by TAZ in Hamilton.
+The top panel of Figure \ref{fig:maps-baseline} shows the average multimodal travel times (weighted by modal shares) by TAZ in Hamilton.
 It is apparent that travel times tend to be lower in much of suburban Hamilton and higher in the urban core and some rural parts of the city, particularly to the west.
@@
 Where the mean travel time of residents of DAs with high median household income is `r table_comparison[1, 4] %>% set_units(h) %>% set_units(min) %>% round() %>% drop_units()` minutes, residents of lower income DAs average `r table_comparison[5, 4] %>% set_units(h) %>% set_units(min) %>% round() %>% drop_units()` minutes in travel time.
-In addition to longer average travel time, residents in lower income DAs also see substantially larger variations in travel times, and some may face considerably longer travel times (see top-left panel in Figure \ref{fig:results}).
+In addition to greater average travel times, the top-left panel in Figure \ref{fig:results} suggests residents in lower income DAs also see substantially larger variability in travel times and some may face considerably longer trip durations.
 
@@
 We do not conduct a systematic search of candidate locations.
-We note that some papers propound the design optimal accessibility landscapes [e.g., @Horner2008optimal; @Paez2013exploring], but such an endeavor is beyond the scope of a short findings paper.
+We note that some papers propound the design of optimal accessibility landscapes [e.g., @Horner2008optimal; @Paez2013exploring], but such an endeavor is beyond the scope of a short findings paper.
 Instead, we repeat the analysis after including the three urban sites shown in white circles in Figure \ref{fig:pharmacies-and-regions}.
@@
 The results indicate that the locations chosen by the province for the pilot vaccination program do not serve urban or rural residents of the city well, and there are some important questions regarding equity of access to the program.
-As shown, a disproportionate burden in the cost of travel (in terms of travel time) fall on lower income urban populations and rural populations.
-Selection of three sensible urban locations does much to alleviate disparities in the burden of transportation - and the fact that this is so despite the lack of a systematic search for optimal locations does not speak well of the effort put by the province in the choice of vaccination sites.
+As shown, a disproportionate burden in the cost of travel (in terms of travel time) falls on lower income urban populations and rural populations.
+The selection of three sensible urban locations does much to alleviate disparities in the burden of transportation - and the fact that this is so despite the lack of a systematic search for optimal locations does not speak well of the effort put forth by the province in the choice of vaccination sites.
 The situation is different for rural populations, mainly because there are not many candidate locations in rural parts of the city; in this case, it is likely that increasing access for residents there will necessitate an expansion of the existing mobile vaccination pop-up clinic program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.

### 2021-05-04 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.Rmd
V2 after revisions in response to reviewer reports

@@
 library(kableExtra)
+library(patchwork)
 library(png)
@@
 
-```{r save-population-map, include=FALSE}
+```{r save-population-map, include=FALSE, fig.height=8, dpi=300}
 ggplot() +
@@
 
-```{r save-figure-maps-baseline, include=FALSE}
-jpeg(filename = "images/baseline-maps.jpeg")
-grid.arrange(ttm_map_baseline +
+```{r save-figure-maps-baseline, include=FALSE, fig.height=8, dpi=600}
+ttm_plot <- ttm_map_baseline +
                geom_sf(data = pharmacy_locations %>%
@@
                      legend.text = element_text(size = 7), 
-                     axis.text = element_text(size = 7)),
-             pht_map_baseline  +
+                     axis.text = element_text(size = 7))
+
+pht_plot <- pht_map_baseline  +
                geom_sf(data = pharmacy_locations %>%
@@
                      legend.text = element_text(size = 7), 
-                     axis.text = element_text(size = 7)),
-             nrow = 2)
-dev.off()
-
-ttm_map_baseline
-ggsave("images/ttm-baseline-map.jpeg")
-
-pht_map_baseline
-ggsave("images/pht-baseline-map.jpeg")
+                     axis.text = element_text(size = 7))
+
+ttm_plot/pht_plot
+
+ggsave("images/baseline-maps.jpeg")
 ```
@@
 
-```{r save-figure-results, include=FALSE}
-jpeg(filename = "images/figure-results.jpeg")
-grid.arrange(ttm_income_baseline + 
+```{r save-figure-results, include=FALSE, fig.height=8, dpi=900}
+ttm_income_baseline <- ttm_income_baseline + 
                ggtitle("By income: Pilot Program") + 
@@
                      axis.text = element_text(size = 7), 
-                     legend.position = "none"), 
-             ttm_income + 
+                     legend.position = "none")
+
+ttm_income <- ttm_income + 
                ggtitle("By income: Scenario") + 
@@
                theme(title = element_text(size = 8), 
-                     axis.text = element_text(size = 7)), 
-             ttm_region_baseline + 
+                     axis.text = element_text(size = 7))
+
+ttm_region_baseline <- ttm_region_baseline + 
                ggtitle("By region: Pilot Program") + 
@@
                      axis.text = element_text(size = 7), 
-                     legend.position = "none"), 
-             ttm_region + 
+                     legend.position = "none")
+
+ttm_region <- ttm_region + 
                ggtitle("By region: Scenario") + 
@@
                      axis.title = element_text(size = 8), 
-                     axis.text = element_text(size = 7)), 
-             nrow = 2)
-dev.off()
-
-ttm_income_baseline + 
-               ggtitle("By income: Pilot Program") + 
-               labs(x = "") + 
-               xlim(c(0,40))
-ggsave("images/ttm-income-pilot.jpeg")
-
-ttm_region_baseline + 
-  ggtitle("By region: Pilot Program") + 
-  xlim(c(0,40))
-ggsave("images/ttm-region-pilot.jpeg")
-
-ttm_income +
-  ggtitle("By income: Scenario") + 
-  labs(x = "", y = "") + 
-  xlim(c(0,40))
-ggsave("images/ttm-income.jpeg")
-
-ttm_region + 
-  ggtitle("By region: Scenario") + 
-  labs(y = "") +
-  xlim(c(0,40))
-ggsave("images/ttm-region.jpeg")
+                     axis.text = element_text(size = 7))
+
+(ttm_income_baseline + ttm_income) / (ttm_region_baseline + ttm_region)
+
+ggsave("images/figure-results.jpeg")
 ```

### 2021-05-04 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.tex
V2 after revisions in response to reviewer reports

@@
 vaccination programs. The Government of Ontario, in Canada, implemented
-in 2021 a pilot program to offer vaccines in pharmacies for people 60
-years and older. In the first incarnation of this program, the province
-approved 325 pharmacies in Toronto, Windsor-Essex, and Kingston (the
-latter including the region of Frontenac-Lennox and Addington). The City
-of Hamilton, the third most populous urban center in the province,
+a pilot program to offer vaccines in pharmacies for people 60 years and
+older in early 2021. In the first incarnation of this program, the
+province approved 325 pharmacies in Toronto, Windsor-Essex, and Kingston
+(the latter including the region of Frontenac-Lennox and Addington). The
+City of Hamilton, the third most populous urban center in the province,
 operated dedicated vaccination centers for people aged 70+ and mobile
@@
 
-We concentrate on the 55 to 69 years old population segment because the
+We concentrate on the 55 to 69 year old population segment because the
 older 70+ group have access to other dedicated facilities besides those
-in the provincial pilot (including pop-up clinics for people 75+).
+in the provincial pharmacy pilot (including pop-up clinics for people
+75+).
 
@@
 The top panel of Figure \ref{fig:maps-baseline} shows the average
-multi-modal travel times (weighted by modal shares) by TAZ in Hamilton.
+multimodal travel times (weighted by modal shares) by TAZ in Hamilton.
 It is apparent that travel times tend to be lower in much of suburban
@@
 residents of lower income DAs average 746 minutes in travel time. In
-addition to longer average travel time, residents in lower income DAs
-also see substantially larger variations in travel times, and some may
-face considerably longer travel times (see top-left panel in Figure
-\ref{fig:results}).
+addition to greater average travel times, the top-left panel in Figure
+\ref{fig:results} suggests residents in lower income DAs also see
+substantially larger variability in travel times and some may face
+considerably longer trip durations.
 
@@
 conduct a systematic search of candidate locations. We note that some
-papers propound the design optimal accessibility landscapes (e.g.,
+papers propound the design of optimal accessibility landscapes (e.g.,
 Horner, 2008; Páez et al., 2013), but such an endeavor is beyond the
@@
 access to the program. As shown, a disproportionate burden in the cost
-of travel (in terms of travel time) fall on lower income urban
-populations and rural populations. Selection of three sensible urban
+of travel (in terms of travel time) falls on lower income urban
+populations and rural populations. The selection of three sensible urban
 locations does much to alleviate disparities in the burden of
@@
 systematic search for optimal locations does not speak well of the
-effort put by the province in the choice of vaccination sites. The
+effort put forth by the province in the choice of vaccination sites. The
 situation is different for rural populations, mainly because there are

### 2021-05-04 README.Rmd
V2 after revisions in response to reviewer reports

--- 2021-04-05 README.Rmd
@@
 
-The province of Ontario in Canada announced the pilot for a new vaccination program, with designated pharmacies across the province now able to offer COVID-19 vaccines.
-The accessibility of this program raises questions about the cost of travel and the distribution of the cost among the population.
-In our examination of the City of Hamilton we find that selected sites do not serve well the rural and urban population of Hamilton, and that the associated cost of travel is expected to be disproportionally borne by lower income populations.
+The Government of Ontario in Canada announced the pilot for a new vaccination program, with designated pharmacies across the province now able to offer COVID-19 vaccines.
+The accessibility of this program raises questions about travel times to vaccination sites and the distribution of these times among the population.
+In our examination of the City of Hamilton we find that selected sites do not serve rural and urban residents well; particularly, the associated cost of travel (in terms of travel time) is expected to be disproportionally borne by lower income urban populations and rural residents.
 Modest additions to the list of pilot sites in the city can substantially alleviate this inequity.
@@
                  min_ttm_walk)
+```
+
+```{r weighted-travel-time-by-mode-baseline, include=FALSE}
+# Baseline analysis (PILOT LOCATIONS ONLY)
+## Bind the three tables and label them by mode:
+min_ttm_mode_baseline <- rbind(data.frame(min_ttm_car, Mode = "Car"),
+                               data.frame(min_ttm_transit, Mode = "Transit"),
+                               data.frame(min_ttm_walk, Mode = "Walking"))
 ```
@@
         dplyr::select(total_weighted_person_hours, ratio))
-```
-
-## Research Questions and Hypotheses
-
-Along with the provision of health care facilities to treat severe cases of COVID-19 [@Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs.
-The Province of Ontario, in Canada, announced on April 1st 2021 the expansion of a pilot program to offer vaccines in pharmacies in the City of Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
-This program is in addition to dedicated vaccination centers for people aged 70+.
-Twenty pharmacies in Hamilton were added to an earlier list of 325 locations in other cities across the province, and the program was extended to people aged 55 years old and over.
+
+colnames(table_comparison) <- c("group", "total_population", "total_weighted_person_hours_baseline", "ratio_baseline",
+                                "total_weighted_person_hours", "ratio")
+
+table_comparison <- table_comparison %>%
+  mutate(total_weighted_person_hours_baseline = round(total_weighted_person_hours_baseline, 2),
+         ratio_baseline = set_units(ratio_baseline, h) %>% 
+           set_units(min) %>% 
+           drop_units(),
+         ratio_baseline = round(ratio_baseline, 2),
+         total_weighted_person_hours = round(total_weighted_person_hours, 2),
+         ratio = set_units(ratio, h) %>% 
+           set_units(min) %>% 
+           drop_units(),
+         ratio = round(ratio, 2))
+```
+
+```{r comparison-travel-time-by-mode, echo=FALSE}
+# Comparison of weighted travel times by mode between pilot and scenario
+comparison_ttm_mode <- cbind(min_ttm_mode_baseline %>%
+  group_by(Mode) %>%
+  summarize(pht_mode_baseline = sum(Population * weighted_travel_time/60),
+            .groups = "drop"),
+  min_ttm_mode %>%
+  group_by(Mode) %>%
+  summarize(pht_mode = sum(Population * weighted_travel_time/60),
+            .groups = "drop") %>%
+    select(pht_mode)) %>% 
+  mutate(diff = paste0(round((pht_mode - pht_mode_baseline)/pht_mode_baseline * 100, 2), "%", sep = ""))
+```
+
+Questions
+==========================
+
+Along with the provision of health care facilities to treat severe cases of COVID-19 [e.g., @Ghorbanzadeh2021spatial; @Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs.
+The Government of Ontario, in Canada, implemented a pilot program to offer vaccines in pharmacies for people 60 years and older in early 2021.
+In the first incarnation of this program, the province approved 325 pharmacies in Toronto, Windsor-Essex, and Kingston (the latter including the region of Frontenac-Lennox and Addington).
+The City of Hamilton, the third most populous urban center in the province, operated dedicated vaccination centers for people aged 70+ and mobile pop-up clinics for people aged 75+; however, it did not have any approved pharmacies under the pilot for other populations until April 1st 2021, with the announcement of 20 pharmacies in the city \footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
+At the same time that the expansion of the pilot was announced, the province extended eligibility for this program to people 55 years and older.
 
@@
 Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast majority of the pharmacies are in suburban Hamilton.
-The issue is somewhat less clear-cut when we consider that Hamilton's population skews suburban (see Figure \ref{fig:population-map}).
-Given the target demographic for the program, it is possible that suburban sites could be convenient for mature adults and young old: the population aged 55 to 69 in Hamilton is approximately `r prettyNum(round((table_comparison[7,2]/1000), 2) * 1000, big.mark = ",")` suburban, `r prettyNum(round((table_comparison[8,2]/1000), 2) * 1000, big.mark = ",")` urban, and only `r prettyNum(round((table_comparison[6,2]/1000), 2) * 1000, big.mark = ",")` rural.
-Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the typically inert Major of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
-As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may introduce inequities in access.
+Whether this is an issue is less clear-cut when we consider that Hamilton's older population skews suburban (see Figure \ref{fig:population-map}).
+Given the target demographic for the program, it is possible that suburban sites could be convenient for mature adults and the young old: the population aged 55 to 69 in Hamilton is approximately `r prettyNum(round((table_comparison[7,2]/1000), 2) * 1000, big.mark = ",")` suburban, `r prettyNum(round((table_comparison[8,2]/1000), 2) * 1000, big.mark = ",")` urban, and only `r prettyNum(round((table_comparison[6,2]/1000), 2) * 1000, big.mark = ",")` rural.
+Nevertheless, the selection of sites by the province raised some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the Mayor of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
+As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may lead to inequities in access by requiring people who do not own cars or who typically travel by other modes (mostly lower income urban residents) to spend more of their time reaching an approved pharmacy.
 
@@
 
-- What is the estimated cost of travel to reach the vaccination sites, assuming that every person requires a vaccine?
-- What is the distribution of this cost across the population of the city?
-- How does the cost and its distribution change with the addition of candidate sites in urban Hamilton?
-
-We concentrate on the 55 to 69 years old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pilot.
-
-```{r pharmacies-map, echo=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:pharmacies-and-regions}Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) and urban locations for scenario analysis (white circles)"}
+- What is the estimated travel time needed to reach the nearest approved pharmacy, assuming that every person requires a vaccine?
+- What is the distribution of travel times across the population of the city?
+- How does the cost of time needed for travel and its distribution change with the addition of candidate vaccination sites in urban Hamilton?
+
+We concentrate on the 55 to 69 year old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pharmacy pilot (including pop-up clinics for people 75+).
+
+```{r pharmacies-map, echo=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:pharmacies-and-regions}Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) along with urban locations for scenario analysis (white circles)."}
 #map_pharmacies <-
@@
         legend.text = element_text(size = 8))
-
-```
-
-```{r population-map, echo=FALSE, warning=FALSE, fig.height=8, fig.align="center", fig.cap="\\label{fig:population-map}Distribution of population age 55+ in the City of Hamilton"}
+```
+
+```{r save-pharmacies-map-1, include=FALSE, eval=FALSE}
+ggplot() +
+  geom_sf(data = urban_types,
+          aes(fill = Type),
+          color = "lightgray") +
+  geom_sf(data = pharmacy_locations %>%
+            filter(Type == "Pilot"),
+          aes(shape = Type,
+              color = Type),
+          size = 2.5,
+          show.legend = FALSE) +
+  scale_shape_manual(values = c("Candidate" = 16,
+                                "Pilot" = 17)) +
+  scale_color_manual(values = c("Candidate" = "white",
+                                "Pilot" = "black")) +
+  scale_fill_manual(values = c("Urban" = "firebrick2",
+                               "Suburban" = "cornflowerblue",
+                               "Rural" = "forestgreen")) +
+  guides(fill = guide_legend(title = "")) + 
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+ggsave("images/pharmacies-map-1.jpeg")
+```
+
+```{r save-pharmacies-map-2, include=FALSE, eval=FALSE}
+ggplot() +
+  geom_sf(data = urban_types,
+          aes(fill = Type),
+          color = "lightgray") +
+  geom_sf(data = pharmacy_locations,
+          aes(shape = Type,
+              color = Type),
+          size = 2.5,
+          show.legend = FALSE) +
+  scale_shape_manual(values = c("Candidate" = 16,
+                                "Pilot" = 17)) +
+  scale_color_manual(values = c("Candidate" = "white",
+                                "Pilot" = "black")) +
+  scale_fill_manual(values = c("Urban" = "firebrick2",
+                               "Suburban" = "cornflowerblue",
+                               "Rural" = "forestgreen")) +
+  guides(fill = guide_legend(title = "")) + 
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+ggsave("images/pharmacies-map-2.jpeg")
+```
+
+```{r population-map, echo=FALSE, warning=FALSE, fig.height=8, fig.align="center", fig.cap="\\label{fig:population-map}Distribution of population aged 55+ in the City of Hamilton."}
 ggplot() +
@@
 
-## Methods and Data
-
-## Data
+```{r save-population-map, include=FALSE, fig.height=8, dpi=300, eval=FALSE}
+ggplot() +
+  geom_sf(data = data_da_2016 %>%
+            filter(`Region Name` == "Hamilton") %>%
+            dplyr::select(Population_55to69, Population_70plus) %>%
+            pivot_longer(cols = -geometry, names_to = "Age", values_to = "Population") %>%
+            mutate(Age = factor(Age,
+                                levels = c("Population_55to69",
+                                           "Population_70plus"),
+                                labels = c("Age: 55 to 69",
+                                           "Age: 70+"),
+                                ordered = TRUE)) %>%
+            st_as_sf(),
+          aes(fill = Population),
+          color = NA) +
+  geom_sf(data = urban_types,
+          aes(color = Type),
+          fill = NA) +
+  scale_fill_distiller(name = "Population",
+                       breaks = c(20, 50, 150, 400),
+                       palette = "OrRd", 
+                       trans = "log",
+                       direction = 1) +
+  scale_color_manual(name = "",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen"))  +
+  facet_wrap(~ Age, nrow = 2) +
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+ggsave("images/population-map.jpeg")
+
+ggplot() +
+  geom_sf(data = data_da_2016 %>%
+            filter(`Region Name` == "Hamilton"),
+          aes(fill = Population_55to69),
+          color = NA) +
+  geom_sf(data = urban_types,
+          aes(color = Type),
+          fill = NA) +
+  scale_fill_distiller(name = "Population aged 55 to 69",
+                       breaks = c(20, 50, 150, 400),
+                       palette = "OrRd", 
+                       trans = "log",
+                       direction = 1) +
+  scale_color_manual(name = "",
+                     values = c("Urban" = "firebrick2",
+                                "Suburban" = "cornflowerblue",
+                                "Rural" = "forestgreen")) +
+  theme_minimal() +
+  theme(panel.grid = element_blank(),
+        axis.text = element_text(size = 7),
+        legend.position = "bottom",
+        legend.text = element_text(size = 8))
+
+ggsave("images/population-55-to-69-map.jpeg")
+```
+
+Methods
+============
 
@@
 
-### Open Hamilton
-
-From the open data portal of the City of Hamilton\footnote{\url{https://open.hamilton.ca/}} we obtained boundaries for the city's various regions (the definition of urban, suburban, and rural regions follows the classification of development applications).
-
-### Statistics Canada
-
-Population and income statistics at the level of Dissemination Areas (DAs) were retrieved using the package `cancensus` [@vonBergmann2021cancensus].
-Dissemination Areas are the smallest publicly available census geography in Canada.
-We use data from the 2016 Population Census.
-
-### Transportation Tomorrow Survey
-
-Data about modal split by age by place of residence were downloaded from the Data Retrieval System of the Transportation Tomorrow Survey (TTS)\footnote{\url{http://dmg.utoronto.ca/}}.
-The data are geocoded at the level of Traffic Analysis Zones (TAZ).
-
-### Other
-
-The locations of pharmacies in the pilot were obtained from public records and geocoded.
-Three urban sites not in the program were also identified and geocoded for comparison purposes.
-In addition, we converted all recorded residential parcels in the City of Hamilton to points on the road network.
-Each point includes information about the number of residential units in the parcel.
-
-## Methods
-
-We used the population aged 55 to 59 y.o.
-in each Dissemination Area to calculate the average number of people per dwelling.
-This value was then assigned proportionally to the number of dwellings per parcel.
-The median total household income of the corresponding DAs was joined to the parcels.
-In addition, we calculated the proportion of trips by mode from the total number of trips by each of the three modes retrieved from the TTS data.
-These proportions were joined to the parcel data based on their corresponding TAZ.
-The package `r5r` [@Pereira2021r5r] was used to calculate the travel time from each parcel to all pharmacies by three modes: car, transit, and walking.
-For routing purposes we used a cutoff value of 180 min and a maximum walking distance of 10,000 m.
-
+- Urban, suburban, and rural boundary definitions from the City of Hamilton\footnote{\url{https://open.hamilton.ca/}}
+- Population and median total household income for 2016 by Dissemination Area (DA) boundary using the `cancensus` package [@vonBergmann2021cancensus]
+- Modal split by age for traffic analysis zones (TAZ) from the 2016 Transportation Tomorrow Survey (TTS)\footnote{\url{http://dmg.utoronto.ca/}}
+- Locations of pilot pharmacies from public records
+- Locations of three additional major chain pharmacies for scenario testing
+- Residential units by land parcel
+
+Using the population aged 55 to 69 y.o.
+we first calculate the average number of people per dwelling and assign them proportionally to the dwellings by parcel.
+Second, median household incomes and modal splits across walking, transit, and car are joined to the parcels.
+Our data are from 2016 (the most recent available), but we assume they are largely representative of spatial, mobility, and demographic trends in the city.
+In the absence of travel behavior information during the pandemic we suggest this is an inevitable limitation of the analysis.
+Third, we use the `r5r` [@Pereira2021r5r] package to calculate the travel time from each parcel to all pharmacies by three modes using a cutoff value of 180 min and a maximum walking distance of 10,000 m.
+These thresholds are chosen to maximize the coverage of the travel time tables, but the analysis, as explained next, is based on travel to the nearest approved pharmacy.
 Once we obtained travel time tables with population, proportion of trips by mode, and income information, we calculated the expected travel time $ett$ from each parcel $i$ to a pharmacy $j$ as follows:
@@
 \noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the vector of travel times from parcel $i$ to the pharmacies.
-In other words, the expected travel time is the weighted sum of travel times to the nearest pharmacy, with the weights given by the expected modal split in the TAZ.
-
-The expected travel time $i$ was multiplied by the population in parcel $i$ to obtain a measure of person-hours of travel ($PHT$) as follows:
+The expected travel time is thus the weighted sum of travel times to the _nearest_ pharmacy, with the weights given by the expected modal split in the TAZ.
+
+The expected travel time $i$ was multiplied by the assigned population in parcel $i$ to obtain a measure of person-hours of travel ($PHT$) that gives a cumulative cost (in travel time) for a population group.
+This is as follows:
 $$
@@
 
-## Findings
-
-The top panel of Figure \ref{fig:maps-baseline} shows the average expected travel time by TAZ in Hamilton.
-It is apparent that travel times tend to be lower in much of suburban Hamilton, and higher in the urban core and some rural parts of the city, particularly to the west.
+Findings
+===================
+
+The top panel of Figure \ref{fig:maps-baseline} shows the average multimodal travel times (weighted by modal shares) by TAZ in Hamilton.
+It is apparent that travel times tend to be lower in much of suburban Hamilton and higher in the urban core and some rural parts of the city, particularly to the west.
 This is unsurprising, given the higher probability of travel by car and the predominantly suburban character of the vaccination sites.
@@
 For comparison purposes we consider a scenario with some modest additions to the list of pharmacies in the provincial pilot.
-We repeat the analysis, but include the three urban sites shown in white circles in Figure \ref{fig:pharmacies-and-regions}.
+For the scenario we choose three plausible sites (i.e., pharmacies that belong to chains already used by the province for this program).
+We do not conduct a systematic search of candidate locations.
+We note that some papers propound the design of optimal accessibility landscapes [e.g., @Horner2008optimal; @Paez2013exploring], but such an endeavor is beyond the scope of a short findings paper.
+Instead, we repeat the analysis after including the three urban sites shown in white circles in Figure \ref{fig:pharmacies-and-regions}.
+
 The results of this scenario appear in the last two columns of Table \ref{tab:distribution-results} and the two right panels of Figure \ref{fig:results}.
-We begin by noting that all income groups benefit from the addition of these three sites with shorter mean trip durations; the most remarkable difference is the large reduction in the disparities between residents in DAs with different levels of income.
+While all income groups benefit from the addition of these three sites with shorter mean trip durations, the most remarkable difference is the large reduction in the disparities between residents in DAs with lower levels of income.
 The top-right panel of Figure \ref{fig:results} shows that the distribution of expected travel time is now more in line for all income groups, even if the bottom two income quintiles still have somewhat wider spreads.
-With respect to the distribution of expected travel time and person-hours of travel by region, unsurprisingly the addition of three urban vaccination sites in the scenario makes a difference for urban residents but not for rural residents.
-
-The results indicate that the locations chosen by the province for the pilot vaccination program do not serve well urban or rural residents of the city, and there are some important questions regarding equity of access to the program, with a disproportionate burden in the cost of travel falling on lower income urban populations and rural populations.
-A scenario did not consider candidate sites in a systematic way.
-Nonetheless, selection of three sensible urban locations does much to alleviate disparities in the burden of transportation.
-On the other hand, unlike the urban context where there are numerous candidate locations that could be chosen for a vaccination program, there are not many candidate locations in rural regions of the city.
-Increasing access in rural Hamilton likely will involve an expansion of existing mobile vaccination pop-up clinics\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
+Unsurprisingly, the addition of three urban vaccination sites does not have a large impact for rural residents.
+The scenario also suggests that important reductions in the need for travel can be achieved.
+While this is true for all three modes of transportation (see Table \ref{tab:distribution-results-by-mode}), the reductions are more substantial for travel by transit and walking.
+At a time that non-pharmaceutical interventions call for reductions in mobility, it is clearly in the public interest to reduce the need to travel even for this essential purpose.
+
+The results indicate that the locations chosen by the province for the pilot vaccination program do not serve urban or rural residents of the city well, and there are some important questions regarding equity of access to the program.
+As shown, a disproportionate burden in the cost of travel (in terms of travel time) falls on lower income urban populations and rural populations.
+The selection of three sensible urban locations does much to alleviate disparities in the burden of transportation - and the fact that this is so despite the lack of a systematic search for optimal locations does not speak well of the effort put forth by the province in the choice of vaccination sites.
+The situation is different for rural populations, mainly because there are not many candidate locations in rural parts of the city; in this case, it is likely that increasing access for residents there will necessitate an expansion of the existing mobile vaccination pop-up clinic program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
 
@@
              nrow = 2)
+```
+
+```{r save-figure-maps-baseline, include=FALSE, fig.height=8, dpi=600, eval=FALSE}
+ttm_plot <- ttm_map_baseline +
+               geom_sf(data = pharmacy_locations %>%
+                         dplyr::filter(Type == "Pilot"),
+                       shape = 17,
+                       size = 1,
+                       color = "black",
+                       show.legend = FALSE) +
+               geom_sf(data = pharmacy_locations %>%
+                         dplyr::filter(Type == "Candidate"),
+                       shape = 16,
+                       color = "white",
+                       size = 1,
+                       show.legend = FALSE) +
+               geom_sf(data = pharmacy_locations %>%
+                         dplyr::filter(Type == "Candidate"),
+                       shape = 1,
+                       color = "black",
+                       size = 1,
+                       show.legend = FALSE) +
+               theme(legend.title = element_text(size = 8),
+                     legend.text = element_text(size = 7), 
+                     axis.text = element_text(size = 7))
+
+pht_plot <- pht_map_baseline  +
+               geom_sf(data = pharmacy_locations %>%
+                         dplyr::filter(Type == "Pilot"),
+                       shape = 17,
+                       size = 1,
+                       color = "black",
+                       show.legend = FALSE) +
+               geom_sf(data = pharmacy_locations %>%
+                         dplyr::filter(Type == "Candidate"),
+                       shape = 16,
+                       color = "white",
+                       size = 1,
+                       show.legend = FALSE) +
+               geom_sf(data = pharmacy_locations %>%
+                         dplyr::filter(Type == "Candidate"),
+                       shape = 1,
+                       color = "black",
+                       size = 1,
+                       show.legend = FALSE) +
+               theme(legend.title = element_text(size = 8),
+                     legend.text = element_text(size = 7), 
+                     axis.text = element_text(size = 7))
+
+ttm_plot/pht_plot
+
+ggsave("images/baseline-maps.jpeg")
 ```
@@
                       "Total PHT", 
-                      "Hours per person", 
+                      "Minutes per person", 
                       "Total PHT",
-                      "Hours per person"),
+                      "Minutes per person"),
         digits = 3,
         align = c("lccccc"),
-        caption = "\\label{tab:distribution-results}Distribution of person-hours of travel (PHT) by median total household income and region: pilot locations only, and scenario with three urban locations added") %>%
+        caption = "\\label{tab:distribution-results}Distribution of person-hours of travel (PHT) by median total household income and region: pilot locations only (baseline), and scenario with three urban locations added") %>%
   add_header_above(c(" " = 2, "Pilot Program" = 2, "Scenario" = 2)) %>%
-  kable_styling(latex_options = c("striped", "scale_down")) %>%
+  kable_styling(bootstrap_options = c("striped")) %>%
   pack_rows("Income Quintile", 1, 5) %>%
@@
 
-```{r figure-results, echo=FALSE, warning=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:results}Distribution of expected travel time for different population groups"}
+```{r table-results-by-mode, echo=FALSE}
+comparison_ttm_mode %>%
+  kable("html",
+        booktabs = TRUE,
+        col.names = c("Mode", 
+                      "Baseline PHT", 
+                      "Scenario PHT", 
+                      "Reduction in PHT"),
+        digits = 3,
+        align = c("lccc"),
+        caption = "\\label{tab:distribution-results-by-mode}Distribution of person-hours of travel (PHT) by mode of transportation: pilot locations only (baseline) and scenario with three urban locations added")  %>%
+  kable_styling(bootstrap_options = c("striped"))
+```
+
+```{r figure-results, echo=FALSE, warning=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:results}Distribution of expected travel time for different population groups."}
 grid.arrange(ttm_income_baseline + 
@@
 
-## References {#references .unnumbered}
+```{r save-figure-results, include=FALSE, fig.height=8, dpi=900, eval=FALSE}
+ttm_income_baseline <- ttm_income_baseline + 
+               ggtitle("By income: Pilot Program") + 
+               labs(x = "") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8), 
+                     axis.text = element_text(size = 7), 
+                     legend.position = "none")
+
+ttm_income <- ttm_income + 
+               ggtitle("By income: Scenario") + 
+               labs(x = "", y = "") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8), 
+                     axis.text = element_text(size = 7))
+
+ttm_region_baseline <- ttm_region_baseline + 
+               ggtitle("By region: Pilot Program") + 
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8),
+                     axis.title = element_text(size = 8), 
+                     axis.text = element_text(size = 7), 
+                     legend.position = "none")
+
+ttm_region <- ttm_region + 
+               ggtitle("By region: Scenario") + 
+               labs(y = "") +
+               xlim(c(0,40)) +
+               theme(title = element_text(size = 8),
+                     axis.title = element_text(size = 8), 
+                     axis.text = element_text(size = 7))
+
+(ttm_income_baseline + ttm_income) / (ttm_region_baseline + ttm_region)
+
+ggsave("images/figure-results.jpeg")
+```
+
+References {#references .unnumbered}
+==========
